---
title: "Final project"
author: "Group 5"
date: "11/25/2021"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
```

### Reproducibility
```{r}
set.seed(1)
```

### Uploading data
```{r}
REIMS <- read.csv("C:/Users/sinth/OneDrive/Desktop/R CSU/Machine Learning/Final project/REIMS machine learning class.csv")
```


### Prediction of USDA grade using REIMS (classification models)  
        a) Dimensionality reduction using Principal Component Analysis (PCA)
```{r}

```

        b) Dimensionality reduction using Recursive Feature Elimination (FS)
```{r}

```

        c) Logistic regression with PCA data
```{r}

```

        d) Logistic regression with FS data
```{r}

```

        e) K-nearest neighbors using PCA data
```{r}

```

        f) K-nearest neighbors using FS data
```{r}

```

        g) PCA using PCA data
```{r}

```

        h) PCA using FS data
```{r}

```

        i) Support vector machine (SVM) using PCA data
```{r}

```

        j) SVM using FS data
```{r}

```

        k) PLS-DA
```{r}

```

        l) PLS-DA
```{r}

```

### Prediction of Tenderness over time using 36h REIMS (regression models)

        Best subset selection
```{r}
#test
```




