---
title: "Final project"
author: "Group 5"
date: "11/25/2021"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
library(FactoMineR)
```

### Reproducibility
```{r}
set.seed(1)
```

### Uploading data
```{r}
REIMS <- read.csv("REIMS machine learning class.csv")

dropBase <- c("SampleID", "Cx", "GS_1", "GS_2", "UNIQ_ID", "Ten42", "Ten56", "Ten70")
responseVars <- c("USDA.grade", "Ten3", "Ten14","Ten28")

# Define the associated dataframes for each analysis. Response variable is column 1, predictors are the remaining columns
gradeDF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("USDA.grade"))))]

#Creating training and test subsets for regressions
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
train<- REIMS[dpart, ]
test<- REIMS[-dpart, ]

ten3DF_train<- na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten3DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten14DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
ten28DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])

# Completed datasets for regressions
ten3DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
```

### Data visualization using PCA
```{r}
#Performing PCA
pca1 <- PCA(gradeDF[,-1], ncp = 100, scale.unit=TRUE ,graph = F)
dat1<-gradeDF[,1]
dat1 <- data.frame(USDA.grade=dat1,pca1$ind$coord[, c(1:2)])
#Plotting PCA faced by USDA grade
p <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=USDA.grade)) +
  geom_point()+ggtitle("PCA graph of Striploin and Tenderloin")
p
```


### Prediction of USDA grade using REIMS (classification models)  
        a) Dimensionality reduction using Principal Component Analysis (PCA)
```{r}
#CODE BY PATRICK
# This section of code performs PCA on the predictors of the data set. It stores the loading vectors in the pca_loadings dataframe
# for use in the models below. It also provides a plot of Proportion of Variance Explained (PVE) vs. Principal Component.

gradeResponse <- as.factor(gradeDF$USDA.grade)
data <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
pca_obj <- prcomp(data, center=TRUE, scale=TRUE)  # perform PCA
pca_summary <- summary(pca_obj)
pca_loadings <- pca_obj$rotation #store the loading vectors for future use

pca_summary

PVE <- pca_summary$importance[2,] # store the proportion of variance explained by each PC
PrincipalComponent <- seq(1, length(PVE), 1) # Vector of Principle components for plot

#PLOT OF PVE VS. Principle Component
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=PVE))+
  geom_point()

```

        b) Dimensionality reduction using Recursive Feature Elimination (FS)
```{r}
FS_REIMS<-read.csv("REIMS_FS_data.csv")
```

        c) Logistic regression with PCA data
```{r}
# CODED BY PATRICK

#This section of code performs 10-fold cross validation on a logistic regression using different number of principal components (varying from 
# 1 principal component to 16 principal components). The optimal number of principal components to use for the logistic regression is identified
# and used to build a final logistic model on the full dataset. A graph of CV_Error vs. Number of Principal Components is provided, and a data
# frame that reports the optimal number of principal components to use, it's CV_Error, and the training error using the full dataset.

set.seed(1)

predictorsDF <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
gradeResponse <- as.factor(gradeDF$USDA.grade)

max_components <- 16 #maximum number of components to use in cross validation
n <- nrow(predictorsDF) 
k<-10 # number of folds
p <- .5 #probability threshold to bin logistic regression response

cv_labels <- sample(seq_len(k), n, replace = TRUE) #labels for folds
cv_mse <- rep(NA, max_components) # place holder vector to record cv_mse for each PC

#loop over each number of principle components
for (i in seq_len(max_components)) {
    num_components <- i
    
    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    #compute the scores (z-values) for the given number of principal components
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
    scoresDF$USDA.grade <- gradeResponse
    
    cv_mse_i <- rep(NA, k) 
    
    #perform 10-fold cross validation
    for(j in seq_len(k)) {
        train <- scoresDF[cv_labels != j,]
        test <- scoresDF[cv_labels == j,]
        
        fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = train)
        pred <- predict(fit, newdata = test, type="response")
        confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=test$USDA.grade)
        
        #it was found that sometimes the predicted response only categorized one response, not two
        # so this error handling was necessary to get the appropriate error rate.
        if ( nrow(confMatrix) == 2) {
          errorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  
        } else {
          errorRate <- 1 - (confMatrix[1,1])/ sum(confMatrix)
        }
        
        cv_mse_i[j] <- errorRate
      }
    
    cv_mse[[i]] <- mean(cv_mse_i) #store the average CV-error of the 10 folds
}

#PLOT OF K-Fold CV_MSE VS. Principle Component
PrincipalComponent <- seq_len(max_components)
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=cv_mse))+
  geom_point()

#store the optimal cv error rate associated with the optimal number of principal components
optimal_CVerror <- min(cv_mse)
optimal_PC <- which.min(cv_mse)


#train a model with the optimal number of principal components and report
loadingDF <- pca_loadings[,1:optimal_PC]
scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
scoresDF$USDA.grade <- gradeResponse
fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = scoresDF)
pred <- predict(fit,type="response")
confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=scoresDF$USDA.grade)
TrainErrorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  

data.frame(optimalNumPrincpleCompoents = c(optimal_PC), optimalCVerror = c(optimal_CVerror), trainError = c(TrainErrorRate))

```

        d) Logistic regression with FS data
```{r}

```

        e) K-nearest neighbors using PCA data
```{r}

```

        f) K-nearest neighbors using FS data
```{r}

```

        g) PCA using PCA data
```{r}

```

        h) PCA using FS data
```{r}

```

        i) Support vector machine (SVM) using PCA data
        
            
            iA) Linear SVM
```{r}

# This calls a script to run the Linear SVM
source("SVM/SVM_linear.R", local = knitr::knit_global())


```


```{r}
#Linear SVM Plotting and Results

svmLinResultsDF <- read.csv("SVM/Results/LinSVM.csv")
BestLinModel <- readRDS("SVM/BestSVMModels/BestLinModel.rds")
BestLinModel

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents_linSVM, y=CV_MSE_linSVM)) +
  geom_line()+
  geom_point()

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents_linSVM, y=Best_Cost)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost_lin, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost_lin, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC = c(svmLinResultsDF$PrincipalComponents_linSVM[which.min(svmLinResultsDF$CV_MSE_linSVM)])
optimalCVError =  c(min(svmLinResultsDF$CV_MSE_linSVM))
optimalCost = c(svmLinResultsDF$Best_Cost[which.min(svmLinResultsDF$CV_MSE_linSVM)])

data.frame(optimalPC = optimalPC, optimalCVError =optimalCVError,optimalCost= optimalCost)

```



            iB) Polynomial SVM with PC

```{r}


# This calls a script to run the Poly SVM
source("SVM/SVM_Poly.R", local = knitr::knit_global())



```

```{r}
#Plotting and Results for Poly SVM

abs_min_cost_poly <- .0001
abs_max_cost_poly <- 10000
BestPolyModel <- readRDS("SVM/BestSVMModels/BestPolyModel.rds")
BestPolyModel

svmPolyResultsDF <- read.csv("SVM/Results/PolySVM.csv")

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE_polySVM)) +
  geom_line()+
  geom_point()

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Cost_polySVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost_poly, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost_poly, linetype="dashed", color = "red") +
  scale_y_log10()

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Degree)) +
  geom_line()+
  geom_point()

optimalPC <- c(svmPolyResultsDF$PrincipalComponents[which.min(svmPolyResultsDF$CV_MSE_polySVM)])
optimalCVError <- c(min(svmPolyResultsDF$CV_MSE_polySVM))
optimalCost <- c(svmPolyResultsDF$Best_Cost_polySVM[which.min(svmPolyResultsDF$CV_MSE_polySVM)])
optimalDegree <- c(svmPolyResultsDF$Best_Degree[which.min(svmPolyResultsDF$CV_MSE_polySVM)])

#Results from tuning across all PCs
data.frame(optimalPC = optimalPC , optimalCVError = optimalCVError, optimalCost = optimalCost)

```


            iB) Radial SVM with PC

```{r}


# This calls a script to run the Radial SVM
source("SVM/SVM_rad.R", local = knitr::knit_global())


```


```{r}
abs_min_cost <- .0001
abs_max_cost <- 10000

abs_min_gamma <- .0001
abs_max_gamma <-10000

svmRadResultsDF <- read.csv("SVM/Results/radSVM.csv")
BestRadModel <- readRDS("SVM/BestSVMModels/BestRadModel.rds")
BestRadModel

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=CV_MSE_radSVM)) +
  geom_line()+
  geom_point()

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Cost_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10()

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Gamma_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_gamma, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_gamma, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- c(svmRadResultsDF$PrincipalComponents_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])
optimalCVError <- c(min(svmRadResultsDF$CV_MSE_radSVM))
optimalCost <- c(svmRadResultsDF$Best_Cost_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])
optimalGamma <- c(svmRadResultsDF$Best_Gamma_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])

#Results from tuning across all PCs
data.frame(optimalPC = optimalPC , optimalCVError = optimalCVError, optimalCost = optimalCost, optimalGamma = optimalGamma)

```


        j) SVM using FS data
```{r}


```


        k) PLS-DA
```{r}

```

        l) PLS-DA
```{r}

```

### Prediction of Tenderness over time using 36h REIMS (regression models)

     Subsets
```{r, echo = FALSE}
#Here are the names of the data sets for all the following models. They correspond to a 60/40 training/test subsets.
ten3DF_train
ten3DF_test
ten14DF_train
ten14DF_test
ten28DF_train
ten28DF_test
```

        a) Best subset selection
```{r}
#


```


        b) PCR
```{r}
#Fitting the PCR Ten3 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten3<- pcr (Ten3~., data=ten3DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten3)
mse <- MSEP(pcr.fit_ten3)
#M of the smallest CV error
mse$comps[which.min(as.data.frame(mse$val)[1,])]
#Calculating MSE with M=6 and the Test set.
pcr.pred <- predict (pcr.fit_ten3, ten3DF_test, ncomp = 6)
pcr.pred-ten3DF_test$Ten3
mse_mean_ten3<-mean ((pcr.pred - ten3DF_test$Ten3)^2)
#Error compare to average
sqrt(mse_mean_ten3)
mean(ten3DF$Ten3)


#Fitting the PCR Ten14 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten14<- pcr (Ten14~., data=ten14DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten14)
mse <- MSEP(pcr.fit_ten14)
#M of the smallest CV error
mse$comps[which.min(as.data.frame(mse$val)[1,])]
#Calculating MSE with M=1 and the Test set.
pcr.pred <- predict (pcr.fit_ten14, ten14DF_test, ncomp = 1)
pcr.pred-ten14DF_test$Ten14
mse_mean_ten14<-mean ((pcr.pred - ten14DF_test$Ten14)^2)
#Error in the same scale
sqrt(mse_mean_ten14)
mean(ten14DF$Ten14)


#Fitting the PCR Ten28 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten28<- pcr (Ten28~., data=ten28DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten28)
mse <- MSEP(pcr.fit_ten28)
#M of the smallest CV error
mse$comps[which.min(as.data.frame(mse$val)[1,])]
#Calculating MSE with M=2 and the Test set.
pcr.pred <- predict (pcr.fit_ten28, ten28DF_test, ncomp = 2)
pcr.pred-ten28DF_test$Ten28
mse_mean_ten28<-mean ((pcr.pred - ten28DF_test$Ten28)^2)
#Error in the same scale
sqrt(mse_mean_ten28)
mean(ten28DF$Ten28)

#Summarizing results

```


        c)Lasso Regression
```{r}
#CODED BY PATRICK
set.seed(1)
#Ten3 Analysis


X_train <- model.matrix(Ten3 ~ ., data = ten3DF_train)[,-1]
Y_train <- ten3DF_train$Ten3
X_test <- model.matrix(Ten3 ~ ., data = ten3DF_test)[,-1]
Y_test <- ten3DF_test$Ten3

lambda <- 10^seq(-2, 3, length.out = 1000)

cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()

lambdaMin <- cv_lasso$lambda.min
lambdaMin

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin, type="response")

testErrorLasso <- mean(( predictions[,'s1'] - Y_test )^2)
testErrorLasso

ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) +
  geom_point(aes(Train_predictions, Y_train))


```
```{r}
# Lasso on principal components for Ten3

source("Lasso/Lasso_ten3.R", local = knitr::knit_global())


```

```{r}

LassoTen3Results <- read.csv("Lasso/Results/Ten3.csv")
BestTen3LassoModel <- readRDS("Lasso/BestModels/BestLassoTen3Model.rds")
BestTen3LassoModel

ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- LassoTen3Results$PrincipalComponents[which.min(LassoTen3Results$CV_MSE)]
optimal_TestError <-  min(LassoTen3Results$CV_MSE)
optimalLambda <- LassoTen3Results$Best_Lambda[which.min(LassoTen3Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

```
```{r}
# Lasso on principal components for Ten14

source("Lasso/Lasso_ten14.R", local = knitr::knit_global())

```

```{r}

LassoTen14Results <- read.csv("Lasso/Results/Ten14.csv")
BestTen14LassoModel <- readRDS("Lasso/BestModels/BestLassoTen14Model.rds")
BestTen14LassoModel

ggplot(data=LassoTen14Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen14Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- LassoTen14Results$PrincipalComponents[which.min(LassoTen14Results$CV_MSE)]
optimal_TestError <-  min(LassoTen14Results$CV_MSE)
optimalLambda <- LassoTen14Results$Best_Lambda[which.min(LassoTen14Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)


```
```{r}
# Lasso on principal components for Ten28

source("Lasso/Lasso_ten28.R", local = knitr::knit_global())


```

```{r}

LassoTen28Results <- read.csv("Lasso/Results/Ten28.csv")
BestTen28LassoModel <- readRDS("Lasso/BestModels/BestLassoTen28Model.rds")
BestTen28LassoModel

ggplot(data=LassoTen28Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen28Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- LassoTen28Results$PrincipalComponents[which.min(LassoTen28Results$CV_MSE)]
optimal_TestError <-  min(LassoTen28Results$CV_MSE)
optimalLambda <- LassoTen28Results$Best_Lambda[which.min(LassoTen28Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

```
