---
title: "Evaluation of Rapid Evaporative Ionization Mass Spectrometry to predict Beef Tenderness and USDA Grade"
author: "Patrick Noe, Logan Perkins, Michael Hernandez"
date: "12/16/2021"
output: 
  bookdown::pdf_document2: default
abstract: "Beef grade and tenderness are important qualities to both understand and predict within the meat processing industry, and most contemporary technologies are limited in their capabilities in making timely and cost effective analysees of these qualities of beef. More modern technologies like Rapid Evaporative Ionization Mass Spectrometry (REIMS) are available to assess the chemical composition of beef in an efficient manner. The purpose of this study was to analyze REIMs data on 83 beef striploins to explore the predictive preformance of REIMS on both USDA quality grade and the beef tenderness at three aging periods (3, 14, and 28 days). The study develops and compares several different statistical learning models trained on REIMs data to predict grade and tenderness. The purpose of these models are to (1) Discover if there is a predictive relationship between REIMs data and beef grade/tenderness, and (2) Select a model that performs the best at prediction."
---

```{r setup, include=FALSE}
library(bookdown)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(fig.pos = 'H')
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
library(FactoMineR)
library(kableExtra)
```

# INTRODUCTION
Beef palatability depends on tenderness, juiciness, and flavor. In the United States, there is a quality grading system that classifies beef carcasses on different quality categories (USDA). Some of the most common USDA quality grades are Prime, Choice, and Select, with Prime the best quality and Select being the lowest quality of the three classifications. The USDA quality grade is based on animal maturity and marbling level (intramuscular fat), and it is correlated with the overall consumer acceptance (USDA). However, tenderness is not predicted with USDA quality grade. Beef industry stakeholders have been searching for a non-destructive, real-time instrument for determining the tenderness of beef for nearly 40 years. Although some technologies allow the determination of beef tenderness, they require sacrificing a loin steak and cannot be performed in real-time. Rapid Evaporative Ionization Mass Spectrometry (REIMS) is a novel environmental mass spectrometry technique that combines a sample vaporization device with a time-of-flight mass spectrometer. When a sampling device, which uses a high-frequency current, comes into contact with the sample surface, an aerosol of molecules is generated and transformed into a mass spectral fingerprint. This technology has the advantage that it can be used in-situ and in the real-time classification of biological tissues without any sample preparation. Therefore, the goal of this study was to evaluate the ability of REIMS to predict USDA quality grade and beef tenderness.

```{r}
### Reproducibility
set.seed(1)
```


```{r}
### Uploading data
REIMS <- read.csv("REIMS machine learning class.csv")

dropBase <- c("SampleID", "Cx", "GS_1", "GS_2", "UNIQ_ID", "Ten42", "Ten56", "Ten70")
responseVars <- c("USDA.grade", "Ten3", "Ten14","Ten28")

# Define the associated dataframes for each analysis. Response variable is column 1, predictors are the remaining columns
gradeDF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("USDA.grade"))))]
```

# METHODOLOGY

## Sample collection and aging
To assess the ability of REIMS to predict beef tenderness and USDA quality grade, samples consisting on 2 USDA quality grades (USDA Select, and USDA upper 2/3 Choice), and 3 aging periods (3, 14, and 28 postmortem days) were evaluated. A total of eithy-three beef carcasses were selected at the time of grading (36h postmortem). Forty-one carcasses graded as USDA Select, and forty-two carcasses graded as upper 2/3 Choice were selected to ensure a balanced data. Additionally to the carcasses data, a sliver of the *longissimus lumborum* was collected. During fabrication, whole strip loins were collected from the carcasses selected and each strip loin were divided into three pieces and randomly assigned to one of three aging treatments (3, 14, 28 days). After 



``` {r}
#Creating training and test subsets for regressions
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
train<- REIMS[dpart, ]
test<- REIMS[-dpart, ]

ten3DF_train<- na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten3DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten14DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
ten28DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])

# Completed datasets for regressions
ten3DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
```

# RESULTS

## Data visualization using Principal Component Analysis (PCA)  

There is not a evident cluster by USDA grade and tenderness at day 3, 14, and 28 days

```{r pcaGradePlot1, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Plot of Principal Component 2 vs. Principal Component 1 colored by USDA Grade. As is seen in the plot, there is no apparent clustering of the first two principal components."}
#Performing PCA
pca1 <- PCA(REIMS[,-c(1:12)], ncp = 100, scale.unit=TRUE ,graph = F)
dat1<-REIMS[,6]
#Building data frame
dat1 <- data.frame(USDA.grade=dat1,Ten3=REIMS[,7],Ten14=REIMS[,8],Ten28=REIMS[,9],pca1$ind$coord[, c(1:50)])
#Plotting PCA faced by USDA grade
p <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=USDA.grade)) +
  geom_point()+
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
print(p + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Principal Component 2", x = "Principal Component 1"))
```

```{r pcaGradePlot2, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Plot of Principal Component 3 vs. Principal Component 2 colored by USDA Grade Data. As is seen in the plot, there is no apparent clustering between the second and third principal components."}
#Plotting PCA faced by USDA grade
p1 <- ggplot(data = dat1, aes(x = Dim.2, y = Dim.3,colour=USDA.grade)) +
  geom_point()+
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
print(p1 + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Principal Component 3", x = "Principal Component 2"))
```

```{r pcaTen3Plot, include=TRUE, fig.width = 6, fig.height = 3, fig.cap = "Plot of Principal Component 2 vs. Principal Component 1 colored by beef tenderness at 3 days. As is seen in the plot, there is no apparent clustering between the second and first principal components."}
#Plotting PCA colored by tenderness at day 3
p2 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten3)) +
  geom_point()+
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
print(p2 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))
```

```{r pcaTen14Plot, include=TRUE, fig.width = 6, fig.height = 3, fig.cap ="Plot of Principal Component 2 vs. Principal Component 1 colored by beef tenderness at 14 days. As is seen in the plot, there is no apparent clustering between the second and first principal components."}
#Plotting PCA colored by tenderness at day 14
p3 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten14)) +
  geom_point()+
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
print(p3 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))
```

```{r pcaTen28Plot, include=TRUE, fig.width = 6, fig.height = 3, fig.cap ="Plot of Principal Component 2 vs. Principal Component 1 colored by beef tenderness at 28 days. As is seen in the plot, there is no apparent clustering between the second and first principal components."}
#Plotting PCA colored by tenderness at day 28
p4 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten28)) +
  geom_point()+
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
print(p4 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))
```

### Proportion of Variance Explained
While analyzing the first several principal components did not show significant clustering of either USDA Grade or beef tenderness, principal components analysis did provide a useful tool to reducing the dimensionality of several of the models that were tested in this study. The Proportion of Variance Explained (PVE) by each principal component is shown in  \@ref(fig:pvePlot). The figure shows that the majority of the variability in the predictors lies within the first twenty principal components. More specifically, almost 31% of variability in the data is captured by the first principal component. There is a noticeable drop in variability explained by principal component 2 (PVE = .19) and principal component 3 (PVE = .06). As shown in the figure, there is another significant drop in variability explained between principal component 9 and 10. The proportion of variance explained drops at a mostly consistent rate between principal components 10 and 24. After this, much less information is contained in each of the remaining principal components. \@ref(tab:PVE-table) shows the cumulative variance that is explained using increasing values of principal components. As shown by the table, nearly 100% of the variability in the data is captured by the first 80 principal components. 

This information and the transformed predictor space provided by PCA allowed for the dimensionality reduction of the predictor space. This reduced dimensionality allowed models that do not have inherent dimension reduction techniques, like logistic regression and support vector machines, to be utilized by this study. The way that Principal Component Analysis was utilized within models will be discussed in further detail within the sections of the report that describe these models. It is worth noting that, often times, the optimal was identified where an "elbow" on the PVE plot did occur.  

```{r pvePlot, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Proportion of Variance Explained in the data by each principal component."}
#CODE BY PATRICK
# This section of code performs PCA on the predictors of the data set. It stores the loading vectors in the pca_loadings dataframe
# for use in the models below. It also provides a plot of Proportion of Variance Explained (PVE) vs. Principal Component.

gradeResponse <- as.factor(gradeDF$USDA.grade)
data <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
pca_obj <- prcomp(data, center=TRUE, scale=TRUE)  # perform PCA
pca_summary <- summary(pca_obj)
pca_loadings <- pca_obj$rotation #store the loading vectors for future use

#pca_summary

PVE <- pca_summary$importance[2,] # store the proportion of variance explained by each PC
PrincipalComponent <- seq(1, length(PVE), 1) # Vector of Principle components for plot

#PLOT OF PVE VS. Principle Component
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=PVE))+
  geom_point()+
  labs(title= "Proportion of Variance Explained (PVE) vs. Principal Component", y="PVE", x = "Prinicipal Component") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

```{r PVE-table, include=TRUE}
cummVarExp <- unname(pca_summary$importance[3,])

cummVar <- c(cummVarExp[1], cummVarExp[5], cummVarExp[10],cummVarExp[15],cummVarExp[20],cummVarExp[30], cummVarExp[40], cummVarExp[50],cummVarExp[60], cummVarExp[80])*100

PCs <- c(1,5, 10,15,20,30, 40, 50, 60, 80)

SummaryDF <- data.frame(PrincipalComponents = PCs, CummulativeVariabilityExplained = cummVar)

SummaryDF%>%
kable(caption="Cummulative percentage of predictor variability explained for increasing numbers of principal components.",digits=1)
```

## Categorical Prediction of USDA Grade 

The capability of using REIMs data to predict the USDA Grade of beef samples is a categorical problem. For the purposes of this study, the response, USDA Grade, could be of two classes: Choice or Select. Several models were built to assess predictive performance including a Logistic Regression Model, a Linear Support Vector Machine, a Polynomial Support Vector Machine, and a Radial Support Vector Machine. This section of the report discusses each of this models, their predictive performance, and a selection of the best performing model. It is important to note that model performance for each of these models was assessed in terms of the 10-fold Cross Validation Error Rate. Because the dataset is small (83 samples), it was determined that cross validation error would be a better estimation of model performance than simple test error because it allowed for less data to be withheld from the training of the model and a more comprehensive assessment of how the model would perform on new samples.

### Logistic Regression


```{r LogRegressionCode}

# CODED BY PATRICK

#This section of code performs 10-fold cross validation on a logistic regression using different number of principal components (varying from 
# 1 principal component to 16 principal components). The optimal number of principal components to use for the logistic regression is identified
# and used to build a final logistic model on the full dataset. A graph of CV_Error vs. Number of Principal Components is provided, and a data
# frame that reports the optimal number of principal components to use, it's CV_Error, and the training error using the full dataset.

set.seed(1)

predictorsDF <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
gradeResponse <- as.factor(gradeDF$USDA.grade)

max_components <- 80 #maximum number of components to use in cross validation
n <- nrow(predictorsDF) 
k<-10 # number of folds
p <- .5 #probability threshold to bin logistic regression response

cv_labels <- sample(seq_len(k), n, replace = TRUE) #labels for folds
cv_mse <- rep(NA, max_components) # place holder vector to record cv_mse for each PC

#loop over each number of principle components
for (i in seq_len(max_components)) {
    num_components <- i
    
    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    #compute the scores (z-values) for the given number of principal components
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
    scoresDF$USDA.grade <- gradeResponse
    
    cv_mse_i <- rep(NA, k) 
    
    #perform 10-fold cross validation
    for(j in seq_len(k)) {
        train <- scoresDF[cv_labels != j,]
        test <- scoresDF[cv_labels == j,]
        
        fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = train)
        pred <- predict(fit, newdata = test, type="response")
        confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=test$USDA.grade)
        
        #it was found that sometimes the predicted response only categorized one response, not two
        # so this error handling was necessary to get the appropriate error rate.
        if ( nrow(confMatrix) == 2) {
          errorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  
        } else {
          errorRate <- 1 - (confMatrix[1,1])/ sum(confMatrix)
        }
        
        cv_mse_i[j] <- errorRate
      }
    
    cv_mse[[i]] <- mean(cv_mse_i) #store the average CV-error of the 10 folds
}

#PLOT OF K-Fold CV_MSE VS. Principle Component
PrincipalComponent <- seq_len(max_components)
#store the optimal cv error rate associated with the optimal number of principal components
optimal_CVerror <- min(cv_mse)
optimal_PC <- which.min(cv_mse)
#train a model with the optimal number of principal components and report
loadingDF <- pca_loadings[,1:optimal_PC]
scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
scoresDF$USDA.grade <- gradeResponse
fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = scoresDF)
pred <- predict(fit,type="response")
confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=scoresDF$USDA.grade)
TrainErrorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  

bestLogModelDF <- data.frame(optimalNumPrincpleCompoents = c(optimal_PC), optimalCVerror = c(optimal_CVerror))
PCvsCV_DF <- data.frame(PrincipalComponents = PrincipalComponent, CV_Error = cv_mse)

write.csv(bestLogModelDF,"LogisticRegression/bestLogModelDF.csv", row.names = FALSE)
write.csv(PCvsCV_DF,"LogisticRegression/PCvsCV.csv", row.names = FALSE)

```

The Logistic Regression technique was implemented with the principal component scores provided by the PCA analysis.The general technique used to arrive at the "best" logistic regression model was to build a logistic regression model using increasing numbers of principal component directions of the predictor dataset. This process was done using between 1 to 80 principal components. 80 was selected as the maximum number of principal components to utilize to build the model because, as discussed in the PCA Analysis above, 99.8% of the predictor variability was contained in the first 80 principal components. For each model containing differing numbers of principal components, 10-fold Cross Validation was performed to identify a cross validation error to assess the model performance. 

```{r log-plot, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Logistic Regression Model CV Error vs. Number of Principal Components included."}
# Log Model Visualization
optimalIndex <- which.min(bestLogModelDF$optimalCVerror)
optimalPC_Log <- bestLogModelDF$optimalNumPrincpleCompoents[optimalIndex]
optimalCVError_Log <- round(min(bestLogModelDF$optimalCVerror),3)

ggplot(data = data.frame(PrincipalComponent = PrincipalComponent, cv_mse = cv_mse), aes(y=cv_mse, x=PrincipalComponent)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

The results of how cross validation error (CV Error) varied for models with differing numbers of principal components are shown in \@ref(fig:log-plot). As is shown in the figure, CV Error for USDA Grade prediction decreased rapidly between models that utilized from 1 to 5 principal components. The best Logistic Regression model was identified as that that minimized CV Error, and this occurred in the model that included the first 9 principal components (see \@ref(tab:log-summary)), and had a CV Error of 20%. This is interesting because, as was mentioned in the discussion of PCA to explore the predictor space, there is a significant decreased in the proportion of variance explained between the 9th and 10th principal component (see \@ref(fig:pvePlot)), and the optimal logistic regression model was determined to be one that utilized roughly 76% of the variability in the data. Furthermore, as the number of principal components that were utilized to build the logistic regression model were increased above 9, there was a noticeable (but not absolute) trend upwards of CV Error.


```{r log-summary, include=TRUE}
# Log Model Performance
summaryDF <- data.frame(PrincipalComponents = bestLogModelDF$optimalNumPrincpleCompoents, CV_Error = bestLogModelDF$optimalCVerror)
summaryDF%>%
kable(caption="Best performing logistic regression model.",digits=2)

```




### K-Nearest Neighbors
```{r}

```


### PCA
```{r}

```

### Support Vector Machines (SVM)

Three different Support Vector Machine (SVM) models were built to predict USDA Grade using data provided by REIMs: a SVM with a linear, polynomial, and radial kernel. The best SVM associated with each kernel were determined similarly to the best Logistic Regression model. For each kernel type, SVM's were built and tuned using an increasing number of principal components, and the best model was identified as the one that minimized 10-fold cross validation error.
            
#### *Linear SVM*



```{r linSVMCode}
# This calls a script to run the Linear SVM. It takes a long time to run so results were saved in a CSV and imported for results.
#source("SVM/SVM_linear.R", local = knitr::knit_global())
```

```{r svmLin-CVPlot, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Linear Support Vector Machine Optimal Cross Validation error for differing numbers of principal components included." }
#Linear SVM Plotting and Results
svmLinResultsDF <- read.csv("SVM/Results/LinSVM.csv")
BestLinModel <- readRDS("SVM/BestSVMModels/BestLinModel.rds")
BestLinModel <- BestLinModel[[1]]
numSupportVectors_linSVM <- nrow(BestLinModel$SV)
optimalPC_linSVM = svmLinResultsDF$PrincipalComponents_linSVM[which.min(svmLinResultsDF$CV_MSE_linSVM)]
optimalCVError_linSVM =  min(svmLinResultsDF$CV_MSE_linSVM)
optimalCost_linSVM = svmLinResultsDF$Best_Cost[which.min(svmLinResultsDF$CV_MSE_linSVM)]

SummaryDF <- data.frame(PC = optimalPC_linSVM , CV_Error = optimalCVError_linSVM, Cost = optimalCost_linSVM, SV = numSupportVectors_linSVM)


ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents_linSVM, y=CV_MSE_linSVM)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        plot.title = element_text(hjust = 0.5))
```


```{r svmLin-CostPlot, include=TRUE,fig.width = 6, fig.height = 2.5, fig.cap="Optimal Cost identfied for each linear support vector machine of differing numbers of principal components. The red dashed lines represent the mininum and maximum of the search range used to tune the support vector machine."}
abs_min_cost <- .01
abs_max_cost <- 100

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents_linSVM, y=Best_Cost)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10()+
  labs(title= "Optimal Cost vs. Principal Components", y="Optimal Cost", x = "Number of Principal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```


```{r svmLin-Results, include=TRUE}
SummaryDF%>%
kable(caption="Best performing Linear SVM Model.",digits=4)

```
79 individual Support Vector Machines with a linear kernel were built that using a progressive number of scores  along principal components directions (from 2 to 80) as predictors. 80 was the maximum number of principal components directions included in a single model because, as discussed in the section on PCA, nearly all of the variability of the data is included in the first 80 principal component directions. On each iteration of the model, the Linear SVM was tuned to the optimal cost using 10-fold cross validation. For each model, the optimal cost was searched for in a range between .01 and 100. Once the optimal CV Error was identified for each Linear SVM containing an increasing number of principal component directions, the best model was identified as the linear SVM that minimized the CV Error.  

Figure \@ref(fig:svmLin-CVPlot) shows the CV Error of each Linear SVM that was built. As shown in the figure, the CV Error, in general, trended downward in models from 2 principal components included to 24 principal components. As the number of principal components included in the linear SVMs were increased, the CV Error generally (but not absolutely) trended upward. \@ref(fig:svmLin-CostPlot) shows the optimal cost for each model. As is shown in the figure, the optimal cost lies within the range of costs that was searched in when tuning the model, so it is assumed that an optimal model was identified in all cases. 

The best Linear SVM model that was identified is summarized in \@ref(tab:svmLin-Results). The model included 24 principal components, had 32 support vectors, and had a CV Error of 11.8%. 

#### *Polynomial SVM*


```{r polySVMCode}
# This calls a script to run the Poly SVM. Code commented out because it takes a long time to run. Results were stored in a csv and brought in for plotting.
#source("SVM/SVM_Poly.R", local = knitr::knit_global())
```

```{r svmPoly-CVPlot, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Polynomial Support Vector Machine Optimal Cross Validation error for differing numbers of principal components included." }
#Plotting and Results for Poly SVM

svmPolyResultsDF <- read.csv("SVM/Results/PolySVM.csv")

BestPolyModel <- readRDS("SVM/BestSVMModels/BestPolyModel.rds")
BestPolyModel <- BestPolyModel[[1]]
numSupportVectors_polySVM <- nrow(BestPolyModel$SV)
optimalPC_polySVM <- c(svmPolyResultsDF$PrincipalComponents[which.min(svmPolyResultsDF$CV_MSE_polySVM)])
optimalCVError_polySVM <- c(min(svmPolyResultsDF$CV_MSE_polySVM))
optimalCost_polySVM <- c(svmPolyResultsDF$Best_Cost_polySVM[which.min(svmPolyResultsDF$CV_MSE_polySVM)])
optimalDegree_polySVM <- c(svmPolyResultsDF$Best_Degree[which.min(svmPolyResultsDF$CV_MSE_polySVM)])

SummaryDF <- data.frame(PC = optimalPC_polySVM , CV_Error = optimalCVError_polySVM, Cost = optimalCost_polySVM, Degree=optimalDegree_polySVM, SV = numSupportVectors_polySVM) 

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE_polySVM)) +
  geom_line()+
  geom_point()+
  labs(title="CV Error vs. Principal Components", y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
```


```{r svmPoly-CostPlot, include=TRUE,fig.width = 6, fig.height = 2.5, fig.cap="Optimal Cost identfied for each polynomial support vector machine of differing numbers of principal components. The red dashed lines represent the mininum and maximum of the search range used to tune the support vector machine."}

abs_min_cost_poly <- .0001
abs_max_cost_poly <- 10000

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Cost_polySVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost_poly, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost_poly, linetype="dashed", color = "red") +
  scale_y_log10()+
  labs(title= "Optimal Cost vs. Principal Components", y="Optimal Cost", x = "Number of Principal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```


```{r svmPoly-DegreePlot, include=TRUE,fig.width = 6, fig.height = 2.5, fig.cap="Optimal Degree identfied for each polynomial support vector machine of differing numbers of principal components."}

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Degree)) +
  geom_line()+
  geom_point()+
  labs(title = "Optimal Degree vs. Principal Components",y="Optimal Degree", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```


```{r svmPoly-Results, include=TRUE}
SummaryDF%>%
kable(caption="Best performing Polynomial SVM Model.",digits=4)

```

79 individual Support Vector Machines with a polynomial were built using a progressive number of scores along principal component direcitons (from 2 to 80) as predictors. Similarly to the linear SVM, 80 was the maximum number of principal components included in a single model because nearly all of the variability of the data is included in the first 80 principal component directions. On each iteration of the model, the Polynomial SVM was tuned to both the optimal cost and optimal degree using 10-fold cross validation. For each model, the optimal cost was searched for in a range between .001 and 1000, and the optimal degree was searched for between degrees 1 to 4. Higher degree polynomials were not included on the findings that higher degree Polynomial SVM's did not produce the optimal results and only served to increase computing times. Once the optimal CV Error was identified for each Polynomial SVM containing an increasing number of principal component directions, the best model was identified as the Polynomial SVM that minimized the CV Error.  

Figure \@ref(fig:svmPoly-CVPlot) shows the optimal CV Error of each Polynomial SVM that was built. As shown in the figure, the CV Error, in general, trended downward in models from 2 principal components included to 16 principal components. As the number of principal components included in the polynomial SVMs were increased, the CV Error generally remained relatively flat, which is different from the results of the linear SVM analysis that showed a clearer minimum.

\@ref(fig:svmPoly-CostPlot) shows the optimal cost for each model. As is shown in the figure, the optimal cost lies within the range of costs that was searched in when tuning the model, so it is assumed that an optimal model was identified in all cases. \@ref(fig:svmPoly-DegreePlot) shows the optimal degree for each Polynomial SVM. Interestingly, all models that included scores from more than 11 principal component directions had an optimal degree of 1, which suggests that as more principal component direction are added to the model, an SVM with a linear kernel perform better than an SVM with a higher order polynomial kernel. 

The best Polynomial SVM model that was identified is summarized in \@ref(tab:svmPoly-Results). The model included 48 principal components, had 48 support vectors, and had a CV Error of 10.7%. 


#### *Radial SVM*

```{r radSVMCode}
# This calls a script to run the Radial SVM
#source("SVM/SVM_rad.R", local = knitr::knit_global())
```

```{r svmRad-CVPlot, include=TRUE,fig.width = 6, fig.height = 3, fig.cap = "Radial Support Vector Machine Optimal Cross Validation error for differing numbers of principal components included." }

svmRadResultsDF <- read.csv("SVM/Results/radSVM.csv")
BestRadModel <- readRDS("SVM/BestSVMModels/BestRadModel.rds")
BestRadModel<- BestRadModel[[1]]
numSupportVectors_radSVM <- nrow(BestRadModel$SV)
optimalPC_radSVM <- c(svmRadResultsDF$PrincipalComponents_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])
optimalCVError_radSVM <- c(min(svmRadResultsDF$CV_MSE_radSVM))
optimalCost_radSVM <- c(svmRadResultsDF$Best_Cost_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])
optimalGamma_radSVM <- c(svmRadResultsDF$Best_Gamma_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])

SummaryDF <- data.frame(PC = optimalPC_radSVM , CV_Error = optimalCVError_radSVM, Cost = optimalCost_radSVM, Gamma = optimalGamma_radSVM, SV = numSupportVectors_radSVM) 

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=CV_MSE_radSVM)) +
  geom_line()+
  geom_point()+
  labs(title = "CV Error vs. Prinicipal Components",y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        plot.title = element_text(hjust = 0.5))


```

```{r svmRad-CostPlot, include=TRUE,fig.width = 6, fig.height = 2.5, fig.cap="Optimal Cost identfied for each radial support vector machine of differing numbers of principal components. The red dashed lines represent the mininum and maximum of the search range for cost used to tune the support vector machine."}

abs_min_cost <- .0001
abs_max_cost <- 10000

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Cost_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10()+
  labs(title= "Optimal Cost vs. Principal Components", y="Optimal Cost", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

```{r svmRad-GammaPlot, include=TRUE,fig.width = 6, fig.height = 2.5, fig.cap="Optimal Gamma identfied for each radial support vector machine of differing numbers of principal components. The red dashed lines represent the mininum and maximum of the search range for gamma used to tune the support vector machine."}

abs_min_gamma <- .0001
abs_max_gamma <-10000

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Gamma_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_gamma, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_gamma, linetype="dashed", color = "red") +
  scale_y_log10()+
  labs(title= "Optimal Gamma vs. Principal Components", y="Optimal Gamma", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))


```


```{r svmRad-Results, include=TRUE}
SummaryDF%>%
kable(caption="Best performing Radial SVM Model.",digits=4)
```

59 individual Support Vector Machines with a radial kernel were built using a progressive number of scores along principal components directions (from 2 to 60) as predictors. Unlike the linear and polynomial SVM analysis, 60, not 80, was the maximum number of principal components included in a single model. This change was due to computing limitations. It was determined that the 97% of the variability in the predictor data can be found in the first 60 principal component directions was sufficient to make conlusions on the best radial SVM. 

Each Radial SVM including differing numbers of principal component directions was tuned to both the optimal cost and optimal gamma using 10-fold cross validation. For each model, the optimal cost was searched for in a range between .001 and 10,000, and the optimal gammas was searched for between .001 to 1,000. Once the optimal CV Error was identified for each Radial SVM containing an increasing number of principal component directions, the best model was identified as the Radial SVM that minimized the CV Error.  

Figure \@ref(fig:svmRad-CVPlot) shows the optimal CV Error of each Radial SVM that was built. As shown in the figure, the CV Error, in general, trended downward in models from 2 principal component directions included to 30 principal component directions. As the number of principal component directions included in the Radial SVMs were increased from 30, the CV Error generally (but not absolutely) increased.

\@ref(fig:svmRad-CostPlot) shows the optimal cost that was identified for each model. As is shown in the figure, the optimal cost lies within the range of costs that was searched in when tuning the model. \@ref(fig:svmRad-GammaPlot) shows the optimal gamma that was identified for each model. As is shown in the figure, the optimal gammas lies within the range of gamma that was searched in when tuning the model. Because the optimal gamma and cost identfied for each model was within their respective search ranges, it is assumed that an optimal model was identified in all cases.

The best Radial SVM model that was identified is summarized in \@ref(tab:svmRad-Results). The model included 30 principal components, had 45 support vectors, and had a CV Error of 10.8%. 


### Summary of Categorical Models

A summary table of the best performing Logistic Regression Model, Linear SVM, Polynomial SVM, and Radial SVM is provided in \@ref(tab:catModelSummary). As shown in the table, simplest, yet worst, performing model was the Logistic Regression that included 9 principal component directions. The CV Error for the Logistic Regression model was more than 8% worse than that of any other model. 

The best performing model as measured by Cross Validation error was the Polynomial SVM that utilized the scores from 48 principal component directions. However, this model only outperformed the Radial SVM that utilized the scores from 30 principal component directions by .1%. Even though the Polynomial SVM performed slightly better, it did require the use of 18 more principal component directions, which is a lot more complexity in the model. In other words, the Radial SVM was able to perform almost just as well with significantly less complexity. Furthermore, the CV Error vs. Number of Principal Components plots for the Radial SVM (\@ref(fig:svmRad-CVPlot)) showed a much more distinct minimum of CV Error against the inlcuded number of principal components than that of the Polynomial SVM (\@ref(fig:svmPoly-CVPlot)) which is much more flat. This potentially suggests the the "better" results seen by the Polynomial SVM could have been more by chance than an actual optimal performance.

The optimal Radial SVM outperformed the Linear SVM with 24 principal component directions included by roughly 1%. Thus, by including 6 less principal component directions, the Linear SVM only performed slightly worse. Again, it is rather subjective, but it was decided that the added complexity of the Radial SVM was worth the 1% of accuracy gained.

In summation, the overall best model was the Radial SVM that included 30 principal component directions. This model was able, through PCA, to reduce the dimensional of the problem from 1,700 to only 30 and still perform with an accuracy of 89.2% of predicting USDA Grade from REIMs data. Overall, this is a good model and could potentially see real world application!

```{r catModelSummary, include=TRUE}
ModelNames <- c('Logistic Regression', 'Linear SVM', 'Polynomial SVM', 'Radial SVM')
PrincipalComponents <- c(optimalPC_Log, optimalPC_linSVM, optimalPC_polySVM, optimalPC_radSVM)
CV_Error <- c(optimalCVError_Log,  optimalCVError_linSVM, optimalCVError_polySVM,optimalCVError_radSVM)

SummaryDF <- data.frame(Model = ModelNames,PrincipalComponents = PrincipalComponents, CV_Error = CV_Error)

SummaryDF%>%
kable(caption="Best performing USDA Grade Models.",digits=4)
```

## Regression Prediction of Tenderness at 3, 14 and 28 Days


```{r, echo = FALSE}
#Here are the names of the data sets for all the following models. They correspond to a 60/40 training/test subsets.
ten3DF_train
ten3DF_test
ten14DF_train
ten14DF_test
ten28DF_train
ten28DF_test
```

### Best subset selection
```{r BestSubsetSelectionCode}
#Tenderness at day 3
set.seed(1)
reg.fwd <- regsubsets(Ten3~., data = ten3DF_train, nvmax = 40,method = "forward") 
#Best subset selection using foward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten3~., data=ten3DF_test)
val.errors <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors[i] <- mean((ten3DF_test$Ten3-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 3",type = "o")

MSE_subset_day3<-min(val.errors)

#Tenderness at day 14
reg.fwd <- regsubsets(Ten14~., data = ten14DF_train, nvmax = 40,method = "forward") 
#Best subset selection using foward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten14~., data=ten14DF_test)
val.errors1 <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors1[i] <- mean((ten14DF_test$Ten14-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors1, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 14",type = "o")

MSE_subset_day14<-min(val.errors1)

#Tenderness at day 28
reg.fwd <- regsubsets(Ten28~., data = ten28DF_train, nvmax = 40,method = "forward") 
#Best subset selection using forward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten28~., data=ten28DF_test)
val.errors2 <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors2[i] <- mean((ten28DF_test$Ten28-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 28",type = "o")

MSE_subset_day28<-min(val.errors2)
```


### PCR
```{r PCRCode}
#Fitting the PCR Ten3 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten3<- pcr (Ten3~., data=ten3DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten3)
mse3 <- MSEP(pcr.fit_ten3)
#M of the smallest CV error
mse3$comps[which.min(as.data.frame(mse3$val)[1,])]
#Calculating MSE with M=6 and the Test set.
pcr.pred <- predict (pcr.fit_ten3, ten3DF_test, ncomp = 6)
pcr.pred-ten3DF_test$Ten3
MSE_PCR_day3<-mean ((pcr.pred - ten3DF_test$Ten3)^2)
#Error compare to average
sqrt(MSE_PCR_day3)
mean(ten3DF$Ten3)


#Fitting the PCR Ten14 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten14<- pcr (Ten14~., data=ten14DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten14)
mse14 <- MSEP(pcr.fit_ten14)
#M of the smallest CV error
mse14$comps[which.min(as.data.frame(mse14$val)[1,])]
#Calculating MSE with M=1 and the Test set.
pcr.pred <- predict (pcr.fit_ten14, ten14DF_test, ncomp = 1)
pcr.pred-ten14DF_test$Ten14
MSE_PCR_day14<-mean ((pcr.pred - ten14DF_test$Ten14)^2)
#Error in the same scale
sqrt(MSE_PCR_day14)
mean(ten14DF$Ten14)



#Fitting the PCR Ten28 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten28<- pcr (Ten28~., data=ten28DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten28)
mse28 <- MSEP(pcr.fit_ten28)
#M of the smallest CV error
mse28$comps[which.min(as.data.frame(mse28$val)[1,])]
size_PCR_day28<-which.min(as.data.frame(mse28$val)[1,])
#Calculating MSE with M=2 and the Test set.
pcr.pred <- predict (pcr.fit_ten28, ten28DF_test, ncomp = 2)
pcr.pred-ten28DF_test$Ten28
MSE_PCR_day28<-mean ((pcr.pred - ten28DF_test$Ten28)^2)
#Error in the same scale
sqrt(MSE_PCR_day28)
mean(ten28DF$Ten28)


```


### Lasso Regression

#### Standard Lasso Regression

```{r LassoNonPCACode}

#Lasso on Ten3 (Non-PCA)

# THIS MODEL IS NO BETTER THAN AVERAGE!!!!
set.seed(1)

X_train <- model.matrix(Ten3 ~ ., data = ten3DF_train)[,-1]
Y_train <- ten3DF_train$Ten3
X_test <- model.matrix(Ten3 ~ ., data = ten3DF_test)[,-1]
Y_test <- ten3DF_test$Ten3
lambda <- 10^seq(-2, 3, length.out = 1000)
cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

lambdaPlot3 <- ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()+
  labs(title= "Optimal Lambda vs. Principal Components for Tenderness at 3 Days Postmortem", y="CV MSE", x = "Lambda") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

lambdaMin_3 <- cv_lasso$lambda.min
cv_error_min_3 <- min(cv_lasso$cvm)

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin_3)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin_3, type="response")

testErrorLasso_3 <- mean(( predictions[,'s1'] - Y_test )^2)

predictionPlot3 <- ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) 

coefs_3 <- coef(lasso_mdl)
size_3 <- length(coefs_3[coefs_3[,1]!=0,]) - 1

Ten3NonPCA_Results <- data.frame(CV_MSE = cv_error_min_3, Lambda = lambdaMin_3, TestError = testErrorLasso_3, Size = size_3)
Ten3NonPCA_Results


#Lasso on Ten14 (Non-PCA)

# THIS MODEL IS NO BETTER THAN AVERAGE!!!!
set.seed(1)

X_train <- model.matrix(Ten14 ~ ., data = ten14DF_train)[,-1]
Y_train <- ten14DF_train$Ten14
X_test <- model.matrix(Ten14 ~ ., data = ten14DF_test)[,-1]
Y_test <- ten14DF_test$Ten14
lambda <- 10^seq(-2, 3, length.out = 1000)
cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

lambdaPlot14 <- ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()+
  labs(title= "Optimal Lambda vs. Principal Components for Tenderness at 14 Days Postmortem",y="CV MSE", x = "Lambda") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

lambdaMin_14 <- cv_lasso$lambda.min
cv_error_min_14 <- min(cv_lasso$cvm)

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin_14)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin_14, type="response")

testErrorLasso_14 <- mean(( predictions[,'s1'] - Y_test )^2)

predictionPlot14 <- ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) 

coefs_14 <- coef(lasso_mdl)
size_14 <- length(coefs_14[coefs_14[,1]!=0,]) - 1

Ten14NonPCA_Results <- data.frame(CV_MSE = cv_error_min_14, Lambda = lambdaMin_14, TestError = testErrorLasso_14, Size = size_14)
Ten14NonPCA_Results

#Lasso on Ten28 (Non-PCA)
set.seed(1)

X_train <- model.matrix(Ten28 ~ ., data = ten28DF_train)[,-1]
Y_train <- ten28DF_train$Ten28
X_test <- model.matrix(Ten28 ~ ., data = ten28DF_test)[,-1]
Y_test <- ten28DF_test$Ten28
lambda <- 10^seq(-2, 3, length.out = 1000)
cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

lambdaPlot28 <- ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()+
  labs(title= "Optimal Lambda vs. Principal Components for Tenderness at 28 Days Postmortem", y="CV MSE", x = "Lambda") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

lambdaMin_28 <- cv_lasso$lambda.min
cv_error_min_28 <- min(cv_lasso$cvm)

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin_28)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin_28, type="response")

testErrorLasso_28 <- mean(( predictions[,'s1'] - Y_test )^2)

predictionPlot28 <- ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) 

coefs_28 <- coef(lasso_mdl)
size_28 <- length(coefs_28[coefs_28[,1]!=0,]) - 1

Ten28NonPCA_Results <- data.frame(CV_MSE = cv_error_min_28, Lambda = lambdaMin_28, TestError = testErrorLasso_28, Size = size_28)
Ten28NonPCA_Results

write.csv(Ten3NonPCA_Results,"Lasso/LassoNonPCA/Ten3.csv", row.names = FALSE)
write.csv(Ten14NonPCA_Results,"Lasso/LassoNonPCA/Ten14.csv", row.names = FALSE)
write.csv(Ten28NonPCA_Results,"Lasso/LassoNonPCA/Ten28.csv", row.names = FALSE)

```




```{r LassoNonPCA-Lambda3Plot, include=TRUE,fig.width = 6, fig.height = 2, fig.cap="CV MSE associated with differening values for lambda in for the Tenderness at Day 3 Lasso Model"}

lambdaPlot3

```
```{r LassoNonPCA-Lambda14Plot, include=TRUE,fig.width = 6, fig.height = 2, fig.cap="CV MSE associated with differening values for lambda in for the Tenderness at Day 14 Lasso Model"}

lambdaPlot14

```

```{r LassoNonPCA-Lambda28Plot, include=TRUE,fig.width = 6, fig.height = 2, fig.cap="CV MSE associated with differening values for lambda in for the Tenderness at Day 28 Lasso Model"}

lambdaPlot28

```


```{r LassoNonPCA-summary, include=TRUE}
#Report the Lasso results
days <- c(3, 14, 28)
cv_mse <- c(cv_error_min_3,cv_error_min_14,cv_error_min_28)
test_error <- c(testErrorLasso_3,testErrorLasso_14,testErrorLasso_28)
size <- c(size_3,size_14,size_28)
lambda <- c(lambdaMin_3,lambdaMin_14,lambdaMin_28)


SummaryLassoNonPCADF <- data.frame(Day = days, Test_MSE=cv_mse, Lambda = lambda, Size = size)
SummaryLassoNonPCADF%>%
kable(caption="Summary of Lasso models for tenderness at days 3, 14, and 28",digits=4)


```

Three Lasso Regression model were built to predict beef tenderness on the 83 samples at 3, 14, and 28 days postmortem. Each of the models were developed using 10-fold cross validation to tune the parameter lambda on a training dataset that represented 51 of the 83 samples. Once the models were trained optimally, they were used to predict the remained samples to predict the test error. 

\@ref(fig:LassoNonPCA-Lambda3Plot), \@ref(fig:LassoNonPCA-Lambda14Plot), \@ref(fig:LassoNonPCA-Lambda28Plot) show how the Cross Validation Mean Squared Error (MSE) of the Tenderness at Day 3, 14, and 28 (respectively) changed with differing values of lambda. To tune the model, the value of lambda was identified that minimized the CV MSE for each model.

Using these tuned models, the Test MSE was calculated on the data left out of the Training set. \@ref(tab:LassoNonPCA-summary) provides a summary of the three models. As shown in the table, the Tenderness at 3 days model shows a Test MSE of 19.05 and it only included one predictor in the model. The tenderness at 14 days model shows a Test MSE of 15.36 and it included no predictors in the model. That is to say, the model only includes an intercept that is the average of the training values and the Lasso algorithm could not identify a model that was simply better than taking the average of the Training dataset tenderness at day 14. Lastly, the tenderness at 28 days model has a Test MSE of 6.61 and included 18 predictors in the model. A more in depth summary of how valuable these test errors are will be include below in the summary section of the regression analysis. 

#### Lasso using Principal Components Regression

```{r LassoTen3Code}
# Lasso on principal components for Ten3
# This code is commented out because it takes a while to run. Results are stored in files.
#source("Lasso/Lasso_ten3_CV.R", local = knitr::knit_global())
```

```{r}
lambda_min_under <- -2
lambda_max_over <- 8

LassoTen3Results <- read.csv("Lasso/Results/Ten3.csv")
BestTen3LassoModel <- readRDS("Lasso/BestModels/BestLassoTen3Model.rds")
BestTen3LassoModel <- BestTen3LassoModel[[1]]
optimalPC_LassoPCATen3 <- LassoTen3Results$PrincipalComponents[which.min(LassoTen3Results$CV_MSE)]
optimal_TestError_LassoPCATen3 <-  min(LassoTen3Results$CV_MSE)
optimalLambda_LassoPCATen3 <- LassoTen3Results$Best_Lambda[which.min(LassoTen3Results$CV_MSE)]
optimalSize_LassoPCATen3 <- LassoTen3Results$Size[which.min(LassoTen3Results$CV_MSE)]

summaryDF_LassoTen3PCA <- data.frame(PrincipalComponents = optimalPC_LassoPCATen3, CV_Error = optimal_TestError_LassoPCATen3, Lambda = optimalLambda_LassoPCATen3, Size = optimalSize_LassoPCATen3)


MSE_Lasso_day3<-optimal_TestError_LassoPCATen3
```

```{r LassoTen14Code}
# Lasso on principal components for Ten14
# This code is commented out because it takes a while to run. Results are stored in files.
# source("Lasso/Lasso_ten14_CV.R", local = knitr::knit_global())
```

```{r}

lambda_min_under <- -4
lambda_max_over <- 8

LassoTen14Results <- read.csv("Lasso/Results/Ten14.csv")
BestTen14LassoModel <- readRDS("Lasso/BestModels/BestLassoTen14Model.rds")
BestTen14LassoModel <- BestTen14LassoModel[[1]]
optimalPC_LassoPCATen14 <- LassoTen14Results$PrincipalComponents[which.min(LassoTen14Results$CV_MSE)]
optimal_TestError_LassoPCATen14 <-  min(LassoTen14Results$CV_MSE)
optimalLambda_LassoPCATen14 <- LassoTen14Results$Best_Lambda[which.min(LassoTen14Results$CV_MSE)]
optimalSize_LassoPCATen14 <- LassoTen14Results$Size[which.min(LassoTen14Results$CV_MSE)]

summaryDF_LassoTen14PCA <- data.frame(PrincipalComponents = optimalPC_LassoPCATen14, CV_Error = optimal_TestError_LassoPCATen14, Lambda = optimalLambda_LassoPCATen14, Size = optimalSize_LassoPCATen14)


MSE_Lasso_day14<-optimal_TestError_LassoPCATen14

```

```{r LassoTen28Code}
# Lasso on principal components for Ten28
# This code is commented out because it takes a while to run. Results are stored in files.
# source("Lasso/Lasso_ten28_CV.R", local = knitr::knit_global())
```

```{r}

lambda_min_under <- -1 
lambda_max_over <- 10

LassoTen28Results <- read.csv("Lasso/Results/Ten28.csv")
BestTen28LassoModel <- readRDS("Lasso/BestModels/BestLassoTen28Model.rds")
BestTen28LassoModel <- BestTen28LassoModel[[1]]
optimalPC_LassoPCATen28 <- LassoTen28Results$PrincipalComponents[which.min(LassoTen28Results$CV_MSE)]
optimal_TestError_LassoPCATen28 <-  min(LassoTen28Results$CV_MSE)
optimalLambda_LassoPCATen28 <- LassoTen28Results$Best_Lambda[which.min(LassoTen28Results$CV_MSE)]
optimalSize_LassoPCATen28 <- LassoTen28Results$Size[which.min(LassoTen28Results$CV_MSE)]

summaryDF_LassoTen28PCA <- data.frame(PrincipalComponents = optimalPC_LassoPCATen28, CV_Error = optimal_TestError_LassoPCATen28, Lambda = optimalLambda_LassoPCATen28, Size = optimalSize_LassoPCATen28)


MSE_Lasso_day28<-optimal_TestError_LassoPCATen28
```

```{r LassoPCA-summary, include=TRUE}
#Summary of Lasso PCA Analysis table
days <- c(3,14,28)
principalComponents <- c(optimalPC_LassoPCATen3,  optimalPC_LassoPCATen14, optimalPC_LassoPCATen28)
CV_MSE <- c(optimal_TestError_LassoPCATen3, optimal_TestError_LassoPCATen14, optimal_TestError_LassoPCATen28)
Lambda <- c(optimalLambda_LassoPCATen3, optimalLambda_LassoPCATen14,optimalLambda_LassoPCATen28)
Size <- c(optimalSize_LassoPCATen3, optimalSize_LassoPCATen14, optimalSize_LassoPCATen28)

summaryLassoPCADF <- data.frame(Day = days, CV_MSE =CV_MSE, PrincipalComponents = principalComponents, Lambda = Lambda, Size = Size)
summaryLassoPCADF%>%
kable(caption="Summary of Lasso models using PCA for tenderness at days 3, 14, and 28",digits=4)

```

```{r LassoPCATen3-CVPlot, include=TRUE,fig.width = 6, fig.height = 3, fig.cap="Cross Validation Mean Squared Error for the Lasso Models that were built using differing numbers of principal components to predict tenderness at three days."}
#Because Ten3 Lasso with PCA had the only meaningful results, show a little more of these results

lambda_min_under <- -2
lambda_max_over <- 8

ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
```


```{r LassoPCATen3-Size, include=TRUE, fig.width = 6, fig.height = 2.5, fig.cap="The number of predictors for tenderness at 3 days that were included in the Lasso Model given the number of principal components that were used to generate the Lasso Model."}
ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = Size)) +
  geom_line() +
  geom_point()+
  labs(y="Size", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
```

```{r LassoPCATen3-Lambda, include=TRUE, fig.width = 6, fig.height = 2.5, fig.cap="The optimal lambda of the Lasso Models used to predict tenderness at 3 days given the number of principal components that were used to generate the Lasso Model."}
ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  scale_y_log10()+
  labs(y="Optimal Lambda", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

A second attempt at Lasso Regression was made utilizing the principal components to answer the question of if there be a direction of variability along a principal component that is unrelated to tenderness at 3, 14, and 28 days. 

For each of the tenderness at 3, 14, and 28 day datasets, a process similar to that which was used for the categorical analysis was performed here. Several Lasso models were created using, in each, a progressive number of principal component directions (from 2 to 40). Each of these lasso models utilized the full dataset and reported a 10-Fold Cross Validation MSE error that was used for model evaluation. Furthermore, each of these models were tuned to use the appropriate value of lambda. The results for this analysis are provided in  \@ref(tab: LassoPCA-summary).  As shown in the table, the optimal value for the tenderness at 14 and 28 day models had a size of zero. That is to say, it was no better than the average values of the responses. The tenderness at day 3 model did have a non-zero size. It used 18 principal components, determined 12 non-zero coefficients for these 18 principal component directions, and it resulted in a CV Error of 17.93.

\@ref(tab:LassoPCATen3-CVPlot) shows the results for CV MSE vs. number of principle components included in the model for tenderness at day 3. As shown in the plot, there is a clear minimum of the CV MSE in the model with 18 principal components. \@ref(tab:LassoPCATen3-Lambda).



### Splines
        
```{r SplinesCode}
# Spline model for Tenderness at day 3
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
ten3<-na.omit( dat1[,-c(1,3:4)])
ten3_train<-ten3[dpart,]
ten3_test<-ten3[-dpart,]
err3<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten3~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten3_train)
preds<- predict(gam, newdata = ten3_test)
preds
err3[i]<-mean((ten3_test$Ten3 - preds)^2)
}
Spline3<-data.frame(N.comp=seq(1,10),MSE=err3)
which.min(Spline3$MSE)
min(Spline3$MSE)


# Spline model for Tenderness at day 14
set.seed(1)
ten14<-na.omit( dat1[,-c(1:2,4)])
ten14_train<-ten14[dpart,]
ten14_test<-ten14[-dpart,]
err14<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten14~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten14_train)
preds<- predict(gam, newdata = ten14_test)
preds
err14[i]<-mean((ten14_test$Ten14 - preds)^2)
}
Spline14<-data.frame(N.comp=seq(1,10),MSE=err14)
which.min(Spline14$MSE)
min(Spline14$MSE)

# Spline model for Tenderness at day 28
set.seed(1)
ten28<-na.omit( dat1[,-c(1:3)])
ten28_train<-ten28[dpart,]
ten28_test<-ten28[-dpart,]
err28<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten28~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten28_train)
preds<- predict(gam, newdata = ten28_test)
preds
err28[i]<-mean((ten28_test$Ten28 - preds)^2)
}
Spline28<-data.frame(N.comp=seq(1,10),MSE=err28)
which.min(Spline28$MSE)
min(Spline28$MSE)

```

### Summary of Regression Models
``` {r RegModelSummary,include=TRUE}
#Summary results of regression

data.frame(  Model = "Best Subset", 
             Size_day3 = "1", MSE_day3=MSE_subset_day3,
             Size_day14 = "1", MSE_day14=MSE_subset_day14,
             Size_day28 = "1", MSE_day28=MSE_subset_day28)%>%
  bind_rows(data.frame(Model = "PCA-Lasso", 
            Size_day3 =toString(optimalSize_LassoPCATen3), MSE_day3=optimal_TestError_LassoPCATen3, 
            Size_day14 =toString(optimalSize_LassoPCATen14), MSE_day14=optimal_TestError_LassoPCATen14, 
            Size_day28 =toString(optimalSize_LassoPCATen28), MSE_day28=optimal_TestError_LassoPCATen28))%>%
  bind_rows(data.frame(Model = "PCR", 
            Size_day3 ="6", MSE_day3= MSE_PCR_day3, 
            Size_day14 ="1", MSE_day14= MSE_PCR_day14, 
            Size_day28 ="2", MSE_day28= MSE_PCR_day28)) %>%
  bind_rows(data.frame(Model = "PCA-Splines", 
            Size_day3 ="5", MSE_day3= min(Spline3$MSE), 
            Size_day14 ="2", MSE_day14= min(Spline14$MSE), 
            Size_day28 ="4", MSE_day28= min(Spline28$MSE))) %>%
  bind_rows(data.frame(Model = "Lasso", 
          Size_day3 =toString(size_3), MSE_day3= testErrorLasso_3, 
          Size_day14 =toString(size_14), MSE_day14= testErrorLasso_14, 
          Size_day28 =toString(size_28), MSE_day28= testErrorLasso_28)) %>% 
kable(caption="Table 2. Mean Squared Error (MSE) of regression models for tenderness at day 3, 14, and 28",digits=2)




```

# CONCLUSION