---
title: "Final project"
author: "Group 5"
date: "11/25/2021"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
```

### Reproducibility
```{r}
set.seed(1)
```

### Uploading data
```{r}
REIMS <- read.csv("REIMS machine learning class.csv")

dropBase <- c("SampleID", "Cx", "GS_1", "GS_2", "UNIQ_ID", "Ten42", "Ten56", "Ten70")
responseVars <- c("USDA.grade", "Ten3", "Ten14","Ten28")

# Define the associated dataframes for each analysis. Respnose variable is column 1, predictors are the remaining columns
gradeDF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("USDA.grade"))))]
ten3DF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))]
ten14DF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))]
ten28DF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))]

gradeDF[,1:5]
ten3DF[,1:5]
ten14DF[,1:5]
ten28DF[,1:5]
```


### Prediction of USDA grade using REIMS (classification models)  
        a) Dimensionality reduction using Principal Component Analysis (PCA)
```{r}
#CODE BY PATRICK
# This section of code performs PCA on the predictors of the data set. It stores the loading vectors in the pca_loadings dataframe
# for use in the models below. It also provides a plot of Proportion of Variance Explained (PVE) vs. Principal Component.

gradeResponse <- as.factor(gradeDF$USDA.grade)
data <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
pca_obj <- prcomp(data, center=TRUE, scale=TRUE)  # perform PCA
pca_summary <- summary(pca_obj)
pca_loadings <- pca_obj$rotation #store the loading vectors for future use

pca_summary

PVE <- pca_summary$importance[2,] # store the proportion of variance explained by each PC
PrincipalComponent <- seq(1, length(PVE), 1) # Vector of Principle components for plot

#PLOT OF PVE VS. Principle Component
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=PVE))+
  geom_point()

```

        b) Dimensionality reduction using Recursive Feature Elimination (FS)
```{r}
FS_REIMS<-read.csv("REIMS_FS_data.csv")
```

        c) Logistic regression with PCA data
```{r}
# CODED BY PATRICK

#This section of code performs 10-fold cross validation on a logistic regression using different number of principal components (varying from 
# 1 principal component to 16 principal components). The optimal number of principal components to use for the logistic regression is identified
# and used to build a final logistic model on the full dataset. A graph of CV_Error vs. Number of Principal Components is provided, and a data
# frame that reports the optimal number of principal components to use, it's CV_Error, and the training error using the full dataset.

set.seed(1)

predictorsDF <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
gradeResponse <- as.factor(gradeDF$USDA.grade)

max_components <- 16 #maximum number of components to use in cross validation
n <- nrow(predictorsDF) 
k<-10 # number of folds
p <- .5 #probability threshold to bin logistic regression response

cv_labels <- sample(seq_len(k), n, replace = TRUE) #labels for folds
cv_mse <- rep(NA, max_components) # place holder vector to record cv_mse for each PC

#loop over each number of principle components
for (i in seq_len(max_components)) {
    num_components <- i
    
    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    #compute the scores (z-values) for the given number of principal components
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
    scoresDF$USDA.grade <- gradeResponse
    
    cv_mse_i <- rep(NA, k) 
    
    #perform 10-fold cross validation
    for(j in seq_len(k)) {
        train <- scoresDF[cv_labels != j,]
        test <- scoresDF[cv_labels == j,]
        
        fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = train)
        pred <- predict(fit, newdata = test, type="response")
        confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=test$USDA.grade)
        
        #it was found that sometimes the predicted response only categorized one response, not two
        # so this error handling was necessary to get the appropriate error rate.
        if ( nrow(confMatrix) == 2) {
          errorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  
        } else {
          errorRate <- 1 - (confMatrix[1,1])/ sum(confMatrix)
        }
        
        cv_mse_i[j] <- errorRate
      }
    
    cv_mse[[i]] <- mean(cv_mse_i) #store the average CV-error of the 10 folds
}

#PLOT OF K-Fold CV_MSE VS. Principle Component
PrincipalComponent <- seq_len(max_components)
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=cv_mse))+
  geom_point()

#store the optimal cv error rate associated with the optimal number of principal components
optimal_CVerror <- min(cv_mse)
optimal_PC <- which.min(cv_mse)


#train a model with the optimal number of principal components and report
loadingDF <- pca_loadings[,1:optimal_PC]
scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
scoresDF$USDA.grade <- gradeResponse
fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = scoresDF)
pred <- predict(fit,type="response")
confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=scoresDF$USDA.grade)
TrainErrorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  

data.frame(optimalNumPrincpleCompoents = c(optimal_PC), optimalCVerror = c(optimal_CVerror), trainError = c(TrainErrorRate))

```

        d) Logistic regression with FS data
```{r}

```

        e) K-nearest neighbors using PCA data
```{r}

```

        f) K-nearest neighbors using FS data
```{r}

```

        g) PCA using PCA data
```{r}

```

        h) PCA using FS data
```{r}

```

        i) Support vector machine (SVM) using PCA data
        
            
            iA) Linear SVM
```{r}
set.seed(1)

#convert USDA.grade to factor
gradeDF$USDA.grade <- as.factor(gradeDF$USDA.grade)

min_components <- 15 # it was found that components less than 15 did not perform optimially
max_components <- 80
component_seq <- seq(min_components,max_components)

best_cost <- rep(NA, length(component_seq))
cv_mse <- rep(NA, length(component_seq))

costVec <- seq(from = .01, to = 2.5, length.out = 50)

for (i in seq(length(component_seq))) {
    num_components <- component_seq[i]

    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF)) #compute the scores (z-values) for the given number of principal components
    scoresDF$USDA.grade <- gradeDF$USDA.grade
    
    TuneObj <- tune.svm(USDA.grade~ ., data = scoresDF, kernel = "linear", cost = costVec)
    
    best_cost[i] <- TuneObj$best.parameters[1,1]
    cv_mse[i] <- TuneObj$best.performance
}

PrincipalComponent <- component_seq

linSVM_Results <- data.frame(PrincipalComponents = PrincipalComponent, CV_MSE = cv_mse, Best_Cost = best_cost)

write.csv(linSVM_Results,"SVM_results/PCA/LinSVM.csv", row.names = FALSE)




```


```{r}
#Linear SVM Continued



svmLinResultsDF <- read.csv("SVM_results/PCA/LinSVM.csv")

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents, y=CV_MSE)) +
  geom_line()+
  geom_point()

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents, y=Best_Cost)) +
  geom_line()+
  geom_point()


data.frame(optimalPC = c(svmLinResultsDF$PrincipalComponents[which.min(svmLinResultsDF$CV_MSE)]), optimalCVError = c(min(svmLinResultsDF$CV_MSE)))

```



            iB) Polynomial SVM with PC

```{r}

set.seed(1)

#convert USDA.grade to factor
gradeDF$USDA.grade <- as.factor(gradeDF$USDA.grade)

min_components <- 1 # it was found that components less than 15 did not perform optimally
max_components <- 50
component_seq <- seq(min_components,max_components)
degreeVec <- seq(1,7)

best_cost <- rep(NA, length(component_seq))
cv_mse <- rep(NA, length(component_seq))



for (i in seq(length(component_seq))) {
    num_components <- component_seq[i]

    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF)) #compute the scores (z-values) for the given number of principal components
    scoresDF$USDA.grade <- gradeDF$USDA.grade
    

    costVec <- seq(from = .01, to = 10, length.out =10)

    TuneObj <- tune.svm(USDA.grade~ ., data = scoresDF, kernel = "polynomial", degree=1, scale=FALSE, cost = costVec)
    
    best_cost[i] <- TuneObj$best.parameters[1,1]
    cv_mse[i] <- TuneObj$best.performance
}

PrincipalComponent <- seq(min_components,max_components)

polySVM_Results <- data.frame(PrincipalComponents = PrincipalComponent, CV_MSE = cv_mse, Best_Cost = best_cost)

write.csv(polySVM_Results,"SVM_results/PCA/PolySVM.csv", row.names = FALSE)



```

```{r}

svmPolyResultsDF <- read.csv("SVM_results/PCA/PolySVM.csv")

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE)) +
  geom_line()+
  geom_point()

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Cost)) +
  geom_line()+
  geom_point()

```


            iB) Radial SVM with PC

```{r}

set.seed(1)

#convert USDA.grade to factor
gradeDF$USDA.grade <- as.factor(gradeDF$USDA.grade)

min_components <- 5 # it was found that components less than 15 did not perform optimially
max_components <- 50
component_seq <- seq(min_components,max_components)

best_cost <- rep(NA, length(component_seq))
best_gamma <- rep(NA, length(component_seq))
cv_mse <- rep(NA, length(component_seq))

costVec <- seq(from = .01, to = 15, length.out = 10)
gammaVec <- seq(from = .0001, to = .2, length.out = 20)

for (i in seq(length(component_seq))) {
    num_components <- component_seq[i]

    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF)) #compute the scores (z-values) for the given number of principal components
    scoresDF$USDA.grade <- gradeDF$USDA.grade
    
    TuneObj <- tune.svm(USDA.grade~ ., data = scoresDF, kernel = "radial", cost = costVec, gamma = gammaVec)
    
    best_cost[i] <- TuneObj$best.parameters[1,2]
    best_gamma[i] <- TuneObj$best.parameters[1,1]
    cv_mse[i] <- TuneObj$best.performance
}

PrincipalComponent <- component_seq

radSVM_Results <- data.frame(PrincipalComponents = PrincipalComponent, CV_MSE = cv_mse, Best_Cost = best_cost, Best_Gamma = best_gamma)

write.csv(radSVM_Results,"SVM_results/PCA/radSVM.csv", row.names = FALSE)


```


```{r}


svmRadResultsDF <- read.csv("SVM_results/PCA/radSVM.csv")

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents, y=CV_MSE)) +
  geom_line()+
  geom_point()

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents, y=Best_Cost)) +
  geom_line()+
  geom_point()

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents, y=Best_Gamma)) +
  geom_line()+
  geom_point()



data.frame(optimalPC = c(svmRadResultsDF$PrincipalComponents[which.min(svmRadResultsDF$CV_MSE)]), optimalCVError = c(min(svmRadResultsDF$CV_MSE)))

```


        j) SVM using FS data
```{r}


```


        k) PLS-DA
```{r}

```

        l) PLS-DA
```{r}

```

### Prediction of Tenderness over time using 36h REIMS (regression models)

        Best subset selection
```{r}
#test
```




