---
title: "Evaluation of Rapid Evaporative Ionization Mass Spectrometry to predict beef tenderness and USDA grade"
author: "Patrick Noe, Logan Perkins, Michael Hernandez"
date: "12/05/2021"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(include = FALSE)
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
library(FactoMineR)
library(kableExtra)
```

### INTRODUCTION
 * Beef palatability depends on tenderness, juiciness, and flavor. In the United States, there is a quality grading system that classifies beef carcasses on different quality categories (USDA). Some of the most common USDA quality grades are Prime, Choice, and Select, with Prime the best quality and Select being the lowest quality of the three classifications. The USDA quality grade is based on animal maturity and marbling level (intramuscular fat), and it is correlated with the overall consumer acceptance (USDA). However, tenderness is not predicted with USDA quality grade. Beef industry stakeholders have been searching for a non-destructive, real-time instrument for determining the tenderness of beef for nearly 40 years. Although some technologies allow the determination of beef tenderness, they require sacrificing a loin steak and cannot be performed in real-time. Rapid Evaporative Ionization Mass Spectrometry (REIMS) is a novel environmental mass spectrometry technique that combines a sample vaporization device with a time-of-flight mass spectrometer. When a sampling device, which uses a high-frequency current, comes into contact with the sample surface, an aerosol of molecules is generated and transformed into a mass spectral fingerprint. This technology has the advantage that it can be used in-situ and in the real-time classification of biological tissues without any sample preparation. Therefore, the goal of this study was to evaluate the ability of REIMS to predict USDA quality grade and beef tenderness.

```{r}
### Reproducibility
set.seed(1)
```


```{r}
### Uploading data
REIMS <- read.csv("REIMS machine learning class.csv")

dropBase <- c("SampleID", "Cx", "GS_1", "GS_2", "UNIQ_ID", "Ten42", "Ten56", "Ten70")
responseVars <- c("USDA.grade", "Ten3", "Ten14","Ten28")

# Define the associated dataframes for each analysis. Response variable is column 1, predictors are the remaining columns
gradeDF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("USDA.grade"))))]
```

### METHODOLOGY
...
``` {r}
#Creating training and test subsets for regressions
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
train<- REIMS[dpart, ]
test<- REIMS[-dpart, ]

ten3DF_train<- na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten3DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten14DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
ten28DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])

# Completed datasets for regressions
ten3DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
```

### RESULTS

# Data visualization using Principal Component Analysis (PCA)  
  There is not a evident cluster by USDA grade and tenderness at day 3, 14, and 28 days
```{r, include=TRUE,fig.width = 8, fig.height = 4}
#Performing PCA
pca1 <- PCA(REIMS[,-c(1:12)], ncp = 100, scale.unit=TRUE ,graph = F)
dat1<-REIMS[,6]
#Building data frame
dat1 <- data.frame(USDA.grade=dat1,Ten3=REIMS[,7],Ten14=REIMS[,8],Ten28=REIMS[,9],pca1$ind$coord[, c(1:50)])
#Plotting PCA faced by USDA grade
p <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=USDA.grade)) +
  geom_point()
print(p + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Comp 2", x = "Comp 1"))

#Plotting PCA faced by USDA grade
p1 <- ggplot(data = dat1, aes(x = Dim.2, y = Dim.3,colour=USDA.grade)) +
  geom_point()
print(p1 + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Comp 2", x = "Comp 1"))
```

```{r, include=TRUE, fig.width = 8, fig.height = 4}
#Plotting PCA colored by tenderness at day 3
p2 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten3)) +
  geom_point()
print(p2 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))

#Plotting PCA colored by tenderness at day 14
p3 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten14)) +
  geom_point()
print(p3 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))

#Plotting PCA colored by tenderness at day 28
p4 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten28)) +
  geom_point()
print(p4 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))
```


### Prediction of USDA grade using REIMS (classification models)  
        a) Dimensionality reduction using Principal Component Analysis (PCA)
```{r}
#CODE BY PATRICK
# This section of code performs PCA on the predictors of the data set. It stores the loading vectors in the pca_loadings dataframe
# for use in the models below. It also provides a plot of Proportion of Variance Explained (PVE) vs. Principal Component.

gradeResponse <- as.factor(gradeDF$USDA.grade)
data <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
pca_obj <- prcomp(data, center=TRUE, scale=TRUE)  # perform PCA
pca_summary <- summary(pca_obj)
pca_loadings <- pca_obj$rotation #store the loading vectors for future use

pca_summary

PVE <- pca_summary$importance[2,] # store the proportion of variance explained by each PC
PrincipalComponent <- seq(1, length(PVE), 1) # Vector of Principle components for plot

#PLOT OF PVE VS. Principle Component
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=PVE))+
  geom_point()

```

        b) Dimensionality reduction using Recursive Feature Elimination (FS)
```{r}
#Dimensionality reduction using Recursive Feature Elimination (FS)
FS_REIMS<-read.csv("REIMS_FS_data.csv")
```

        c) Logistic regression with PCA data
```{r}
# CODED BY PATRICK

#This section of code performs 10-fold cross validation on a logistic regression using different number of principal components (varying from 
# 1 principal component to 16 principal components). The optimal number of principal components to use for the logistic regression is identified
# and used to build a final logistic model on the full dataset. A graph of CV_Error vs. Number of Principal Components is provided, and a data
# frame that reports the optimal number of principal components to use, it's CV_Error, and the training error using the full dataset.

set.seed(1)

predictorsDF <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
gradeResponse <- as.factor(gradeDF$USDA.grade)

max_components <- 30 #maximum number of components to use in cross validation
n <- nrow(predictorsDF) 
k<-10 # number of folds
p <- .5 #probability threshold to bin logistic regression response

cv_labels <- sample(seq_len(k), n, replace = TRUE) #labels for folds
cv_mse <- rep(NA, max_components) # place holder vector to record cv_mse for each PC

#loop over each number of principle components
for (i in seq_len(max_components)) {
    num_components <- i
    
    loadingDF <- pca_loadings[,1:num_components] # store the correct loadings DF
    
    #compute the scores (z-values) for the given number of principal components
    scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
    scoresDF$USDA.grade <- gradeResponse
    
    cv_mse_i <- rep(NA, k) 
    
    #perform 10-fold cross validation
    for(j in seq_len(k)) {
        train <- scoresDF[cv_labels != j,]
        test <- scoresDF[cv_labels == j,]
        
        fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = train)
        pred <- predict(fit, newdata = test, type="response")
        confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=test$USDA.grade)
        
        #it was found that sometimes the predicted response only categorized one response, not two
        # so this error handling was necessary to get the appropriate error rate.
        if ( nrow(confMatrix) == 2) {
          errorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  
        } else {
          errorRate <- 1 - (confMatrix[1,1])/ sum(confMatrix)
        }
        
        cv_mse_i[j] <- errorRate
      }
    
    cv_mse[[i]] <- mean(cv_mse_i) #store the average CV-error of the 10 folds
}

#PLOT OF K-Fold CV_MSE VS. Principle Component
PrincipalComponent <- seq_len(max_components)
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=cv_mse))+
  geom_point()

#store the optimal cv error rate associated with the optimal number of principal components
optimal_CVerror <- min(cv_mse)
optimal_PC <- which.min(cv_mse)


#train a model with the optimal number of principal components and report
loadingDF <- pca_loadings[,1:optimal_PC]
scoresDF <- as.data.frame(as.matrix(predictorsDF) %*% as.matrix(loadingDF))
scoresDF$USDA.grade <- gradeResponse
fit <- glm(formula = USDA.grade ~ ., family = "binomial", data = scoresDF)
pred <- predict(fit,type="response")
confMatrix <-table(pred=ifelse(pred > p, "SE", "CH"), actual=scoresDF$USDA.grade)
TrainErrorRate <- 1 - (confMatrix[1,1] + confMatrix[2,2])/ sum(confMatrix)  

bestLogModelDF <- data.frame(optimalNumPrincpleCompoents = c(optimal_PC), optimalCVerror = c(optimal_CVerror))

PCvsCV_DF <- data.frame(PrincipalComponents = PrincipalComponent, CV_Error = cv_mse)

bestLogModelDF
PCvsCV_DF

write.csv(bestLogModelDF,"LogisticRegression/bestLogModelDF.csv", row.names = FALSE)
write.csv(PCvsCV_DF,"LogisticRegression/PCvsCV.csv", row.names = FALSE)

```

        e) K-nearest neighbors using PCA data
```{r}

```

        g) PCA using PCA data
```{r}

```

        i) Support vector machine (SVM) using PCA data
        
            
            iA) Linear SVM
```{r}

# This calls a script to run the Linear SVM
source("SVM/SVM_linear.R", local = knitr::knit_global())


```


```{r}
#Linear SVM Plotting and Results

svmLinResultsDF <- read.csv("SVM/Results/LinSVM.csv")
BestLinModel <- readRDS("SVM/BestSVMModels/BestLinModel.rds")
BestLinModel

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents_linSVM, y=CV_MSE_linSVM)) +
  geom_line()+
  geom_point()

ggplot(data= svmLinResultsDF, aes(x=PrincipalComponents_linSVM, y=Best_Cost)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost_lin, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost_lin, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC = c(svmLinResultsDF$PrincipalComponents_linSVM[which.min(svmLinResultsDF$CV_MSE_linSVM)])
optimalCVError =  c(min(svmLinResultsDF$CV_MSE_linSVM))
optimalCost = c(svmLinResultsDF$Best_Cost[which.min(svmLinResultsDF$CV_MSE_linSVM)])

data.frame(optimalPC = optimalPC, optimalCVError =optimalCVError,optimalCost= optimalCost)

```



            iB) Polynomial SVM with PC

```{r}


# This calls a script to run the Poly SVM
source("SVM/SVM_Poly.R", local = knitr::knit_global())



```

```{r}
#Plotting and Results for Poly SVM

abs_min_cost_poly <- .0001
abs_max_cost_poly <- 10000
BestPolyModel <- readRDS("SVM/BestSVMModels/BestPolyModel.rds")
BestPolyModel

svmPolyResultsDF <- read.csv("SVM/Results/PolySVM.csv")

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE_polySVM)) +
  geom_line()+
  geom_point()

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Cost_polySVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost_poly, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost_poly, linetype="dashed", color = "red") +
  scale_y_log10()

ggplot(data= svmPolyResultsDF, aes(x=PrincipalComponents, y=Best_Degree)) +
  geom_line()+
  geom_point()

optimalPC <- c(svmPolyResultsDF$PrincipalComponents[which.min(svmPolyResultsDF$CV_MSE_polySVM)])
optimalCVError <- c(min(svmPolyResultsDF$CV_MSE_polySVM))
optimalCost <- c(svmPolyResultsDF$Best_Cost_polySVM[which.min(svmPolyResultsDF$CV_MSE_polySVM)])
optimalDegree <- c(svmPolyResultsDF$Best_Degree[which.min(svmPolyResultsDF$CV_MSE_polySVM)])

#Results from tuning across all PCs
data.frame(optimalPC = optimalPC , optimalCVError = optimalCVError, optimalCost = optimalCost)

```


            iB) Radial SVM with PC

```{r}


# This calls a script to run the Radial SVM
source("SVM/SVM_rad.R", local = knitr::knit_global())


```


```{r}
abs_min_cost <- .0001
abs_max_cost <- 10000

abs_min_gamma <- .0001
abs_max_gamma <-10000

svmRadResultsDF <- read.csv("SVM/Results/radSVM.csv")
BestRadModel <- readRDS("SVM/BestSVMModels/BestRadModel.rds")
BestRadModel

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=CV_MSE_radSVM)) +
  geom_line()+
  geom_point()

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Cost_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10()

ggplot(data= svmRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Gamma_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_gamma, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_gamma, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- c(svmRadResultsDF$PrincipalComponents_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])
optimalCVError <- c(min(svmRadResultsDF$CV_MSE_radSVM))
optimalCost <- c(svmRadResultsDF$Best_Cost_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])
optimalGamma <- c(svmRadResultsDF$Best_Gamma_radSVM[which.min(svmRadResultsDF$CV_MSE_radSVM)])

#Results from tuning across all PCs
data.frame(optimalPC = optimalPC , optimalCVError = optimalCVError, optimalCost = optimalCost, optimalGamma = optimalGamma)

```

### Prediction of Tenderness over time using 36h REIMS (regression models)

     Subsets
```{r, echo = FALSE}
#Here are the names of the data sets for all the following models. They correspond to a 60/40 training/test subsets.
ten3DF_train
ten3DF_test
ten14DF_train
ten14DF_test
ten28DF_train
ten28DF_test
```

        a) Best subset selection
```{r}
#Tenderness at day 3
set.seed(1)
reg.fwd <- regsubsets(Ten3~., data = ten3DF_train, nvmax = 40,method = "forward") 
#Best subset selection using foward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten3~., data=ten3DF_test)
val.errors <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors[i] <- mean((ten3DF_test$Ten3-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 3",type = "o")

MSE_subset_day3<-min(val.errors)

#Tenderness at day 14
reg.fwd <- regsubsets(Ten14~., data = ten14DF_train, nvmax = 40,method = "forward") 
#Best subset selection using foward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten14~., data=ten14DF_test)
val.errors1 <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors1[i] <- mean((ten14DF_test$Ten14-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors1, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 14",type = "o")

MSE_subset_day14<-min(val.errors1)

#Tenderness at day 28
reg.fwd <- regsubsets(Ten28~., data = ten28DF_train, nvmax = 40,method = "forward") 
#Best subset selection using forward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten28~., data=ten28DF_test)
val.errors2 <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors2[i] <- mean((ten28DF_test$Ten28-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 28",type = "o")

MSE_subset_day28<-min(val.errors2)
```


        b) PCR
```{r}
#Fitting the PCR Ten3 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten3<- pcr (Ten3~., data=ten3DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten3)
mse3 <- MSEP(pcr.fit_ten3)
#M of the smallest CV error
mse3$comps[which.min(as.data.frame(mse3$val)[1,])]
#Calculating MSE with M=6 and the Test set.
pcr.pred <- predict (pcr.fit_ten3, ten3DF_test, ncomp = 6)
pcr.pred-ten3DF_test$Ten3
MSE_PCR_day3<-mean ((pcr.pred - ten3DF_test$Ten3)^2)
#Error compare to average
sqrt(MSE_PCR_day3)
mean(ten3DF$Ten3)


#Fitting the PCR Ten14 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten14<- pcr (Ten14~., data=ten14DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten14)
mse14 <- MSEP(pcr.fit_ten14)
#M of the smallest CV error
mse14$comps[which.min(as.data.frame(mse14$val)[1,])]
#Calculating MSE with M=1 and the Test set.
pcr.pred <- predict (pcr.fit_ten14, ten14DF_test, ncomp = 1)
pcr.pred-ten14DF_test$Ten14
MSE_PCR_day14<-mean ((pcr.pred - ten14DF_test$Ten14)^2)
#Error in the same scale
sqrt(MSE_PCR_day14)
mean(ten14DF$Ten14)



#Fitting the PCR Ten28 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten28<- pcr (Ten28~., data=ten28DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten28)
mse28 <- MSEP(pcr.fit_ten28)
#M of the smallest CV error
mse28$comps[which.min(as.data.frame(mse28$val)[1,])]
size_PCR_day28<-which.min(as.data.frame(mse28$val)[1,])
#Calculating MSE with M=2 and the Test set.
pcr.pred <- predict (pcr.fit_ten28, ten28DF_test, ncomp = 2)
pcr.pred-ten28DF_test$Ten28
MSE_PCR_day28<-mean ((pcr.pred - ten28DF_test$Ten28)^2)
#Error in the same scale
sqrt(MSE_PCR_day28)
mean(ten28DF$Ten28)


```


        c)Lasso Regression
        
```{r}

#Lasso on Ten3 (Non-PCA)

# THIS MODEL IS NO BETTER THAN AVERAGE!!!!
set.seed(1)

X_train <- model.matrix(Ten3 ~ ., data = ten3DF_train)[,-1]
Y_train <- ten3DF_train$Ten3
X_test <- model.matrix(Ten3 ~ ., data = ten3DF_test)[,-1]
Y_test <- ten3DF_test$Ten3
lambda <- 10^seq(-2, 3, length.out = 1000)
cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

lambdaPlot3 <- ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()

lambdaMin <- cv_lasso$lambda.min
cv_error_min <- min(cv_lasso$cvm)

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin, type="response")

testErrorLasso <- mean(( predictions[,'s1'] - Y_test )^2)

predictionPlot3 <- ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) 

coefs_3 <- coef(lasso_mdl)
size_3 <- length(coefs_3[coefs_3[,1]!=0,]) - 1

Ten3NonPCA_Results <- data.frame(CV_MSE = cv_error_min, Lambda = lambdaMin, TestError = testErrorLasso, Size = size_3)
Ten3NonPCA_Results


#Lasso on Ten14 (Non-PCA)

# THIS MODEL IS NO BETTER THAN AVERAGE!!!!
set.seed(1)

X_train <- model.matrix(Ten14 ~ ., data = ten14DF_train)[,-1]
Y_train <- ten14DF_train$Ten14
X_test <- model.matrix(Ten14 ~ ., data = ten14DF_test)[,-1]
Y_test <- ten14DF_test$Ten14
lambda <- 10^seq(-2, 3, length.out = 1000)
cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

lambdaPlot14 <- ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()

lambdaMin <- cv_lasso$lambda.min
cv_error_min <- min(cv_lasso$cvm)

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin, type="response")

testErrorLasso <- mean(( predictions[,'s1'] - Y_test )^2)

predictionPlot14 <- ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) 

coefs_14 <- coef(lasso_mdl)
size_14 <- length(coefs_14[coefs_14[,1]!=0,]) - 1

Ten14NonPCA_Results <- data.frame(CV_MSE = cv_error_min, Lambda = lambdaMin, TestError = testErrorLasso, Size = size_14)
Ten14NonPCA_Results

#Lasso on Ten28 (Non-PCA)
set.seed(1)

X_train <- model.matrix(Ten28 ~ ., data = ten28DF_train)[,-1]
Y_train <- ten28DF_train$Ten28
X_test <- model.matrix(Ten28 ~ ., data = ten28DF_test)[,-1]
Y_test <- ten28DF_test$Ten28
lambda <- 10^seq(-2, 3, length.out = 1000)
cv_lasso <- cv.glmnet(X_train, Y_train, alpha = 1, lambda = lambda)

lambdaPlot28 <- ggplot() +
  geom_line(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  geom_point(aes(cv_lasso$lambda, cv_lasso$cvm)) +
  scale_x_log10()

lambdaMin <- cv_lasso$lambda.min
cv_error_min <- min(cv_lasso$cvm)

lasso_mdl <- glmnet(X_train, Y_train, alpha = 1, lambda = lambdaMin)

predictions <- predict(lasso_mdl,newx=X_test,s=lambdaMin, type="response")

testErrorLasso <- mean(( predictions[,'s1'] - Y_test )^2)

predictionPlot28 <- ggplot() +
  geom_point(aes(predictions, Y_test))+
  geom_abline(slope=1, intercept = 0) 

coefs_28 <- coef(lasso_mdl)
size_28 <- length(coefs_28[coefs_28[,1]!=0,]) - 1

Ten28NonPCA_Results <- data.frame(CV_MSE = cv_error_min, Lambda = lambdaMin, TestError = testErrorLasso, Size = size_28)
Ten28NonPCA_Results

```

```{r}
# Lasso on principal components for Ten3

# This code is commented out because it takes a while to run. Results are stored in files.
#source("Lasso/Lasso_ten3_CV.R", local = knitr::knit_global())


```

```{r}
lambda_min_under <- -2
lambda_max_over <- 8

LassoTen3Results <- read.csv("Lasso/Results/Ten3.csv")
BestTen3LassoModel <- readRDS("Lasso/BestModels/BestLassoTen3Model.rds")
BestTen3LassoModel

ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()


ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = Size)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen3Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- LassoTen3Results$PrincipalComponents[which.min(LassoTen3Results$CV_MSE)]
optimal_TestError <-  min(LassoTen3Results$CV_MSE)
optimalLambda <- LassoTen3Results$Best_Lambda[which.min(LassoTen3Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

MSE_Lasso_day3<-optimal_TestError
```

```{r}
# Lasso on principal components for Ten14

# This code is commented out because it takes a while to run. Results are stored in files.
# source("Lasso/Lasso_ten14_CV.R", local = knitr::knit_global())

```

```{r}

lambda_min_under <- -4
lambda_max_over <- 8

LassoTen14Results <- read.csv("Lasso/Results/Ten14.csv")
BestTen14LassoModel <- readRDS("Lasso/BestModels/BestLassoTen14Model.rds")
BestTen14LassoModel

ggplot(data=LassoTen14Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen14Results, aes(x= PrincipalComponents, y = Size)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen14Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- LassoTen14Results$PrincipalComponents[which.min(LassoTen14Results$CV_MSE)]
optimal_TestError <-  min(LassoTen14Results$CV_MSE)
optimalLambda <- LassoTen14Results$Best_Lambda[which.min(LassoTen14Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

MSE_Lasso_day14<-optimal_TestError


```
```{r}
# Lasso on principal components for Ten28

# This code is commented out because it takes a while to run. Results are stored in files.
# source("Lasso/Lasso_ten28_CV.R", local = knitr::knit_global())


```

```{r}

lambda_min_under <- -1 
lambda_max_over <- 10

LassoTen28Results <- read.csv("Lasso/Results/Ten28.csv")
BestTen28LassoModel <- readRDS("Lasso/BestModels/BestLassoTen28Model.rds")
BestTen28LassoModel

ggplot(data=LassoTen28Results, aes(x= PrincipalComponents, y = CV_MSE)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen28Results, aes(x= PrincipalComponents, y = Size)) +
  geom_line() +
  geom_point()

ggplot(data=LassoTen28Results, aes(x= PrincipalComponents, y = Best_Lambda)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=10^lambda_min_under, linetype="dashed", color = "red") +
  geom_hline(yintercept=10^lambda_max_over, linetype="dashed", color = "red") +
  scale_y_log10()

optimalPC <- LassoTen28Results$PrincipalComponents[which.min(LassoTen28Results$CV_MSE)]
optimal_TestError <-  min(LassoTen28Results$CV_MSE)
optimalLambda <- LassoTen28Results$Best_Lambda[which.min(LassoTen28Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

MSE_Lasso_day28<-optimal_TestError
```

        d) Splines
        
```{r}
# Spline model for Tenderness at day 3
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
ten3<-na.omit( dat1[,-c(1,3:4)])
ten3_train<-ten3[dpart,]
ten3_test<-ten3[-dpart,]
err3<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten3~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten3_train)
preds<- predict(gam, newdata = ten3_test)
preds
err3[i]<-mean((ten3_test$Ten3 - preds)^2)
}
Spline3<-data.frame(N.comp=seq(1,10),MSE=err3)
which.min(Spline3$MSE)
min(Spline3$MSE)


# Spline model for Tenderness at day 14
set.seed(1)
ten14<-na.omit( dat1[,-c(1:2,4)])
ten14_train<-ten14[dpart,]
ten14_test<-ten14[-dpart,]
err14<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten14~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten14_train)
preds<- predict(gam, newdata = ten14_test)
preds
err14[i]<-mean((ten14_test$Ten14 - preds)^2)
}
Spline14<-data.frame(N.comp=seq(1,10),MSE=err14)
which.min(Spline14$MSE)
min(Spline14$MSE)

# Spline model for Tenderness at day 28
set.seed(1)
ten28<-na.omit( dat1[,-c(1:3)])
ten28_train<-ten28[dpart,]
ten28_test<-ten28[-dpart,]
err28<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten28~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten28_train)
preds<- predict(gam, newdata = ten28_test)
preds
err28[i]<-mean((ten28_test$Ten28 - preds)^2)
}
Spline28<-data.frame(N.comp=seq(1,10),MSE=err28)
which.min(Spline28$MSE)
min(Spline28$MSE)

```

``` {r}
#Summary results of regression

data.frame(  Model = "Best Subset", 
             Size_day3 = "1", MSE_day3=MSE_subset_day3,
             Size_day14 = "1", MSE_day14=MSE_subset_day14,
             Size_day28 = "1", MSE_day28=MSE_subset_day28)%>%
  bind_rows(data.frame(Model = "PCA-Lasso", 
            Size_day3 ="18", MSE_day3=MSE_Lasso_day3, 
            Size_day14 ="10", MSE_day14=MSE_Lasso_day14, 
            Size_day28 ="58", MSE_day28=MSE_Lasso_day28))%>%
  bind_rows(data.frame(Model = "PCR", 
            Size_day3 ="6", MSE_day3= MSE_PCR_day3, 
            Size_day14 ="1", MSE_day14= MSE_PCR_day14, 
            Size_day28 ="2", MSE_day28= MSE_PCR_day28)) %>%
  bind_rows(data.frame(Model = "PCA-Splines", 
            Size_day3 ="5", MSE_day3= min(Spline3$MSE), 
            Size_day14 ="2", MSE_day14= min(Spline14$MSE), 
            Size_day28 ="4", MSE_day28= min(Spline28$MSE))) %>%
kable(caption="Table 2. Mean Squared Error (MSE) of regression models for tenderness at day 3, 14, and 28",digits=2)



```

