---
title: "Patrick-Presentation"
author: "Patrick Noe"
date: "12/4/2021"
output:
  beamer_presentation:
    theme: "Frankfurt"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(knitr)
```

# USDA Grade Categorical Prediction

```{r}
# Import Data
LogModelDF <- read.csv("LogisticRegression/PCvsCV.csv")
SVMLinResultsDF <- read.csv("SVM/Results/LinSVM.csv")
SVMPolyResultsDF <- read.csv("SVM/Results/PolySVM.csv")
SVMRadResultsDF <- read.csv("SVM/Results/radSVM.csv")

#Process Data
optimalIndex <- which.min(SVMLinResultsDF$CV_MSE_linSVM)
optimalPC_SVMLin <- SVMLinResultsDF$PrincipalComponents_linSVM[optimalIndex]
optimalCVError_SVMLin <- min(SVMLinResultsDF$CV_MSE_linSVM)
optimalCost_SVMLin <- SVMLinResultsDF$Best_Cost[optimalIndex]

optimalIndex <- which.min(SVMPolyResultsDF$CV_MSE_polySVM)
optimalPC_SVMPoly <- SVMPolyResultsDF$PrincipalComponents[optimalIndex]
optimalCVError_SVMPoly <- min(SVMPolyResultsDF$CV_MSE_polySVM)
optimalCost_SVMPoly <- SVMPolyResultsDF$Best_Cost_polySVM[optimalIndex]
optimalDegree_SVMPoly <- SVMPolyResultsDF$Best_Degree[optimalIndex] 

optimalIndex <- which.min(SVMRadResultsDF$CV_MSE_radSVM)
optimalPC_SVMRad <- SVMRadResultsDF$PrincipalComponents_radSVM[optimalIndex]
optimalCVError_SVMRad <- min(SVMRadResultsDF$CV_MSE_radSVM)
optimalCost_SVMRad <- SVMRadResultsDF$Best_Cost_radSVM[optimalIndex]
optimalGamma_SVMRad <- SVMRadResultsDF$Best_Gamma_radSVM[optimalIndex] 

optimalIndex <- which.min(LogModelDF$CV_Error)
optimalPC_Log <- LogModelDF$PrincipalComponents[optimalIndex]
optimalCVError_Log <- min(LogModelDF$CV_Error)

```

## Methodology

## Overview 

- Several models were built utilizing the Principal Components to predict the USDA Grade
  - Support Vector Machines (Linear, Polynomial, and Radial)
  - Logistic Regression
- Results:

```{r}
ModelNames <- c('Radial SVM', 'Linear SVM', 'Polynomial SVM', 'Logistic Regression')
PrincipalComponents <- c(optimalPC_SVMRad, optimalPC_SVMLin, optimalPC_SVMPoly, optimalPC_Log)
CV_Error <- c(optimalCVError_SVMRad, optimalCVError_SVMLin, optimalCVError_SVMPoly, optimalCVError_Log)

SummaryDF <- data.frame(Model = ModelNames,PrincipalComponents = PrincipalComponents, CV_Error = CV_Error)

kable(SummaryDF)
```
- Best Model: SVM Radial


## Radial SVM Summary

```{r}

BestRadModel <- readRDS("SVM/BestSVMModels/BestRadModel.rds")
BestRadModel<- BestRadModel[[1]]
numSupportVectors <- nrow(BestRadModel$SV)

kable(data.frame(PC = optimalPC_SVMRad , CV_Error = optimalCVError_SVMRad, Cost = optimalCost_SVMRad, Gamma = optimalGamma_SVMRad, SV = numSupportVectors))



```

## Radial SVM CV Error vs. Principal Components

```{r}

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=CV_MSE_radSVM)) +
  geom_line()+
  geom_point()





```

## Radial SVM: Optimal Cost vs. Principal Components

```{r}
abs_min_cost <- .0001
abs_max_cost <- 10000

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Cost_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10()

```

## Radial SVM: Optimal Gamma vs. Principal Components
```{r}
abs_min_gamma <- .0001
abs_max_gamma <-10000

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Gamma_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_gamma, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_gamma, linetype="dashed", color = "red") +
  scale_y_log10()

```


## Linear Support Vector Machine

```{r}

ggplot(data= SVMLinResultsDF, aes(x=PrincipalComponents_linSVM, y=CV_MSE_linSVM)) +
  geom_line()+
  geom_point()


```

## Polynomial Support Vector Machine

```{r}

ggplot(data= SVMPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE_polySVM)) +
  geom_line()+
  geom_point()
```

## Logistic Regression



```{r pressure, echo=FALSE}

bestLogModelDF <- read.csv("LogisticRegression/bestLogModelDF.csv")
PCvsCVLog <- read.csv("LogisticRegression/PCvsCV.csv")

#bestLogModelDF

ggplot(data = PCvsCVLog, aes(x = PrincipalComponents, y= CV_Error)) +
  geom_line()+
  geom_point()

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
