---
title: "Patrick-Presentation"
author: "Patrick Noe"
date: "12/4/2021"
output:
  beamer_presentation:
    theme: "Frankfurt"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(knitr)
```

# USDA Grade Categorical Prediction

```{r}
# Import Data
LogModelDF <- read.csv("LogisticRegression/PCvsCV.csv")
SVMLinResultsDF <- read.csv("SVM/Results/LinSVM.csv")
SVMPolyResultsDF <- read.csv("SVM/Results/PolySVM.csv")
SVMRadResultsDF <- read.csv("SVM/Results/radSVM.csv")

#Process Data
optimalIndex <- which.min(SVMLinResultsDF$CV_MSE_linSVM)
optimalPC_SVMLin <- SVMLinResultsDF$PrincipalComponents_linSVM[optimalIndex]
optimalCVError_SVMLin <- round(min(SVMLinResultsDF$CV_MSE_linSVM),3)
optimalCost_SVMLin <- SVMLinResultsDF$Best_Cost[optimalIndex]

optimalIndex <- which.min(SVMPolyResultsDF$CV_MSE_polySVM)
optimalPC_SVMPoly <- SVMPolyResultsDF$PrincipalComponents[optimalIndex]
optimalCVError_SVMPoly <- round(min(SVMPolyResultsDF$CV_MSE_polySVM),3)
optimalCost_SVMPoly <- SVMPolyResultsDF$Best_Cost_polySVM[optimalIndex]
optimalDegree_SVMPoly <- SVMPolyResultsDF$Best_Degree[optimalIndex] 

optimalIndex <- which.min(SVMRadResultsDF$CV_MSE_radSVM)
optimalPC_SVMRad <- SVMRadResultsDF$PrincipalComponents_radSVM[optimalIndex]
optimalCVError_SVMRad <- round(min(SVMRadResultsDF$CV_MSE_radSVM),3)
optimalCost_SVMRad <- SVMRadResultsDF$Best_Cost_radSVM[optimalIndex]
optimalGamma_SVMRad <- SVMRadResultsDF$Best_Gamma_radSVM[optimalIndex] 

optimalIndex <- which.min(LogModelDF$CV_Error)
optimalPC_Log <- LogModelDF$PrincipalComponents[optimalIndex]
optimalCVError_Log <- round(min(LogModelDF$CV_Error),3)

```

## Methodology

- Several models were built to predict USDA Grade using the principal components:
  - Logistic Regression
  - Support Vector Machines (Linear, Polynomial, Radial)
- 10-Fold cross validation was performed on each model using an increasing number of principal components
- Optimal CV Error was recorded for each model and this was used to identify the optimal number of principal components to utilize for each model.


## Logistic Regression


```{r pressure, echo=FALSE}

bestLogModelDF <- read.csv("LogisticRegression/bestLogModelDF.csv")
PCvsCVLog <- read.csv("LogisticRegression/PCvsCV.csv")

#bestLogModelDF

ggplot(data = PCvsCVLog, aes(x = PrincipalComponents, y= CV_Error)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=16, face="bold"),
        axis.title=element_text(size=16, face="bold"))

```

## Radial SVM Summary

- Optimal Model:

```{r}

BestRadModel <- readRDS("SVM/BestSVMModels/BestRadModel.rds")
BestRadModel<- BestRadModel[[1]]
numSupportVectors <- nrow(BestRadModel$SV)

kable(data.frame(PC = optimalPC_SVMRad , CV_Error = optimalCVError_SVMRad, Cost = optimalCost_SVMRad, Gamma = optimalGamma_SVMRad, SV = numSupportVectors))



```
- Built models using an increasing number of principal components from 2 to 60.
- For each iteration of principal components, a SVM was tuned using:
  - Costs from .0001 to 10,000
  - $\gamma$ from .0001 to 10,000

## Radial SVM CV Error vs. Principal Components

```{r}

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=CV_MSE_radSVM)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=16, face="bold"),
        axis.title=element_text(size=16, face="bold"))




```

## Radial SVM: Optimal Cost vs. Principal Components

```{r}
abs_min_cost <- .0001
abs_max_cost <- 10000

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Cost_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10() +
  labs(y="Optimal Cost", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=16, face="bold"),
        axis.title=element_text(size=16, face="bold"))

```

## Radial SVM: Optimal $\gamma$ vs. Principal Components
```{r}
abs_min_gamma <- .0001
abs_max_gamma <-10000

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Gamma_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_gamma, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_gamma, linetype="dashed", color = "red") +
  scale_y_log10() +
  labs(y="Optimal Gamma", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=16, face="bold"),
        axis.title=element_text(size=16, face="bold"))

```


## Linear Support Vector Machine

```{r}

ggplot(data= SVMLinResultsDF, aes(x=PrincipalComponents_linSVM, y=CV_MSE_linSVM)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=16, face="bold"),
        axis.title=element_text(size=16, face="bold"))


```

## Polynomial Support Vector Machine

```{r}

ggplot(data= SVMPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE_polySVM)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=16, face="bold"),
        axis.title=element_text(size=16, face="bold"))
```




## Summary of Results 


```{r}
ModelNames <- c('Radial SVM', 'Linear SVM', 'Polynomial SVM', 'Logistic Regression')
PrincipalComponents <- c(optimalPC_SVMRad, optimalPC_SVMLin, optimalPC_SVMPoly, optimalPC_Log)
CV_Error <- c(optimalCVError_SVMRad, optimalCVError_SVMLin, optimalCVError_SVMPoly, optimalCVError_Log)

SummaryDF <- data.frame(Model = ModelNames,PrincipalComponents = PrincipalComponents, CV_Error = CV_Error)

kable(SummaryDF)
```
- Best Model: Radial SVM with 30 Principal Components: CV Error of .108
  - Even though the Polynomial SVM outperformed the Radial SVM slightly, it required 48 more Principal Components to be utilized.
  - The Linear SVM performed slightly worse than the Radial SVM, but required 6 less Principal Components.