---
title: "Presentation"
author: "Michael Hernandez"
date: "12/4/2021"
output:
  beamer_presentation:
    theme: Frankfurt
    colortheme: default
    fonttheme: structurebold
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(include = FALSE)
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
library(FactoMineR)
```


# INTRODUCTION

## Beef industry context
- Beef palatability depends on tenderness, juiciness, and flavor. 
- In the United States, there is a quality grading system that classifies beef carcasses on different quality categories (USDA) based on animal maturity and marbling level (intramuscular fat).
- Tenderness is not predicted with USDA quality grade.
- Although some technologies allow the determination of beef tenderness, they require sacrificing a loin steak and cannot be performed in real-time.

## Rapid Evaporative Ionization Mass Spectrometry (REIMS)

- Rapid Evaporative Ionization Mass Spectrometry (REIMS) is a novel environmental mass spectrometry technique that combines a sample vaporization device with a time-of-flight mass spectrometer. 
- The sampling device, which uses a high-frequency current, comes into contact with the sample surface, an aerosol of molecules is generated and transformed into a mass spectral fingerprint. 
- REIMS can classify biological tissues in-situ, in real-time and without any sample preparation. 
- The goal of this study was to evaluate the ability of REIMS to predict USDA quality grade and beef tenderness.


# METHODOLOGY

## Data collection

- Eighty-three whole striploins were collected at 36h post-mortem (42 USDA quality grade Choice, and 41 USDA quality grade Select).
- They were cut into three sections, which were aged at 3 different aging treatments (3 days, 14 days, 28 days).
- Tenderness were measure at each aging period.
- The number of observations is 83, and the number of predictors is 1700.
- Classification models were performed to predict USDA quality grade.
- Regression models were performed to predict tenderness.

# RESULTS

## Data structure (Principal Component Analysis)

```{r}
### Reproducibility
set.seed(1)
```

```{r}
### Uploading data
REIMS <- read.csv("REIMS machine learning class.csv")

dropBase <- c("SampleID", "Cx", "GS_1", "GS_2", "UNIQ_ID", "Ten42", "Ten56", "Ten70")
responseVars <- c("USDA.grade", "Ten3", "Ten14","Ten28")

# Define the associated dataframes for each analysis. Response variable is column 1, predictors are the remaining columns
gradeDF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("USDA.grade"))))]
```

```{r, include=TRUE,fig.width = 6, fig.height = 2}
#Performing PCA
pca1 <- PCA(REIMS[,-c(1:12)], ncp = 100, scale.unit=TRUE ,graph = F)
dat1<-REIMS[,6]
#Building data frame
dat1 <- data.frame(USDA.grade=dat1,Ten3=REIMS[,7],Ten14=REIMS[,8],Ten28=REIMS[,9],pca1$ind$coord[, c(1:5)])
#Plotting PCA faced by USDA grade
p <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=USDA.grade)) +
  geom_point()
print(p + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Comp 2", x = "Comp 1"))

#Plotting PCA faced by USDA grade
p1 <- ggplot(data = dat1, aes(x = Dim.2, y = Dim.3,colour=USDA.grade)) +
  geom_point()
print(p1 + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Comp 2", x = "Comp 1"))
```

## Data structure (Tenderness)  
```{r, include=TRUE, fig.width = 8, fig.height = 4}
#Plotting PCA colored by tenderness at day 3
p2 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten3)) +
  geom_point()
print(p2 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))

#Plotting PCA colored by tenderness at day 14
p3 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten14)) +
  geom_point()
print(p3 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))

#Plotting PCA colored by tenderness at day 28
p4 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten28)) +
  geom_point()
print(p4 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))
```
