---
title: "Presentation"
author: "Patrick Noe, Logan Perkins, Michael Hernandez"
date: "12/4/2021"
output:
  beamer_presentation:
    theme: Frankfurt
    colortheme: default
    fonttheme: structurebold
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(include = FALSE)
library(ggplot2)
library(MASS)
library(GGally)
library(caret)
library(tidyverse)
library(ggrepel)
library(ISLR)
library(hrbrthemes)
library(class)
library(boot)
library(leaps)
library(glmnet)
library(pls)
library(gam)
library(tree)
library(gbm)
library(randomForest)
library(e1071)
library(FactoMineR)
library(kableExtra)
```


# INTRODUCTION

## Beef industry context
- Beef palatability depends on tenderness, juiciness, and flavor. 
- In the United States, there is a quality grading system that classifies beef carcasses on different quality categories (USDA) based on animal maturity and marbling level (intramuscular fat).
- Tenderness is not predicted with USDA quality grade.
- Although some technologies allow the determination of beef tenderness, they require sacrificing a loin steak and cannot be performed in real-time.

## Rapid Evaporative Ionization Mass Spectrometry (REIMS)

- Rapid Evaporative Ionization Mass Spectrometry (REIMS) is a novel environmental mass spectrometry technique that combines a sample vaporization device with a time-of-flight mass spectrometer. 
- The sampling device, which uses a high-frequency current, comes into contact with the sample surface, an aerosol of molecules is generated and transformed into a mass spectral fingerprint. 
- REIMS can classify biological tissues in-situ, in real-time and without any sample preparation. 
- The goal of this study was to evaluate the ability of REIMS to predict USDA quality grade and beef tenderness.


# METHODOLOGY

## Data collection

- Eighty-three whole striploins were collected at 36h post-mortem (42 USDA quality grade Choice, and 41 USDA quality grade Select).
- They were cut into three sections, which were aged at 3 different aging treatments (3 days, 14 days, 28 days).
- Tenderness were measure at each aging period.
- The number of observations is 83, and the number of predictors is 1700.
- Classification models were performed to predict USDA quality grade.
- Regression models were performed to predict tenderness.

# PCA

## USDA quality grade
```{r}
### Reproducibility
set.seed(1)
```

```{r}
### Uploading data
REIMS <- read.csv("REIMS machine learning class.csv")

dropBase <- c("SampleID", "Cx", "GS_1", "GS_2", "UNIQ_ID", "Ten42", "Ten56", "Ten70")
responseVars <- c("USDA.grade", "Ten3", "Ten14","Ten28")

# Define the associated dataframes for each analysis. Response variable is column 1, predictors are the remaining columns
gradeDF <- REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("USDA.grade"))))]
```

```{r, include=TRUE,fig.width = 6, fig.height = 2.2}
#Performing PCA
pca1 <- PCA(REIMS[,-c(1:12)], ncp = 100, scale.unit=TRUE ,graph = F)
dat1<-REIMS[,6]
#Building data frame
dat1 <- data.frame(USDA.grade=dat1,Ten3=REIMS[,7],Ten14=REIMS[,8],Ten28=REIMS[,9],pca1$ind$coord[, c(1:40)])
#Plotting PCA faced by USDA grade
p <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=USDA.grade)) +
  geom_point()
print(p + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Comp 2", x = "Comp 1"))

#Plotting PCA faced by USDA grade
p1 <- ggplot(data = dat1, aes(x = Dim.2, y = Dim.3,colour=USDA.grade)) +
  geom_point()
print(p1 + labs(title= "PCA of REIMS data faced by USDA quality grade",y="Comp 2", x = "Comp 1"))
```

## Tenderness 
```{r, include=TRUE, fig.width = 6, fig.height = 2.2}
#Plotting PCA colored by tenderness at day 3
p2 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten3)) +
  geom_point()
print(p2 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))

#Plotting PCA colored by tenderness at day 28
p4 <- ggplot(data = dat1, aes(x = Dim.1, y = Dim.2,colour=Ten28)) +
  geom_point()
print(p4 + labs(title= "PCA of REIMS data colored by tenderness at day 3",y="Comp 2", x = "Comp 1"))
```

## Proportion of Variance Explained

```{r, include=TRUE,fig.width = 6, fig.height = 4}
#Performing PCA

gradeResponse <- as.factor(gradeDF$USDA.grade)
data <- gradeDF[,!(names(gradeDF) %in% c("USDA.grade"))] # dataframe without response
pca_obj <- prcomp(data, center=TRUE, scale=TRUE)  # perform PCA
pca_summary <- summary(pca_obj)
pca_loadings <- pca_obj$rotation #store the loading vectors for future use

#pca_summary

PVE <- pca_summary$importance[2,] # store the proportion of variance explained by each PC
PrincipalComponent <- seq(1, length(PVE), 1) # Vector of Principle components for plot

#PLOT OF PVE VS. Principle Component
ggplot() +
  geom_line(aes(x=PrincipalComponent, y=PVE))+
  geom_point()+
  labs(title= "Proportion of Variance Explained (PVE) vs. Principal Component", y="PVE", x = "Prinicipal Component") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```


# CATEGORIAL MODELS: USDA GRADE

```{r}
# Import Data
LogModelDF <- read.csv("LogisticRegression/PCvsCV.csv")
SVMLinResultsDF <- read.csv("SVM/Results/LinSVM.csv")
SVMPolyResultsDF <- read.csv("SVM/Results/PolySVM.csv")
SVMRadResultsDF <- read.csv("SVM/Results/radSVM.csv")

#Process Data
optimalIndex <- which.min(SVMLinResultsDF$CV_MSE_linSVM)
optimalPC_SVMLin <- SVMLinResultsDF$PrincipalComponents_linSVM[optimalIndex]
optimalCVError_SVMLin <- round(min(SVMLinResultsDF$CV_MSE_linSVM),3)
optimalCost_SVMLin <- SVMLinResultsDF$Best_Cost[optimalIndex]

optimalIndex <- which.min(SVMPolyResultsDF$CV_MSE_polySVM)
optimalPC_SVMPoly <- SVMPolyResultsDF$PrincipalComponents[optimalIndex]
optimalCVError_SVMPoly <- round(min(SVMPolyResultsDF$CV_MSE_polySVM),3)
optimalCost_SVMPoly <- SVMPolyResultsDF$Best_Cost_polySVM[optimalIndex]
optimalDegree_SVMPoly <- SVMPolyResultsDF$Best_Degree[optimalIndex] 

optimalIndex <- which.min(SVMRadResultsDF$CV_MSE_radSVM)
optimalPC_SVMRad <- SVMRadResultsDF$PrincipalComponents_radSVM[optimalIndex]
optimalCVError_SVMRad <- round(min(SVMRadResultsDF$CV_MSE_radSVM),3)
optimalCost_SVMRad <- SVMRadResultsDF$Best_Cost_radSVM[optimalIndex]
optimalGamma_SVMRad <- SVMRadResultsDF$Best_Gamma_radSVM[optimalIndex] 

optimalIndex <- which.min(LogModelDF$CV_Error)
optimalPC_Log <- LogModelDF$PrincipalComponents[optimalIndex]
optimalCVError_Log <- round(min(LogModelDF$CV_Error),3)

```

## Methodology

- Several models were built to predict USDA Grade using the principal components:
  - Logistic Regression
  - Support Vector Machines (Linear, Polynomial, Radial)
- 10-Fold cross validation was performed on each model using an increasing number of principal components
- Optimal CV Error was recorded for each model and this was used to identify the optimal number of principal components to utilize for each model.

## Radial Support Vector Machine

- Support Vector Machine with a Radial kernel.
- Built models using an increasing number of principal components from 2 to 60.
- For each iteration of principal components, a SVM was tuned using:
  - Costs from .0001 to 10,000
  - $\gamma$ from .0001 to 10,000
  

```{r, include=TRUE}

BestRadModel <- readRDS("SVM/BestSVMModels/BestRadModel.rds")
BestRadModel<- BestRadModel[[1]]
numSupportVectors <- nrow(BestRadModel$SV)
SummaryDF <- data.frame(PC = optimalPC_SVMRad , CV_Error = optimalCVError_SVMRad, Cost = optimalCost_SVMRad, Gamma = optimalGamma_SVMRad, SV = numSupportVectors) 

SummaryDF%>%
kable(caption="Best performing Radial SVM Model.",digits=4)



```

## Radial Support Vector Machine


```{r, include=TRUE,fig.width = 6, fig.height = 4}

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=CV_MSE_radSVM)) +
  geom_line()+
  geom_point()+
  labs(title = "CV Error vs. Prinicipal Components",y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        plot.title = element_text(hjust = 0.5))




```

## Radial SVM: Optimal Parameters

```{r, include=TRUE,fig.width = 6, fig.height = 2}
abs_min_cost <- .0001
abs_max_cost <- 10000

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Cost_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10() +
  labs(title= "Optimal Cost vs. Principal Components", y="Optimal Cost", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

```{r, include=TRUE,fig.width = 6, fig.height = 2}
abs_min_gamma <- .0001
abs_max_gamma <-10000

ggplot(data= SVMRadResultsDF, aes(x=PrincipalComponents_radSVM, y=Best_Gamma_radSVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_gamma, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_gamma, linetype="dashed", color = "red") +
  scale_y_log10() +
  labs(title = "Optimal Gamma vs. Principal Components",y="Optimal Gamma", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

## Linear Support Vector Machine

- Support Vector Machine with a Linear kernel.
- Built models using an increasing number of principal components from 2 to 80.
- For each iteration of principal components, a SVM was tuned using:
  - Costs from .01 to 100
  
```{r, include=TRUE}

BestLinModel <- readRDS("SVM/BestSVMModels/BestLinModel.rds")
BestLinModel<- BestLinModel[[1]]
numSupportVectors <- nrow(BestLinModel$SV)
SummaryDF <- data.frame(PC = optimalPC_SVMLin , CV_Error = optimalCVError_SVMLin, Cost = optimalCost_SVMLin, SV = numSupportVectors) 

SummaryDF%>%
kable(caption="Best performing Linear SVM Model.",digits=4)



```
## Linear Support Vector Machine

```{r, include=TRUE,fig.width = 6, fig.height = 4}

ggplot(data= SVMLinResultsDF, aes(x=PrincipalComponents_linSVM, y=CV_MSE_linSVM)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        plot.title = element_text(hjust = 0.5))


```

## Linear SVM: Optimal Parameters

```{r, include=TRUE,fig.width = 6, fig.height = 2}
abs_min_cost <- .01
abs_max_cost <- 100

ggplot(data= SVMLinResultsDF, aes(x=PrincipalComponents_linSVM, y=Best_Cost)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10() +
  labs(title= "Optimal Cost vs. Principal Components", y="Optimal Cost", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```



## Polynomial Support Vector Machine
- Support Vector Machine with a Polynomal kernel.
- Built models using an increasing number of principal components from 2 to 60.
- For each iteration of principal components, a SVM was tuned using:
  - Costs from .01 to 100
  - Degrees from 1 to 4
  
```{r, include=TRUE}

BestPolyModel <- readRDS("SVM/BestSVMModels/BestPolyModel.rds")
BestPolyModel<- BestPolyModel[[1]]
numSupportVectors <- nrow(BestPolyModel$SV)
SummaryDF <- data.frame(PC = optimalPC_SVMPoly , CV_Error = optimalCVError_SVMPoly, Cost = optimalCost_SVMPoly, Degree=optimalDegree_SVMPoly, SV = numSupportVectors) 

SummaryDF%>%
kable(caption="Best performing Linear SVM Model.",digits=4)



```

## Polynomial Support Vector Machine

```{r, include=TRUE,fig.width = 6, fig.height = 4}

ggplot(data= SVMPolyResultsDF, aes(x=PrincipalComponents, y=CV_MSE_polySVM)) +
  geom_line()+
  geom_point()+
  labs(title="CV Error vs. Principal Components", y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))
```

## Polynomial SVM: Optimal Parameters

```{r, include=TRUE,fig.width = 6, fig.height = 2}
abs_min_cost <- .0001
abs_max_cost <- 10000

ggplot(data= SVMPolyResultsDF, aes(x=PrincipalComponents, y=Best_Cost_polySVM)) +
  geom_line()+
  geom_point()+
  geom_hline(yintercept=abs_min_cost, linetype="dashed", color = "red") +
  geom_hline(yintercept=abs_max_cost, linetype="dashed", color = "red") +
  scale_y_log10() +
  labs(title= "Optimal Cost vs. Principal Components", y="Optimal Cost", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

```{r, include=TRUE,fig.width = 6, fig.height = 2}

ggplot(data= SVMPolyResultsDF, aes(x=PrincipalComponents, y=Best_Degree)) +
  geom_line()+
  geom_point() +
  labs(title = "Optimal Degree vs. Principal Components",y="Optimal Degree", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))

```

## Logistic Regression Model
\tiny
```{r, include=TRUE}


SummaryDF <- data.frame(PrincipalComponents = c(optimalPC_Log), CV_Error = c(optimalCVError_Log))
SummaryDF%>%
kable(caption="Best performing logistic regression model.",digits=2)

```

```{r, include=TRUE,fig.width = 6, fig.height = 3}

bestLogModelDF <- read.csv("LogisticRegression/bestLogModelDF.csv")
PCvsCVLog <- read.csv("LogisticRegression/PCvsCV.csv")



plot <- ggplot(data = PCvsCVLog, aes(x = PrincipalComponents, y= CV_Error)) +
  geom_line()+
  geom_point()+
  labs(y="CV Error", x = "Number of Prinicipal Components") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10),
        plot.title = element_text(hjust = 0.5))


print(plot)
```


## Summary of Categorical Model Results 


```{r, include=TRUE}
ModelNames <- c('Radial SVM', 'Linear SVM', 'Polynomial SVM', 'Logistic Regression')
PrincipalComponents <- c(optimalPC_SVMRad, optimalPC_SVMLin, optimalPC_SVMPoly, optimalPC_Log)
CV_Error <- c(optimalCVError_SVMRad, optimalCVError_SVMLin, optimalCVError_SVMPoly, optimalCVError_Log)

SummaryDF <- data.frame(Model = ModelNames,PrincipalComponents = PrincipalComponents, CV_Error = CV_Error)

SummaryDF%>%
kable(caption="Best performing USDA Grade Models.",digits=4)
```
- Best Model: Radial SVM with 30 Principal Components: CV Error of .108
  - Even though the Polynomial SVM outperformed the Radial SVM slightly, it required 48 more Principal Components to be utilized.
  - The Linear SVM performed slightly worse than the Radial SVM, but required 6 less Principal Components.


# REGRESSION MODELS: TENDERNESS

``` {r}
#Data
#Creating training and test subsets for regressions
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
train<- REIMS[dpart, ]
test<- REIMS[-dpart, ]

ten3DF_train<- na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten3DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten14DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF_train<-na.omit(train[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
ten28DF_test<-na.omit(test[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])

# Completed datasets for regressions
ten3DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten3"))))])
ten14DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten14"))))])
ten28DF <- na.omit(REIMS[,!(names(REIMS) %in% union(dropBase,setdiff(responseVars, c("Ten28"))))])
```

```{r}
#PCR
#Fitting the PCR Ten3 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten3<- pcr (Ten3~., data=ten3DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten3)
mse3 <- MSEP(pcr.fit_ten3)
#M of the smallest CV error
mse3$comps[which.min(as.data.frame(mse3$val)[1,])]
#Calculating MSE with M=6 and the Test set.
pcr.pred <- predict (pcr.fit_ten3, ten3DF_test, ncomp = 6)
pcr.pred-ten3DF_test$Ten3
MSE_PCR_day3<-mean ((pcr.pred - ten3DF_test$Ten3)^2)
#Error compare to average
sqrt(MSE_PCR_day3)
mean(ten3DF$Ten3)

#Fitting the PCR Ten14 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten14<- pcr (Ten14~., data=ten14DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten14)
mse14 <- MSEP(pcr.fit_ten14)
#M of the smallest CV error
mse14$comps[which.min(as.data.frame(mse14$val)[1,])]
#Calculating MSE with M=1 and the Test set.
pcr.pred <- predict (pcr.fit_ten14, ten14DF_test, ncomp = 1)
pcr.pred-ten14DF_test$Ten14
MSE_PCR_day14<-mean ((pcr.pred - ten14DF_test$Ten14)^2)
#Error in the same scale
sqrt(MSE_PCR_day14)
mean(ten14DF$Ten14)

#Fitting the PCR Ten28 model using 10-fold cross-validation
set.seed(1)
pcr.fit_ten28<- pcr (Ten28~., data=ten28DF_train, scale = TRUE,
validation="CV")
summary (pcr.fit_ten28)
mse28 <- MSEP(pcr.fit_ten28)
#M of the smallest CV error
mse28$comps[which.min(as.data.frame(mse28$val)[1,])]
size_PCR_day28<-which.min(as.data.frame(mse28$val)[1,])
#Calculating MSE with M=2 and the Test set.
pcr.pred <- predict (pcr.fit_ten28, ten28DF_test, ncomp = 2)
pcr.pred-ten28DF_test$Ten28
MSE_PCR_day28<-mean ((pcr.pred - ten28DF_test$Ten28)^2)
#Error in the same scale
sqrt(MSE_PCR_day28)
mean(ten28DF$Ten28)
```

```{r}
#Best subset

#Tenderness at day 3
set.seed(1)
reg.fwd <- regsubsets(Ten3~., data = ten3DF_train, nvmax = 40,method = "forward") 
#Best subset selection using foward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten3~., data=ten3DF_test)
val.errors <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors[i] <- mean((ten3DF_test$Ten3-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 3",type = "o")

MSE_subset_day3<-min(val.errors)

#Tenderness at day 14
reg.fwd <- regsubsets(Ten14~., data = ten14DF_train, nvmax = 40,method = "forward") 
#Best subset selection using foward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten14~., data=ten14DF_test)
val.errors1 <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors1[i] <- mean((ten14DF_test$Ten14-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors1, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 14",type = "o")

MSE_subset_day14<-min(val.errors1)

#Tenderness at day 28
reg.fwd <- regsubsets(Ten28~., data = ten28DF_train, nvmax = 40,method = "forward") 
#Best subset selection using forward selection.
summary(reg.fwd)
res.sum <- summary(reg.fwd)

#Loop to calculate CV test error
test.mat <- model.matrix(Ten28~., data=ten28DF_test)
val.errors2 <- rep(NA, 20)
coefi<-seq(0)

for (i in 1:20) {
  coefi<-coef(reg.fwd , id = i)
  pred <- test.mat[, names(coefi)] %*% coefi
  val.errors2[i] <- mean((ten28DF_test$Ten28-pred)^2)
}

# Graph of Model size vs. MSE
plot(seq(1,20), val.errors, log="y", xlab = "number of variables", ylab="MSE",
     main="Model size vs. MSE (test) tenderness at day 28",type = "o")

MSE_subset_day28<-min(val.errors2)
```

```{r}
#Splines
#Tenderness at day 3
set.seed(1)
dpart <- createDataPartition(REIMS$USDA.grade, p = 0.6, list = F)
ten3<-na.omit( dat1[,-c(1,3:4)])
ten3_train<-ten3[dpart,]
ten3_test<-ten3[-dpart,]
err3<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten3~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten3_train)
preds<- predict(gam, newdata = ten3_test)
preds
err3[i]<-mean((ten3_test$Ten3 - preds)^2)
}
Spline3<-data.frame(N.comp=seq(1,10),MSE=err3)
which.min(Spline3$MSE)
min(Spline3$MSE)


#Tenderness at day 14
set.seed(1)
ten14<-na.omit( dat1[,-c(1:2,4)])
ten14_train<-ten14[dpart,]
ten14_test<-ten14[-dpart,]
err14<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten14~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten14_train)
preds<- predict(gam, newdata = ten14_test)
preds
err14[i]<-mean((ten14_test$Ten14 - preds)^2)
}
Spline14<-data.frame(N.comp=seq(1,10),MSE=err14)
which.min(Spline14$MSE)
min(Spline14$MSE)

# Tenderness at day 28
set.seed(1)
ten28<-na.omit( dat1[,-c(1:3)])
ten28_train<-ten28[dpart,]
ten28_test<-ten28[-dpart,]
err28<-0
for (i in 1:10) {
p<-i
form<-as.formula(paste0("Ten28~",paste0("s(Dim.",1:p,",3)",collapse="+")))
gam<-gam(formula=form,data=ten28_train)
preds<- predict(gam, newdata = ten28_test)
preds
err28[i]<-mean((ten28_test$Ten28 - preds)^2)
}
Spline28<-data.frame(N.comp=seq(1,10),MSE=err28)
which.min(Spline28$MSE)
min(Spline28$MSE)

```

```{r}
#PCA-Lasso
lambda_max_over<-4
#Tenderness day 3
LassoTen3Results <- read.csv("Lasso/Results/Ten3.csv")
BestTen3LassoModel <- readRDS("Lasso/BestModels/BestLassoTen3Model.rds")
BestTen3LassoModel

optimalPC <- LassoTen3Results$PrincipalComponents[which.min(LassoTen3Results$CV_MSE)]
optimal_TestError <-  min(LassoTen3Results$CV_MSE)
optimalLambda <- LassoTen3Results$Best_Lambda[which.min(LassoTen3Results$CV_MSE)]
data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

MSE_Lasso_day3<-optimal_TestError

#Tenderness day 14
LassoTen14Results <- read.csv("Lasso/Results/Ten14.csv")
BestTen14LassoModel <- readRDS("Lasso/BestModels/BestLassoTen14Model.rds")
BestTen14LassoModel

optimalPC <- LassoTen14Results$PrincipalComponents[which.min(LassoTen14Results$CV_MSE)]
optimal_TestError <-  min(LassoTen14Results$CV_MSE)
optimalLambda <- LassoTen14Results$Best_Lambda[which.min(LassoTen14Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

MSE_Lasso_day14<-optimal_TestError

#Tenderness day 28
LassoTen28Results <- read.csv("Lasso/Results/Ten28.csv")
BestTen28LassoModel <- readRDS("Lasso/BestModels/BestLassoTen28Model.rds")
BestTen28LassoModel

optimalPC <- LassoTen28Results$PrincipalComponents[which.min(LassoTen28Results$CV_MSE)]
optimal_TestError <-  min(LassoTen28Results$CV_MSE)
optimalLambda <- LassoTen28Results$Best_Lambda[which.min(LassoTen28Results$CV_MSE)]

data.frame(optimalPC = optimalPC, optimal_TestError = optimal_TestError, optimalLambda = optimalLambda)

MSE_Lasso_day28<-optimal_TestError
```

## Summary results
\tiny
```{r,include=TRUE,size="huge"}

data.frame(  Model = "Best Subset", 
             Size = "1", MSE_day3=MSE_subset_day3,
             Size = "1", MSE_day14=MSE_subset_day14,
             Size = "1", MSE_day28=MSE_subset_day28)%>%
  bind_rows(data.frame(Model = "PCA-Lasso", 
            Size ="18", MSE_day3=MSE_Lasso_day3, 
            Size ="10", MSE_day14=MSE_Lasso_day14, 
            Size ="58", MSE_day28=MSE_Lasso_day28))%>%
  bind_rows(data.frame(Model = "PCR", 
            Size ="6", MSE_day3= MSE_PCR_day3, 
            Size ="1", MSE_day14= MSE_PCR_day14, 
            Size ="2", MSE_day28= MSE_PCR_day28)) %>%
  bind_rows(data.frame(Model = "PCA-Splines", 
            Size ="5", MSE_day3= min(Spline3$MSE), 
            Size ="2", MSE_day14= min(Spline14$MSE), 
            Size ="4", MSE_day28= min(Spline28$MSE)))%>% 
kable(caption="Mean Squared Error (MSE) of regression models for tenderness at day 3, 14, and 28",digits=2)
```