---
title: "Illinois DOC Sentencing Machine Learning Analysis"
author: Juan Gonzalez, Aaron Graff, Ilijah Pearson, Hope Winsor
format: pptx
editor: visual
---

## Dataset and Preparation

Our data comes from Kaggle, it consists of 3 CSV files with data scraped off the Illinois Department Of Corrections website.

::: incremental
-   Contains information 61,052 individuals who have been incarcerated in Illinois from 1952-2018.

-   "Person" dataset includes simple descriptive information on the person: Name, DOB, Sentencing Date, Race, Eye Color, Hair Color, etc...

-   "Marks" has multiple rows per person listing any additional descriptive features like scars and tattoos.

-   "Sentencing" is information on the charges faced by the individual, the crime categorized by Illinois' felony classes, and our response variable, sentencing time.
:::

## Research Questions

::: incremental
-   What predictors have the strongest effect on sentencing length?

-   How accurately can we predict sentencing length?

-   What predictors have the strongest effect on offence category?

-   How accurately can we predict offence categories?
:::

## KNN

## Linear Regression

## LASSO

:::: columns
::: column

### LASSO model = RSS + $\lambda \sum_{j = 1}^{p} \beta_{j}^2$

where $\lambda$ is the tuning parameter, to improve interpretability, use penalty to perform feature selection

- cross validation selected $\lambda = 0.01$

```{r, echo=FALSE, message = FALSE}
library(tidymodels)
library(knitr)
library(tidyverse)
library(ggplot2)
```
:::
::: column

```{r}
lasso_terms <- read.csv("../Code/LinearRegressionOutputs/lasso_terms.csv")
kable(lasso_terms[,c("term", "estimate")])
```

:::
::::


## Forward and Backward Subset
:::: columns
::: column
```{r}
vars <- read.csv("../Code/LinearRegressionOutputs/ranked_vars.csv")
merge_vars <- vars[1:8,c("step", "variable")]
merge_vars$"step (cont)" <- vars[9:16, "step"]
merge_vars$"variable (cont)" <- vars[9:16, "variable"]
kable(merge_vars)
```
:::

::: column
```{r}
rmse <- read.csv("../Code/LinearRegressionOutputs/rmse_v_terms.csv")

ggplot(rmse) + geom_point(aes(num_terms, RMSE, color = RMSE == min(RMSE))) + scale_color_discrete(labels = c("larger than optimum", "minimum RMSE")) + labs(x = "Number of Terms", color = "", caption = "dropped terms: num_scars, sex, num_other_marks, height")
```
:::
::::

## Generalized Additive Model (GAM)

::: incremental
-   Extension of linear regression that allows each predictor to have a continuous nonlinear effect on the response

-   80:20 split

-   Used 10 fold CV to tune the degrees of freedom

-   On the test set the GAM achieved:

-   RMSE = 12.20631

-   Meaning that our prediction error is $12.20631$ years.

-   R\^2=0.7139504

-   Meaning that our model explains 71% of the variability in sentencing length on new data.
:::

## GAM

```{r, echo=FALSE, fig.width=12, fig.height=8.25}

library(tidyverse)
library(tidymodels)
library(knitr)
gam_fit <- readRDS("~/project-5/Code/gam_fit.rds")
sent_train <- readRDS("~/project-5/Code/GAM_train.rds")
train_pred <- readRDS("~/project-5/Code/Gam_pred.rds")




pred_dat <- train_pred |>
  group_by(year_adm) |>
  summarise(
    .pred       = mean(.pred,       na.rm = TRUE),
    .pred_lower = mean(.pred_lower, na.rm = TRUE),
    .pred_upper = mean(.pred_upper, na.rm = TRUE),
    .groups     = "drop"
  )

out_dat <- sent_train |>
  select(year_adm, current_sentence)

ggplot() +
  geom_point(
    data = out_dat,
    aes(x = year_adm, y = current_sentence),
    alpha = 0.007
  ) +
  geom_line(
    data = pred_dat,
    aes(x = year_adm, y = .pred),
    color = "coral4"
  ) +
  geom_line(
    data = pred_dat,
    aes(x = year_adm, y = .pred_lower),
    color = "coral4",
    linetype = "dashed"
  ) +
  geom_line(
    data = pred_dat,
    aes(x = year_adm, y = .pred_upper),
    color = "coral4",
    linetype = "dashed"
  ) +
  coord_cartesian(xlim = c(1980, 2018)) +
  labs(
    x = "Year admitted",
    y = "Sentence length",
    title = "GAM effect for year admitted 1980–2018"
  ) +
  theme_bw(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text  = element_text(size = 14)
  )



```

## Random Forest

::::: columns
::: column
-   Chosen to deal with correlation amongst trees.
-   58,836 observations. 70:30 Split $\sim$ (41,200 \| 17,600)
-   Fit 500 trees with 4 randomly chosen predictors for each tree.
-   Out-of-Bag error is calculated by checking the performance of each tree on a subset of unseen data during the training process.\
:::

::: column
```{r, echo=FALSE, fig.width=12, fig.height=8.25}

library(vip)
library(tidyverse)
library(randomForest)
library(pdp)

#You must use download link to download the rds file locally and adjust it to your pathway to the rds.
#
#rf_fit <- readRDS("~/Documents/RFfit_model_juan.rds")
rf_fit <- readRDS("../Code/RFfit_model_juan.rds")



vi <- vi(rf_fit$fit) %>%
  top_n(5, Importance) %>%
  mutate(Variable = recode(Variable,
                           year_adm = "Year Admitted",
                           parent_institution = "Parent Institution",
                           total_counts = "Total Counts",
                           time_sentenced_prior = "Prior Sentencing Time",
                           class = "Illinois Felony Class",
                           ))

vip(vi, aesthetics = list(fill = "coral4"))  +
  ggtitle("5 Most Important Predictors in our Random Forests Model") + theme_bw() + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text = element_text(size = 13),
        plot.title = element_text(size = 16, hjust = 0.5))



```
:::
:::::

## Predictive Performance - Random Forests

::::: columns
::: column
-   **12.11** OOB RMSE / **12. 18** Test RMSE. Similar RMSEs is good (generalizes well).
-   As sentencing length goes up, our models prediction becomes more variable. However, the models' errors scale proportionally to sentencing length. Meaning we don't typically get crazy predictions for lower sentencing lengths.
:::

::: column
```{r, echo=FALSE, fig.width=8}

rmses <- read_csv("../CSV Files/RF_RMSEbySentencing.csv")


ggplot(data = rmses, aes(x = current_sentence_rnd, y = rmse)) + 
  labs(title = "RMSE by Sentence Length", x = "Sentencing Time", y = "RMSE") +
  geom_smooth(method = "loess", se = FALSE, color = "coral4") + geom_line() +
  theme_bw()+theme(plot.title = element_text(hjust = 0.5))

```
:::
:::::

## Conclusion

## References

-https://www.kaggle.com/datasets/davidjfisher/illinois-doc-labeled-faces-dataset -https://www.idoc.state.il.us

## Questions?

## Bonus Slides - Complete Variable Importance for RF

```{r, echo=FALSE, fig.width=8}
vi_all <- vi(rf_fit$fit) %>%
  top_n(20, Importance) #%>%
  # mutate(Variable = recode(Variable,
  #                          year_adm = "Year Admitted",
  #                          parent_institution = "Parent Institution",
  #                          total_counts = "Total Counts",
  #                          time_sentenced_prior = "Prior Sentencing Time",
  #                          class = "Illinois Felony Class",
  # ))

vip(vi_all, num_features = 20,aesthetics = list(fill = "coral4"))  +
  ggtitle("Variable Importance Plot for our Random Forests Model") + theme_bw()




```
