---
title: "DSCI445 - Group Project"
authors: "Amanda Bartels, Mohamed Habibniaykoochesfahani, Isabella Kukor, Casey Martin"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


## Basic R

# Remove NA fields

```{r, include=FALSE}

dataset <- read.csv("apartments_for_rent_classified_10K.csv", sep = ";")

head(dataset)
sapply(dataset, class)

# View summary statistics for each column
summary(dataset)

table(dataset$category)
table(dataset$currency)
table(dataset$state)
table(dataset$bedrooms)
table(dataset$fee)
table(dataset$has_photo)
table(dataset$pets_allowed)
table(dataset$price_type)
table(dataset$source)

```

# Remove columns that do not contain useful information:

Removed 'currency' column because all values in it are "USD".
Removed 'fee' column because all values are "No" so this will not provide insights.
Kept value 'has_photo' because it could correlate with listing quality and affect price.
Removed 'price_display' column because it is a repeat of price column.
Kept(?) 'source' column tracks the source of the listing (i.e. different websites). This is only useful for comparing listings across sources which we will/not be doing in this analysis.
'title' and 'body' columns are removed because they contain text descriptions which are not going to be analyzed in this study.
Removed 'address' because it's not useful and has a large ammount of NAs.
Removed 'latitude', 'longitude', 'address', 'cityname', and 'time' columns because they are not interpretable in this study.
Only 4 rows in the entire dataset have \textit{category} values other than 'apartment', so these rows and consequently the "category" column are deleted.  
Only 2 rows in the entire dataset have \textit{price\_type} values other than 'Monthly', so these rows and consequently the "price\_type" column are deleted.

```{r, include=FALSE}
library(tidyverse)

# Remove rows
dataset <- dataset[dataset$category=='housing/rent/apartment',]
dataset <- dataset[dataset$price_type=='Monthly',]

# Remove columns
dataset <- dataset %>%
  select(
    -currency,
    -fee,
    -price_display,
    -title,
    -body,
    -price_type,
    -address,
    -cityname,
    -time,
    -category,
    -longitude,
    -latitude
    
    
    # Uncomment the line below to remove 'source' (website listing was posted on)
    # -source
  )

# View cleaned dataset
str(dataset)


```


# Remove null values.

```{r, include=FALSE}
# Replace "null" with NA across all character columns in the dataset for easier handling
dataset[] <- lapply(dataset, function(x) {
  if (is.character(x)) {
    x[x == "null"] <- NA
  }
  return(x)
})

# Get a summary of NA counts for each column in the dataset
colSums(is.na(dataset))

# Remove rows with any NA values
dataset <- na.omit(dataset)

# Check the structure and summary of the dataset to verify changes
str(dataset)
summary(dataset)

```


# Standardizes categorical values for consistency.

```{r}

# Change `has_photo` entries containing "No Thumbnail" into "No" category so only categories are "Yes" or "No"
dataset$has_photo[dataset$has_photo == "No Thumbnail"] <- "No"

# Correct the special character in `source`
dataset$source[dataset$source == "RENTCaf\xe9"] <- "RENTCafé"

```

# Spread amenities column

```{r}
dataset_spread <- dataset %>%
  # Split and widen amenities
  mutate(amenities_split = str_split(amenities, ",")) %>%
  unnest(amenities_split) %>%
  mutate(has_amenity = 1) %>%
  pivot_wider(names_from = amenities_split, values_from = has_amenity, values_fill = 0) %>%
  select(-amenities, -id)
```

# Train test split
```{r}
library(tidymodels)

set.seed(445)

dataset_split <- initial_split(dataset_spread, prop = 0.6)
dataset_train <- training(dataset_split)
dataset_test <- testing(dataset_split)
```

#"Lasso" and "Ridge" regressions are used to include regularization to the linear regression and prevent overfitting.
```{r}

library(tidymodels)
library(glmnet)
library(ggplot2)

set.seed(445)

rec <- recipe(price ~ ., data = dataset_train) %>%
  step_novel(all_nominal_predictors()) %>% # Handle unseen factor levels
  step_dummy(all_nominal_predictors()) %>% # Convert categorical variables to dummy variables
  step_zv(all_predictors()) %>%  # Remove zero-variance 
  step_normalize(all_numeric_predictors()) # Normalize numeric predictors

prepped_rec <- prep(rec, training = dataset_train)

x_train <- bake(prepped_rec, new_data = dataset_train)
x_test <- bake(prepped_rec, new_data = dataset_test)

x_train <- as.matrix(select(x_train, -price))
y_train <- dataset_train$price

x_test <- as.matrix(select(x_test, -price))
y_test <- dataset_test$price

# Fit ridge regression model with cross-validation
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_ridge$lambda.min

# Fit final ridge regression model with best lambda
ridge_model <- glmnet(x_train, y_train, alpha = 0, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(ridge_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)


# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(ridge_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_ridge$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_ridge$lambda

# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  # Add circle around best_lambda
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20, label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  # Add annotation for best lambda

# Print the plot
print(cv_rmse_plot)

comparison_plot <- data.frame(Actual = y_test, Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Reference line y=x
  labs(title = "Actual vs Predicted Values",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)

comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  # Predicted values

comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   # Line for actual values
  geom_line(aes(y = Predicted, color = "Predicted"), size = 1, linetype = "dashed",alpha = 0.8) +  # Line for predicted values
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  # Custom colors
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```

```{r}
set.seed(445)

# Fit Lasso regression model with cross-validation (alpha = 1 for Lasso)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_lasso$lambda.min

# Fit final Lasso regression model with best lambda
lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(lasso_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)

# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(lasso_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_lasso$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_lasso$lambda



# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  # Add circle around best_lambda
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20, label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  # Add annotation for best lambda

# Print the plot
print(cv_rmse_plot)

# Plot: Actual vs Predicted for Test Data
comparison_plot <- data.frame(Actual = y_test, Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Reference line y=x
  labs(title = "Actual vs Predicted Values for Lasso",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)



comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  # Predicted values

comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   # Line for actual values
  geom_line(aes(y = Predicted, color = "Predicted",alpha = 0.8), size = 1, linetype = "dashed", alpha = 0.8) +  # Line for predicted values
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  # Custom colors
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```

Despite of utilizing cross-validation to select an optimal regularization parameter, the Ridge regression model shows signs of overfitting. There is a noticeable gap between training and test performance, with the model performing better on the training data, indicating it may have learned patterns specific to the training set that don't generalize well. While regularization has helped reduce overfitting to some extent, the results remain relatively unchanged when switching to Lasso, suggesting the issue isn't due to the type of regularization but rather the dataset's complexity. Given that cross-validation with ten folds is a rational choice, the next steps should focus on using more complex models like decision trees, random forests, or gradient boosting, which may better capture the underlying patterns of the data.