---
output: pdf_document
geometry: portrait
fontsize: 12pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,message=FALSE)
```
\newpage
\begin{center}
\vspace*{\fill} % Push content to the vertical center
 
{\Large DSCI445 - Group Project}
\vspace{3cm} % Add vertical space

{\Huge Apartment Rental Rates Across the United States}

\vspace{3cm} % Add vertical space

{\large Amanda Bartels, Mohamad Habibniaykoochesfahani, Isabella Kukor, Casey Martin}

\vspace{3cm} % Add vertical space

{\normalsize December, 2024}

\vspace*{\fill} % Push content to the vertical center
\end{center}
\newpage


# Dataset introduction and cleaning
The "Apartment for Rent Classified" dataset, donated on December 25, 2019, is a multivariate dataset containing classified ads for apartments for rent in the USA. This dataset consists of 10,000 instances with 21 categorical and integer features.
Attributes  include details like apartment amenities (e.g., AC, gym, pool), bathrooms, bedrooms, rental price, square footage, location (address, city, state), and more. Each variable in the dataset was examined for NA values and all NA data points were removed. Categorical data counts explored potential levels with 1 count. For example:
  - Only 1 apartment was pay type “Weekly.”
  - 1 apartment was both “Monthly and Weekly.”
  - 9998 apartments were “Monthly.”
This led to removing unnecessary variables in the dataset, such as address, time, longitude, latitude, etc.


# Libraries
Different libraries are used in this study and each of these libraries serves a distinct purpose in data manipulation, modeling, evaluation, and parallel computing, helping streamline and optimize the workflow.
```{r}
library(tidyverse)
library(tidymodels)
library(glmnet)
library(leaps)
library(rpart.plot)
library(vip)
library(doParallel)
library(knitr)
```


# Dataset introduction

```{r}
dataset <- tryCatch(
  {
    read.csv("apartments_for_rent_classified_10K.csv", sep = ";")
  },
  error = function(e) {
    # If the first path fails, try the second path
    read.csv("group project/apartments_for_rent_classified_10K.csv", sep = ";")
  }
)

head(dataset)
sapply(dataset, class)

# View summary statistics for each column
summary(dataset)

table(dataset$category)
table(dataset$currency)
table(dataset$state)
table(dataset$bedrooms)
table(dataset$fee)
table(dataset$has_photo)
table(dataset$pets_allowed)
table(dataset$price_type)
table(dataset$source)

```

# Removing Useless Columns
The detailed procedure of cleaning the data is as follows:  
- Removed 'currency' column because all values in it are "USD".  
- Removed 'fee' column because all values are "No" so this will not provide insights.  
- Kept value 'has_photo' because it could correlate with listing quality and affect price.  
- Removed 'price_display' column because it is a repeat of price column.  
- Kept 'source' column tracks the source of the listing (i.e. different websites). This is only useful for comparing listings across  sources which we will be doing in this analysis.  
- 'title' and 'body' columns are removed because they contain text descriptions which are not going to be analyzed in this study.
- Removed 'address' because it's not useful and has a large amount of NAs.  
- Removed 'latitude', 'longitude', 'address', 'cityname', and 'time' columns because they are not interpretable in this study.  
- Only 4 rows in the entire dataset have \textit{category} values other than 'apartment', so these rows and consequently the "category" column are deleted.    
- Only 2 rows in the entire dataset have \textit{price\_type} values other than 'Monthly', so these rows and consequently the "price\_type" column are deleted.  

```{r}
# Remove rows
dataset <- dataset[dataset$category=='housing/rent/apartment',]
dataset <- dataset[dataset$price_type=='Monthly',]

# Remove columns
dataset <- dataset %>%
  select(
    -currency,
    -fee,
    -price_display,
    -title,
    -body,
    -price_type,
    -address,
    -cityname,
    -time,
    -category,
    -longitude,
    -latitude
    
    
    # Uncomment the line below to remove 'source' (website listing was posted on)
    # -source
  )

# View cleaned dataset
str(dataset)


```


# Remove null values.

```{r}
# Replace "null" with NA across all character columns in the dataset for easier handling
dataset[] <- lapply(dataset, function(x) {
  if (is.character(x)) {
    x[x == "null"] <- NA
  }
  return(x)
})

# Get a summary of NA counts for each column in the dataset
colSums(is.na(dataset))

# Remove rows with any NA values
dataset <- na.omit(dataset)

# Check the structure and summary of the dataset to verify changes
str(dataset)
summary(dataset)

```


# Standardizing Categorical Values for Consistency.

```{r}

# Change `has_photo` entries containing "No Thumbnail" into "No" 
dataset$has_photo[dataset$has_photo == "No Thumbnail"] <- "No"

# Correct the special character in `source`
dataset$source[dataset$source == "RENTCaf\xe9"] <- "RENTCafe"

```

# Spread Amenities Column

```{r}
dataset_spread <- dataset %>%
  mutate(amenities_split = str_split(amenities, ",")) %>%
  unnest(amenities_split) %>%
  mutate(has_amenity = 1) %>%
  pivot_wider(names_from = amenities_split, values_from = has_amenity, values_fill = 0) %>%
  select(-amenities, -id)

dataset_spread_copy <- dataset_spread

```

# Train-Test Split
```{r}
set.seed(445)
dataset_spread$bathrooms <- as.numeric(dataset_spread$bathrooms)
dataset_spread$bedrooms <- as.numeric(dataset_spread$bedrooms)
mean(dataset_spread$price)
dataset_spread$bi_price[dataset_spread$price >=1356.544] <- "1"
dataset_spread$bi_price[dataset_spread$price < 1356.544] <- "0"
dataset_spread$bi_price <- as.factor(dataset_spread$bi_price)
dataset_split <- initial_split(dataset_spread, prop = 0.6)
dataset_train <- training(dataset_split)
dataset_test <- testing(dataset_split)
```
\newpage
# Distribution Map of the Apartments
In order to understand the pricing of apartments in the US, it is important to visualize where our data is courced from as the cost of living heavily depends on the location. The map below shows that the vast majority of states are represented in the data set. The map does not show DC data points although they do exist. Texas had the majority of apartment data while states in gray like West Virigina did not have any apartment data points.

```{r}
##Create dataset for the states with longitude and latitude. map_data function is in ggplot2
USmap <- map_data("state")

##Need to change abbreviations to full state name to match with USMap datatset 
dataset_spread$state_full <- state.name[match(dataset_spread$state, state.abb)]
##Lowercase letters for exact match
dataset_spread$state_full <- tolower(dataset_spread$state_full)
##Save the new name for state
dataset_spread$state <- as.factor(dataset_spread$state_full)

##Determine frequency for each state by the amount of records. *DC not mapped on map*
state_counts <- dataset_spread %>%
  group_by(state) %>%
  summarise(frequency = n())

##Join frequency for each state in the dataset
USmap <- USmap %>%
  left_join(state_counts, by = c("region" = "state"))

##Map out the United States using ggplot, and color by frequency used. 
ggplot(USmap, aes(x = long, y = lat, group = group, fill = frequency)) +
  geom_polygon(color = "black") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "gray") +
  theme_minimal() +
  labs(title = "US Map with Highlighted States", fill = "Frequency")


```

# Summary of the Statistics
Below are small examples of the average prices of certain predictors of interest. Although we will use all the remaining predictors, these tables do provide some information on the raw data before any analysis occurs. 
```{r}
## Average income for each state 
table <- dataset_spread %>%
  group_by(state) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table

## Average price for each listing source
table2 <- dataset_spread %>%
  group_by(source) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table2
## Average price based on bathrooms
table3 <- dataset_spread %>%
  group_by(bathrooms) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table3
##Average price based on bedrooms 
table4 <- dataset_spread %>%
  group_by(bedrooms) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table4


```

\newpage

# Linear Model
The simpliest model to first examine was the linear model since the resposnse variable of price is continious. In this case we were more interested in seeing how the model preforms and less interested in which predictors were signifcant. 
```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
#glimpse(dataset_train)

pricelm <- recipe(price ~ square_feet + bathrooms + bedrooms + has_photo +
                    pets_allowed + state + source , data = dataset_train)

workflow <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm)

fitted_model <- workflow %>% 
  fit(data = dataset_train)

predictions <- fitted_model %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics <- predictions %>% 
  metrics(truth = price, estimate = .pred)

print(metrics)
```

\newpage

# Logistic Regression
Average price of house, above mean 1, below 0. See if we can get better predictions from binary logistic regression. In this model we get 19.45% error rate using a logistic model with the same predictors. 

```{r}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog <- recipe(bi_price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed + state + source , data = dataset_train)

workflow_log <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog)

fitted_model_log <- workflow_log %>% 
  fit(data = dataset_train)

predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)
predictions_log

class_predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)
class_predictions_log

metrics_log <- class_predictions_log %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log <-class_predictions_log %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

print(matrix_log)

(239+138)/(1104+239+138+457)

print(metrics_log)

```

\newpage

# Choosing the correct predictors:
In our model, we have a significant amount of factor variables, where there is a lot of categorical data. When these categories are broke up there are 90+ variables that could be included in the model. For example: state includes all 47 states that is contained in the data set. Knowing which predictors are significant to the model is difficult to determine so we used forward and backward selection to weed out of all the variables which would be the best to make predictions on. 

When talking about price of apartments, we know location can have an impact over the general price. For these models, we were more curious about separating out amenities, bedrooms, ad type, etc. 

Based on forward selection and using the BIC and ADJR we see that the predictors of availability of a playground, hot tub, view, wood floors, washer dryer, patio/ deck, elevator, garbage disposal, refrigerator are all amenities that influence price in the model. Rent Lingo and ListedBuy are the only two sources of posting that tend to influence the predictability of price. Lastly, the square footage and bathrooms are significant to the price of the apartment. 

The adjusted r square value indicates a slightly different model would be best but it only explains about 25.28% of the variability. That model includes all the above predictors and adds Tennis and Pool to the list of significant predictors. We know from our linear model that included every variable, specifically the state variable is going to explain a lot of our variability, and with all predictors in that model, naturally R^2 will be higher. 

The backwards selection indicated the same model as the forward selection did for BIC and AdjR. This leads us to decrease our predictors on our models and disregard insignificant amenities that do not influence the price point of the apartment. 

```{r,include=FALSE}

sub_for <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "forward", nvmax=20)
summary(sub_for)
summary(sub_for)$bic
summary(sub_for)$adjr2

sub_back <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "backward", nvmax = 20)
summary(sub_back)
summary(sub_back)$bic
summary(sub_back)$adjr2
```
Linear model with predictors shows slight improvement in our R squared value where our model is explaining 58.9%.

```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
#glimpse(dataset_train)
#dataset_spread$`Garbage Disposal`

pricelm2 <- recipe(price ~ square_feet + bathrooms + Pool + View + Refrigerator+
                     `Patio/Deck` +Playground +`Hot Tub` + Elevator +
                     `Wood Floors`+ `Garbage Disposal`+ state + source ,
                   data = dataset_train)

workflow2 <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm2)

fitted_model2 <- workflow2 %>% 
  fit(data = dataset_train)

predictions2 <- fitted_model2 %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics2 <- predictions2 %>% 
  metrics(truth = price, estimate = .pred)

print(metrics2)
```
\newpage

# Logisitc Regression with Predictors

Not much changed in our prediction error or accuracy, but the mean of all prices was selected as an arbitrary cut off point. Logistic regression tends to outperform the linear model in accuracy for predicting whether the prices will be lower or higher than average, but since we dont know if this cut off point of the average of the overall price is a good classification compared to the other models. It is not yet determined whether the logistic model utilizing classification of low and high is better predictor than examining the prices of the model as a continuous response variable in a linear model, or other options.

```{r}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog2 <- recipe(bi_price ~ square_feet + bathrooms + Pool + View + Refrigerator+ `Patio/Deck` +Playground +`Hot Tub` + Elevator + `Wood Floors`+ `Garbage Disposal`+ state + source, data = dataset_train)

workflow_log2 <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog2)

fitted_model_log2 <- workflow_log2 %>% 
  fit(data = dataset_train)

predictions_log2 <- fitted_model_log2 %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)
predictions_log2

class_predictions_log2 <- fitted_model_log2 %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)
class_predictions_log2

metrics_log2 <- class_predictions_log2 %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log2 <-class_predictions_log2 %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

print(matrix_log2)

(235+138)/(1104+235+138+461)

print(metrics_log2)

``` 

\newpage

# Ridge Regression Technique 
In this step, Ridge and Lasso techniques are used to improve the performance of the linear model through employing regularization to avoid overfitting. This is achieved through calibrating and choosing the best $\lambda$ parameter based on the least rmse value.
```{r,warning=FALSE}
set.seed(445)

rec <- recipe(price ~ ., data = dataset_train) %>%
  step_novel(all_nominal_predictors()) %>% # Handle unseen factor levels
  step_dummy(all_nominal_predictors()) %>% # Convert categorical variables to dummy variables
  step_zv(all_predictors()) %>%  # Remove zero-variance 
  step_normalize(all_numeric_predictors()) # Normalize numeric predictors

prepped_rec <- prep(rec, training = dataset_train)

x_train <- bake(prepped_rec, new_data = dataset_train)
x_test <- bake(prepped_rec, new_data = dataset_test)

x_train <- as.matrix(select(x_train, -price))
y_train <- dataset_train$price

x_test <- as.matrix(select(x_test, -price))
y_test <- dataset_test$price

# Fit ridge regression model with cross-validation
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_ridge$lambda.min

# Fit final ridge regression model with best lambda
ridge_model <- glmnet(x_train, y_train, alpha = 0, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(ridge_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)


# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(ridge_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / 
  sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_ridge$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_ridge$lambda

# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20,
           label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1) 

# Print the plot
print(cv_rmse_plot)

comparison_plot <- data.frame(Actual = y_test, 
                              Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  
  labs(title = "Actual vs Predicted Values",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)

comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  
comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   
  geom_line(aes(y = Predicted, color = "Predicted"), size = 1, 
            linetype = "dashed",alpha = 0.8) + 
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```

\newpage

# Lasso Regression Technique 

```{r,warning=FALSE}
set.seed(445)

# Fit Lasso regression model with cross-validation (alpha = 1 for Lasso)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_lasso$lambda.min

# Fit final Lasso regression model with best lambda
lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(lasso_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)

# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(lasso_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_lasso$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_lasso$lambda



# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20,
           label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  

# Print the plot
print(cv_rmse_plot)

# Plot: Actual vs Predicted for Test Data
comparison_plot <- data.frame(Actual = y_test, 
                              Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  
  labs(title = "Actual vs Predicted Values for Lasso",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)



comparison_df <- data.frame(Index = 1:length(y_test),  
                            Actual = y_test,         
                            Predicted = as.vector(predictions_test))  
comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +  
  geom_line(aes(y = Predicted, color = "Predicted",alpha = 0.8), size = 1, 
            linetype = "dashed", alpha = 0.8) +  
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE",
             "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

print(error_summary)
```

Cross-validation is used to select an optimal regularization parameter, helping reduce overfitting to some extent; However, the results for both Lasso and Ridge models show a slight improvement comparing to the previously developed linear model. This suggests that the issue isn't due to the type of regularization but rather the dataset's complexity. Given that cross-validation with ten folds is a rational choice, the next steps should focus on using more complex models like decision trees, random forests, or gradient boosting, which may better capture the underlying patterns of the data.

\newpage

# Tree Based Methods

## Remove rare levels

```{r}
# Threshold for minimum occurrences
threshold <- 10

# Identify rows to keep
rows_to_keep <- seq(TRUE, nrow(dataset_spread_copy))

# Process each column
for (col_name in colnames(dataset_spread_copy)) {
  if (is.character(dataset_spread_copy[[col_name]]) || is.factor(dataset_spread_copy[[col_name]])) {
    col_table <- table(dataset_spread_copy[[col_name]])
    valid_categories <- names(col_table[col_table >= threshold])
    rows_to_keep <- rows_to_keep & dataset_spread_copy[[col_name]] %in% valid_categories
  }
}

# Filter the dataframe
dataset_spread_trees <- dataset_spread_copy[rows_to_keep, ]
```

Due to the extreme inbalance of the dataset, levels with less than 10 values were removed as they could cause a mismatch between the training and testing sets as well as the different folds for cross validation.

## Recreate train test split
```{r}
set.seed(445)

dataset_spread_trees <- dataset_spread_trees %>% select(-state, -source)


dataset_trees_split <- initial_split(dataset_spread_trees, prop = 0.6)
dataset_trees_train <- training(dataset_trees_split)
dataset_trees_test <- testing(dataset_trees_split)
```

\newpage

# Decision Tree
## Tune depth
```{r, cache=TRUE}
set.seed(445)
train_cv_5fold <- vfold_cv(dataset_trees_train, v = 5)

tree_spec <- decision_tree(tree_depth = tune("depth")) |>
  set_engine("rpart") |>
  set_mode("regression")

tree_rec <- recipe(price ~ ., data = dataset_trees_train)

tree_wf <- workflow() |>
  add_model(tree_spec) |>
  add_recipe(tree_rec)

tune_df <- data.frame(depth = seq(1:10))

tree_tune_fit <- tree_wf |>
  tune_grid(resamples = train_cv_5fold, grid = tune_df)

tree_tune_fit |>
  autoplot()

show_best(tree_tune_fit, metric = "rmse", n = 1)

tree_final <- finalize_workflow(tree_wf, select_best(tree_tune_fit, metric = "rmse"))

tree_final_fit <- tree_final |>
  fit(data = dataset_trees_train)

tree_final_fit |>
  extract_fit_engine() |>
  rpart.plot(roundint=FALSE)

vip(tree_final_fit)

dataset_pred_tree <- predict(tree_final_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_tree))*(sum(dataset_trees_test$price - dataset_pred_tree$.pred))^2)
RMSE
```

Tuning depth gives an RMSE of 35.64 and depth equal to just one layer.

## Tune alpha
```{r, cache=TRUE}
set.seed(445)
train_cv_5fold <- vfold_cv(dataset_trees_train, v = 5)

tree_spec <- decision_tree(tree_depth = 1, cost_complexity = tune("alpha")) |>
  set_engine("rpart") |>
  set_mode("regression")

tree_rec <- recipe(price ~ ., data = dataset_trees_train)

tree_wf <- workflow() |>
  add_model(tree_spec) |>
  add_recipe(tree_rec)

tune_df <- data.frame(alpha = 10^seq(-3, -1, length.out = 10))

tree_tune_fit <- tree_wf |>
  tune_grid(resamples = train_cv_5fold, grid = tune_df)

tree_tune_fit |>
  autoplot()

show_best(tree_tune_fit, metric = "rmse", n = 1)

tree_final <- finalize_workflow(tree_wf, select_best(tree_tune_fit, metric = "rmse"))

tree_final_fit <- tree_final |>
  fit(data = dataset_trees_train)

tree_final_fit |>
  extract_fit_engine() |>
  rpart.plot(roundint=FALSE)

vip(tree_final_fit)

dataset_pred_tree <- predict(tree_final_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_tree))*(sum(dataset_trees_test$price - dataset_pred_tree$.pred))^2)
RMSE
```

Tuning alpha gives a value of 0.001 and an RMSE score of 35.64, which is the same as with just the depth.

\newpage

# Random Forest

```{r, cache=TRUE}
set.seed(445)

rf_spec <- rand_forest(mtry = sqrt(.cols()), trees = 100) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("regression")

rf_rec <- recipe(price ~ ., data = dataset_trees_train)

rf_wf <- workflow() |>
  add_model(rf_spec) |>
  add_recipe(rf_rec)

rf_fit <- rf_wf |>
  fit(dataset_trees_train)

vip(rf_fit)

dataset_pred_rf <- predict(rf_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_rf))*(sum(dataset_trees_test$price - dataset_pred_rf$.pred))^2)
RMSE
```

A random forest with trees set to 100 gives an RMSE of 135.63, which is a lot worse than the decision tree, but still better than lasso, ridge, and basic linear regression. Cross validation was skipped due to extremely long computation times, but manually choosing 100 trees seemed to give approximately the best performance.

## Tune lambda and depth

```{r, cache=TRUE}
#Setup parallel processing
all_cores <- parallel::detectCores(logical = FALSE)
if(all_cores >= 32){
  all_cores <- 32
}
cl <- makePSOCKcluster(all_cores)
registerDoParallel(cl)

set.seed(445)

boost_spec <- boost_tree(trees = 5000, tree_depth = tune('depth'), learn_rate = tune("lambda")) |>
  set_engine("xgboost") |>
  set_mode("regression")

boost_rec <- recipe(price ~ ., data = dataset_trees_train) |>
  step_dummy(all_nominal_predictors())

boost_wf <- workflow() |>
  add_model(boost_spec) |>
  add_recipe(boost_rec)

tune_df <- data.frame(depth = seq(1, 5, length.out = 5), lambda = 10^seq(-3, 1, length.out = 5))

start.time <- Sys.time()

boost_tune_fit <- boost_wf |>
  tune_grid(resamples = train_cv_5fold, grid = tune_df)

end.time <- Sys.time()
time.taken <- end.time - start.time
print(paste("Time elapsed tuning:", time.taken))

boost_tune_fit |>
  autoplot()

show_best(boost_tune_fit, metric = "rmse", n = 1)

boost_final <- finalize_workflow(boost_wf, select_best(boost_tune_fit, metric = "rmse"))

boost_final_fit <- boost_final |>
  fit(data = dataset_trees_train)

vip(boost_final_fit)

dataset_pred_boost <- predict(boost_final_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_boost))*(sum(dataset_trees_test$price - dataset_pred_boost$.pred))^2)
RMSE
```

Tuning depth and lambda simultaneously gives a depth of 2 and lambda equal to 0.01. The RMSE is 173.16, making it the worst of the tree based methods.

\newpage

# Conclusion


The table below presents the Root Mean Squared Error (RMSE) for various models. The Decision Trees model has the lowest RMSE at 35, showing the best performance among the models. Linear regression shows the highest RMSE of 518, followed by Lasso and Ridge models at 474 and 473, respectively. This comparison highlights the different performance of each model, with Decision Trees outperforming others in terms of accuracy.

| Model         | RMSE   |
|---------------|--------|
| Linear        | 518    |
| Ridge         | 473    |
| Lasso         | 474    |
| **Decision Trees**| **35**     |
| Random Forest | 173    |

The key takeaways are:  

- The first model used in this study was the linear model, but the error rate didn't show an acceptable performance.  
- Logistic regression had higher accuracy when predicting on higher or lower than average apartment prices based on the predictors provided in the dataset but little improvements can be made to the model past this. The linear regression model was considered to be improved upon.  
- Although lasso and ridge regression techniques were used to avoid overfitting through employing regularization, linear model didn't show much improvement, showing that the complexity of the dataset might be high. In this step it was decided to use nonlinear models.  
- The nonlinear regression model with the best RMSE is the decision tree model, which is surprising given its simpleness compared to random forests and boosted models, especially considering that XG boost handles inbalanced datasets better.  


\newpage

# References
- “Apartment for Rent Classified.” UCI Machine Learning Repository. https://doi.org/10.24432/C5X623.

