---
title: "DSCI445 - Group Project"
authors: "Amanda Bartels, Mohamed Habibniaykoochesfahani, Isabella Kukor, Casey Martin"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


## Basic R

# Remove NA fields

```{r, include=FALSE}

dataset <- read.csv("group project/apartments_for_rent_classified_10K.csv", sep = ";")

head(dataset)
sapply(dataset, class)

# View summary statistics for each column
summary(dataset)

table(dataset$category)
table(dataset$currency)
table(dataset$state)
table(dataset$bedrooms)
table(dataset$fee)
table(dataset$has_photo)
table(dataset$pets_allowed)
table(dataset$price_type)
table(dataset$source)

```

# Remove columns that do not contain useful information:

Removed 'currency' column because all values in it are "USD".
Removed 'fee' column because all values are "No" so this will not provide insights.
Kept value 'has_photo' because it could correlate with listing quality and affect price.
Removed 'price_display' column because it is a repeat of price column.
Kept(?) 'source' column tracks the source of the listing (i.e. different websites). This is only useful for comparing listings across sources which we will/not be doing in this analysis.
'title' and 'body' columns are removed because they contain text descriptions which are not going to be analyzed in this study.
Removed 'address' because it's not useful and has a large ammount of NAs.
Removed 'latitude', 'longitude', 'address', 'cityname', and 'time' columns because they are not interpretable in this study.
Only 4 rows in the entire dataset have \textit{category} values other than 'apartment', so these rows and consequently the "category" column are deleted.  
Only 2 rows in the entire dataset have \textit{price\_type} values other than 'Monthly', so these rows and consequently the "price\_type" column are deleted.

```{r, include=FALSE}
library(tidyverse)

# Remove rows
dataset <- dataset[dataset$category=='housing/rent/apartment',]
dataset <- dataset[dataset$price_type=='Monthly',]

# Remove columns
dataset <- dataset %>%
  select(
    -currency,
    -fee,
    -price_display,
    -title,
    -body,
    -price_type,
    -address,
    -cityname,
    -time,
    -category,
    -longitude,
    -latitude
    
    
    # Uncomment the line below to remove 'source' (website listing was posted on)
    # -source
  )

# View cleaned dataset
str(dataset)


```


# Remove null values.

```{r, include=FALSE}
# Replace "null" with NA across all character columns in the dataset for easier handling
dataset[] <- lapply(dataset, function(x) {
  if (is.character(x)) {
    x[x == "null"] <- NA
  }
  return(x)
})

# Get a summary of NA counts for each column in the dataset
colSums(is.na(dataset))

# Remove rows with any NA values
dataset <- na.omit(dataset)

# Check the structure and summary of the dataset to verify changes
str(dataset)
summary(dataset)

```


# Standardizes categorical values for consistency.

```{r}

# Change `has_photo` entries containing "No Thumbnail" into "No" category so only categories are "Yes" or "No"
dataset$has_photo[dataset$has_photo == "No Thumbnail"] <- "No"

# Correct the special character in `source`
dataset$source[dataset$source == "RENTCaf\xe9"] <- "RENTCafé"

```

# Spread amenities column

```{r}
dataset_spread <- dataset %>%
  # Split and widen amenities
  mutate(amenities_split = str_split(amenities, ",")) %>%
  unnest(amenities_split) %>%
  mutate(has_amenity = 1) %>%
  pivot_wider(names_from = amenities_split, values_from = has_amenity, values_fill = 0) %>%
  select(-amenities, -id)
```

# Train test split, need to add 
```{r}
library(tidymodels)

set.seed(445)
library(tidymodels)
dataset_spread$bathrooms <- as.numeric(dataset_spread$bathrooms)
dataset_spread$bedrooms <- as.numeric(dataset_spread$bedrooms)
mean(dataset_spread$price)
dataset_spread$bi_price[dataset_spread$price >=1356.544] <- "1"
dataset_spread$bi_price[dataset_spread$price < 1356.544] <- "0"
dataset_spread$bi_price <- as.factor(dataset_spread$bi_price)
dataset_split <- initial_split(dataset_spread, prop = 0.6)
dataset_train <- training(dataset_split)
dataset_test <- testing(dataset_split)
dataset_train
```

##Map of Apartments
```{r}
##Create dataset for the states with longitude and latitude. map_data function is in ggplot2
USmap <- map_data("state")
USmap
dataset_spread
##Need to change abbreviations to full state name to match with USMap datatset 
dataset_spread$state_full <- state.name[match(dataset_spread$state, state.abb)]
##Lowercase letters for exact match
dataset_spread$state_full <- tolower(dataset_spread$state_full)
##Save the new name for state
dataset_spread$state <- as.factor(dataset_spread$state_full)

##Determine frequency for each state by the amount of records. *DC not mapped on map*
state_counts <- dataset_spread %>%
  group_by(state) %>%
  summarise(frequency = n())

##Join frequency for each state in the dataset
USmap <- USmap %>%
  left_join(state_counts, by = c("region" = "state"))
USmap

##Map out the United States using ggplot, and color by frequency used. 
ggplot(USmap, aes(x = long, y = lat, group = group, fill = frequency)) +
  geom_polygon(color = "black") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "gray") +
  theme_minimal() +
  labs(title = "US Map with Highlighted States", fill = "Frequency")


```

Summary Statistics
```{r}
## Average income for each state 
table <- dataset_spread %>%
  group_by(state) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table

```
Linear Model
```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
glimpse(dataset_train)

pricelm <- recipe(price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed + state + source , data = dataset_train)

workflow <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm)

fitted_model <- workflow %>% 
  fit(data = dataset_train)

predictions <- fitted_model %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics <- predictions %>% 
  metrics(truth = price, estimate = .pred)

print(metrics)
```

###Logisitic Regression
Average price of house, above mean 1, below 0. See if we can get better predictions from binary logistic regression. In this model we get 19.45% error rate using a logistic model with the same predictors. 
```{r}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog <- recipe(bi_price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed + state + source , data = dataset_train)

workflow_log <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog)

fitted_model_log <- workflow_log %>% 
  fit(data = dataset_train)

predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)
predictions_log

class_predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)
class_predictions_log

metrics_log <- class_predictions_log %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log <-class_predictions_log %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

print(matrix_log)

(239+138)/(1104+239+138+457)

print(metrics_log)

```
### Choosing the correct predictors:
In our model, we have a significant amount of factor variables, where there is a lot of categorical data. When these categories are broke up there are 90+ variables that could be included in the model. For example: state includes all 47 states that is contained in the data set. Knowing which predictors are significant to the model is difficult to determine so we used forward and backward selection to weed out of all the variables which would be the best to make predictions on. 

When talking about price of apartments, we know location can have an impact over the general price. For these models, we were more curious about separating out amenities, bedrooms, ad type, etc. 

Based on forward selection and using the BIC and ADJR we see that the predictors of availability of a playground, hot tub, view, wood floors, washer dryer, patio/ deck, elevator, garbage disposal, refrigerator are all amenities that influence price in the model. Rent Lingo and ListedBuy are the only two sources of posting that tend to influence the predictability of price. Lastly, the square footage and bathrooms are significant to the price of the apartment. 

The adjusted r square value indicates a slightly different model would be best but it only explains about 25.28% of the variability. That model includes all the above predictors and adds Tennis and Pool to the list of significant predictors. We know from our linear model that included every variable, specifically the state variable is going to explain a lot of our variability, and with all predictors in that model, naturally R^2 will be higher. 

The backwards selection indicated the same model as the forward selection did for BIC and AdjR. This leads us to decrease our predictors on our models and disregard insignificant amenities that do not influence the price point of the apartment. 

```{r}
library(leaps)
sub_for <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "forward", nvmax=20)
summary(sub_for)
summary(sub_for)$bic
summary(sub_for)$adjr2

sub_back <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "backward", nvmax = 20)
summary(sub_back)
summary(sub_back)$bic
summary(sub_back)$adjr2
```


```{r}

```