---
title: "DSCI445 - Group Project"
authors: "Amanda Bartels, Mohamed Habibniaykoochesfahani, Isabella Kukor, Casey Martin"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


## Basic R

# Remove NA fields

```{r, include=FALSE}

dataset <- read.csv("apartments_for_rent_classified_10K.csv", sep = ";")

head(dataset)
sapply(dataset, class)

# View summary statistics for each column
summary(dataset)

table(dataset$category)
table(dataset$currency)
table(dataset$state)
table(dataset$bedrooms)
table(dataset$fee)
table(dataset$has_photo)
table(dataset$pets_allowed)
table(dataset$price_type)
table(dataset$source)

```

# Remove columns that do not contain useful information:

Removed 'currency' column because all values in it are "USD".
Removed 'fee' column because all values are "No" so this will not provide insights.
Kept value 'has_photo' because it could correlate with listing quality and affect price.
Removed 'price_display' column because it is a repeat of price column.
Removed 'address' because it's not useful and has a large ammount of NAs.
Kept(?) 'source' column tracks the source of the listing (i.e. different websites). This is only useful for comparing listings across sources which we will/not be doing in this analysis.
'title' and 'body' columns are removed because they contain text descriptions which are not going to be analyzed in this study.

```{r, include=FALSE}

library(tidyverse)

# Remove columns
dataset <- dataset %>%
  select(
    -currency,
    -fee,
    -price_display,
    -title,
    -body,
    -address
    # Uncomment the line below to remove 'source' (website listing was posted on)
    # -source
  )

# View cleaned dataset
str(dataset)


```


# Remove null values.

```{r, include=FALSE}
# Replace "null" with NA across all character columns in the dataset for easier handling
dataset[] <- lapply(dataset, function(x) {
  if (is.character(x)) {
    x[x == "null"] <- NA
  }
  return(x)
})

# Get a summary of NA counts for each column in the dataset
colSums(is.na(dataset))

# Remove rows with any NA values
dataset <- na.omit(dataset)

# Check the structure and summary of the dataset to verify changes
str(dataset)
summary(dataset)

```


# Standardizes categorical values for consistency.

```{r}

# Change `has_photo` entries containing "No Thumbnail" into "No" category so only categories are "Yes" or "No"
dataset$has_photo[dataset$has_photo == "No Thumbnail"] <- "No"

# Correct the special character in `source`
dataset$source[dataset$source == "RENTCaf\xe9"] <- "RENTCafé"

```

# Split ammenities into binary columns

```{r}
dataset_split <- dataset %>%
  # Split and widen amenities
  mutate(amenities_split = str_split(amenities, ",")) %>%
  unnest(amenities_split) %>%
  mutate(has_amenity = 1) %>%
  pivot_wider(names_from = amenities_split, values_from = has_amenity, values_fill = 0) %>%
  select(-amenities)
```

