---
title: "DSCI445 - Group Project"
authors: "Amanda Bartels, Mohamed Habibniaykoochesfahani, Isabella Kukor, Casey Martin"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


## Basic R

# Remove NA fields

```{r, include=FALSE}

dataset <- read.csv("group project/apartments_for_rent_classified_10K.csv", sep = ";")

head(dataset)
sapply(dataset, class)

# View summary statistics for each column
summary(dataset)

table(dataset$category)
table(dataset$currency)
table(dataset$state)
table(dataset$bedrooms)
table(dataset$fee)
table(dataset$has_photo)
table(dataset$pets_allowed)
table(dataset$price_type)
table(dataset$source)

```

# Remove columns that do not contain useful information:

Removed 'currency' column because all values in it are "USD".
Removed 'fee' column because all values are "No" so this will not provide insights.
Kept value 'has_photo' because it could correlate with listing quality and affect price.
Removed 'price_display' column because it is a repeat of price column.
Kept(?) 'source' column tracks the source of the listing (i.e. different websites). This is only useful for comparing listings across sources which we will/not be doing in this analysis.
'title' and 'body' columns are removed because they contain text descriptions which are not going to be analyzed in this study.
Removed 'address' because it's not useful and has a large ammount of NAs.
Removed 'latitude', 'longitude', 'address', 'cityname', and 'time' columns because they are not interpretable in this study.
Only 4 rows in the entire dataset have \textit{category} values other than 'apartment', so these rows and consequently the "category" column are deleted.  
Only 2 rows in the entire dataset have \textit{price\_type} values other than 'Monthly', so these rows and consequently the "price\_type" column are deleted.

```{r, include=FALSE}
library(tidyverse)

# Remove rows
dataset <- dataset[dataset$category=='housing/rent/apartment',]
dataset <- dataset[dataset$price_type=='Monthly',]

# Remove columns
dataset <- dataset %>%
  select(
    -currency,
    -fee,
    -price_display,
    -title,
    -body,
    -price_type,
    -address,
    -cityname,
    -time,
    -category,
    -longitude,
    -latitude
    
    
    # Uncomment the line below to remove 'source' (website listing was posted on)
    # -source
  )

# View cleaned dataset
str(dataset)


```


# Remove null values.

```{r, include=FALSE}
# Replace "null" with NA across all character columns in the dataset for easier handling
dataset[] <- lapply(dataset, function(x) {
  if (is.character(x)) {
    x[x == "null"] <- NA
  }
  return(x)
})

# Get a summary of NA counts for each column in the dataset
colSums(is.na(dataset))

# Remove rows with any NA values
dataset <- na.omit(dataset)

# Check the structure and summary of the dataset to verify changes
str(dataset)
summary(dataset)

```


# Standardizes categorical values for consistency.

```{r}

# Change `has_photo` entries containing "No Thumbnail" into "No" category so only categories are "Yes" or "No"
dataset$has_photo[dataset$has_photo == "No Thumbnail"] <- "No"

# Correct the special character in `source`
dataset$source[dataset$source == "RENTCaf\xe9"] <- "RENTCafé"

```

# Spread amenities column

```{r}
dataset_spread <- dataset %>%
  # Split and widen amenities
  mutate(amenities_split = str_split(amenities, ",")) %>%
  unnest(amenities_split) %>%
  mutate(has_amenity = 1) %>%
  pivot_wider(names_from = amenities_split, values_from = has_amenity, values_fill = 0) %>%
  select(-amenities, -id)
```

# Train test split
```{r}
library(tidymodels)
set.seed(445)
dataset_spread$bathrooms <- as.numeric(dataset_spread$bathrooms)
dataset_spread$bedrooms <- as.numeric(dataset_spread$bedrooms)
mean(dataset_spread$price)
dataset_spread$bi_price[dataset_spread$price >=1356.544] <- "1"
dataset_spread$bi_price[dataset_spread$price < 1356.544] <- "0"
dataset_spread$bi_price <- as.factor(dataset_spread$bi_price)
dataset_split <- initial_split(dataset_spread, prop = 0.6)
dataset_train <- training(dataset_split)
dataset_test <- testing(dataset_split)
dataset_train
```

##Map of Apartments
```{r}
##Create dataset for the states with longitude and latitude. map_data function is in ggplot2
USmap <- map_data("state")
USmap
dataset_spread
##Need to change abbreviations to full state name to match with USMap datatset 
dataset_spread$state_full <- state.name[match(dataset_spread$state, state.abb)]
##Lowercase letters for exact match
dataset_spread$state_full <- tolower(dataset_spread$state_full)
##Save the new name for state
dataset_spread$state <- as.factor(dataset_spread$state_full)

##Determine frequency for each state by the amount of records. *DC not mapped on map*
state_counts <- dataset_spread %>%
  group_by(state) %>%
  summarise(frequency = n())

##Join frequency for each state in the dataset
USmap <- USmap %>%
  left_join(state_counts, by = c("region" = "state"))
USmap

##Map out the United States using ggplot, and color by frequency used. 
ggplot(USmap, aes(x = long, y = lat, group = group, fill = frequency)) +
  geom_polygon(color = "black") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "gray") +
  theme_minimal() +
  labs(title = "US Map with Highlighted States", fill = "Frequency")


```

Summary Statistics
```{r}
## Average income for each state 
table <- dataset_spread %>%
  group_by(state) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table

## Average price for each listing source
table2 <- dataset_spread %>%
  group_by(source) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table2
## Average price based on bathrooms
table3 <- dataset_spread %>%
  group_by(bathrooms) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table3
##Average price based on bedrooms 
table4 <- dataset_spread %>%
  group_by(bedrooms) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table4


```
Linear Model
```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
glimpse(dataset_train)

pricelm <- recipe(price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed + state + source , data = dataset_train)

workflow <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm)

fitted_model <- workflow %>% 
  fit(data = dataset_train)

predictions <- fitted_model %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics <- predictions %>% 
  metrics(truth = price, estimate = .pred)

print(metrics)
```

###Logisitic Regression
Average price of house, above mean 1, below 0. See if we can get better predictions from binary logistic regression. In this model we get 19.45% error rate using a logistic model with the same predictors. 

```{r}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog <- recipe(bi_price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed + state + source , data = dataset_train)

workflow_log <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog)

fitted_model_log <- workflow_log %>% 
  fit(data = dataset_train)

predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)
predictions_log

class_predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)
class_predictions_log

metrics_log <- class_predictions_log %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log <-class_predictions_log %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

print(matrix_log)

(239+138)/(1104+239+138+457)

print(metrics_log)

```
### Choosing the correct predictors:
In our model, we have a significant amount of factor variables, where there is a lot of categorical data. When these categories are broke up there are 90+ variables that could be included in the model. For example: state includes all 47 states that is contained in the data set. Knowing which predictors are significant to the model is difficult to determine so we used forward and backward selection to weed out of all the variables which would be the best to make predictions on. 

When talking about price of apartments, we know location can have an impact over the general price. For these models, we were more curious about separating out amenities, bedrooms, ad type, etc. 

Based on forward selection and using the BIC and ADJR we see that the predictors of availability of a playground, hot tub, view, wood floors, washer dryer, patio/ deck, elevator, garbage disposal, refrigerator are all amenities that influence price in the model. Rent Lingo and ListedBuy are the only two sources of posting that tend to influence the predictability of price. Lastly, the square footage and bathrooms are significant to the price of the apartment. 

The adjusted r square value indicates a slightly different model would be best but it only explains about 25.28% of the variability. That model includes all the above predictors and adds Tennis and Pool to the list of significant predictors. We know from our linear model that included every variable, specifically the state variable is going to explain a lot of our variability, and with all predictors in that model, naturally R^2 will be higher. 

The backwards selection indicated the same model as the forward selection did for BIC and AdjR. This leads us to decrease our predictors on our models and disregard insignificant amenities that do not influence the price point of the apartment. 

```{r}
library(leaps)
sub_for <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "forward", nvmax=20)
summary(sub_for)
summary(sub_for)$bic
summary(sub_for)$adjr2

sub_back <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "backward", nvmax = 20)
summary(sub_back)
summary(sub_back)$bic
summary(sub_back)$adjr2
```
Linear model with predictors
Slight improvement in our R squared value where our model is explaining 58.9% 

```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
glimpse(dataset_train)
dataset_spread$`Garbage Disposal`

pricelm2 <- recipe(price ~ square_feet + bathrooms + Pool + View + Refrigerator+ `Patio/Deck` +Playground +`Hot Tub` + Elevator + `Wood Floors`+ `Garbage Disposal`+ state + source , data = dataset_train)

workflow2 <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm2)

fitted_model2 <- workflow2 %>% 
  fit(data = dataset_train)

predictions2 <- fitted_model2 %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics2 <- predictions2 %>% 
  metrics(truth = price, estimate = .pred)

print(metrics2)
```
###Logisitc Regression with Predictors

Not much changed in our prediction error or accuracy, but the mean of all prices was selected as an arbitrary cut off point. Logistic regression tends to outperform the linear model in accuracy for predicting whether the prices will be lower or higher than average, but since we dont know if this cut off point of the average of the overall price is a good classification compared to the other models. It is not yet determined whether the logistic model utilizing classification of low and high is better predictor than examining the prices of the model as a continuous response variable in a linear model, or other options.

```{r}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog2 <- recipe(bi_price ~ square_feet + bathrooms + Pool + View + Refrigerator+ `Patio/Deck` +Playground +`Hot Tub` + Elevator + `Wood Floors`+ `Garbage Disposal`+ state + source, data = dataset_train)

workflow_log2 <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog2)

fitted_model_log2 <- workflow_log2 %>% 
  fit(data = dataset_train)

predictions_log2 <- fitted_model_log2 %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)
predictions_log2

class_predictions_log2 <- fitted_model_log2 %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)
class_predictions_log2

metrics_log2 <- class_predictions_log2 %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log2 <-class_predictions_log2 %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

print(matrix_log2)

(235+138)/(1104+235+138+461)

print(metrics_log2)

``` 


#"Lasso" and "Ridge" regressions are used to include regularization to the linear regression and prevent overfitting.
```{r}

library(tidymodels)
library(glmnet)
library(ggplot2)

set.seed(445)

rec <- recipe(price ~ ., data = dataset_train) %>%
  step_novel(all_nominal_predictors()) %>% # Handle unseen factor levels
  step_dummy(all_nominal_predictors()) %>% # Convert categorical variables to dummy variables
  step_zv(all_predictors()) %>%  # Remove zero-variance 
  step_normalize(all_numeric_predictors()) # Normalize numeric predictors

prepped_rec <- prep(rec, training = dataset_train)

x_train <- bake(prepped_rec, new_data = dataset_train)
x_test <- bake(prepped_rec, new_data = dataset_test)

x_train <- as.matrix(select(x_train, -price))
y_train <- dataset_train$price

x_test <- as.matrix(select(x_test, -price))
y_test <- dataset_test$price

# Fit ridge regression model with cross-validation
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_ridge$lambda.min

# Fit final ridge regression model with best lambda
ridge_model <- glmnet(x_train, y_train, alpha = 0, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(ridge_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)


# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(ridge_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_ridge$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_ridge$lambda

# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  # Add circle around best_lambda
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20, label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  # Add annotation for best lambda

# Print the plot
print(cv_rmse_plot)

comparison_plot <- data.frame(Actual = y_test, Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Reference line y=x
  labs(title = "Actual vs Predicted Values",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)

comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  # Predicted values

comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   # Line for actual values
  geom_line(aes(y = Predicted, color = "Predicted"), size = 1, linetype = "dashed",alpha = 0.8) +  # Line for predicted values
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  # Custom colors
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```

```{r}
set.seed(445)

# Fit Lasso regression model with cross-validation (alpha = 1 for Lasso)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_lasso$lambda.min

# Fit final Lasso regression model with best lambda
lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(lasso_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)

# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(lasso_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_lasso$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_lasso$lambda



# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  # Add circle around best_lambda
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20, label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  # Add annotation for best lambda

# Print the plot
print(cv_rmse_plot)

# Plot: Actual vs Predicted for Test Data
comparison_plot <- data.frame(Actual = y_test, Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Reference line y=x
  labs(title = "Actual vs Predicted Values for Lasso",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)



comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  # Predicted values

comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   # Line for actual values
  geom_line(aes(y = Predicted, color = "Predicted",alpha = 0.8), size = 1, linetype = "dashed", alpha = 0.8) +  # Line for predicted values
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  # Custom colors
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```

Despite of utilizing cross-validation to select an optimal regularization parameter, the Ridge regression model shows signs of overfitting. There is a noticeable gap between training and test performance, with the model performing better on the training data, indicating it may have learned patterns specific to the training set that don't generalize well. While regularization has helped reduce overfitting to some extent, the results remain relatively unchanged when switching to Lasso, suggesting the issue isn't due to the type of regularization but rather the dataset's complexity. Given that cross-validation with ten folds is a rational choice, the next steps should focus on using more complex models like decision trees, random forests, or gradient boosting, which may better capture the underlying patterns of the data.
>>>>>>> origin/main
