---
output: pdf_document
geometry: landscape
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE,message=FALSE)
```

\newpage
\begin{center}
\vspace*{\fill} % Push content to the vertical center

{\Huge Apartment Rental Rates Across the United States}

\vspace{3cm} % Add vertical space

{\large Amanda Bartels, Mohamad Habibniaykoochesfahani, Isabella Kukor, Casey Martin}

\vspace{3cm} % Add vertical space

{\normalsize December 2, 2024}

\vspace*{\fill} % Push content to the vertical center
\end{center}
\newpage

\LARGE
# Dataset introduction and cleaning

\large
- The "Apartment for Rent Classified" dataset, donated on December 25, 2019, is a multivariate dataset containing classified ads for apartments for rent in the USA.
- The dataset consists of 10,000 instances with 21 categorical and integer features.
- Attributes  includes details like apartment amenities (e.g., AC, gym, pool), bathrooms, bedrooms, rental price, square footage, location (address, city, state), and more.
- Each variable in the dataset was examined for NA values.

- Categorical data counts explored potential levels with 1 count. For example:
  - Only 1 apartment was pay type “Weekly.”
  - 1 apartment was both “Monthly and Weekly.”
  - 9998 apartments were “Monthly.”

- This led to removing unnecessary variables in the dataset, such as address, time, longitude, latitude, etc.

- All NA data points were removed.




\newpage

```{r,include=FALSE}

dataset <- read.csv("apartments_for_rent_classified_10K.csv", sep = ";")

head(dataset)
sapply(dataset, class)

# View summary statistics for each column
summary(dataset)

table(dataset$category)
table(dataset$currency)
table(dataset$state)
table(dataset$bedrooms)
table(dataset$fee)
table(dataset$has_photo)
table(dataset$pets_allowed)
table(dataset$price_type)
table(dataset$source)

```


```{r}
library(tidyverse)
library(knitr)

# Remove rows
dataset <- dataset[dataset$category=='housing/rent/apartment',]
dataset <- dataset[dataset$price_type=='Monthly',]

# Remove columns
dataset <- dataset %>%
  dplyr::select(
    -currency,
    -fee,
    -price_display,
    -title,
    -body,
    -price_type,
    -address,
    -cityname,
    -time,
    -category,
    -longitude,
    -latitude
    
    
    # Uncomment the line below to remove 'source' (website listing was posted on)
    # -source
  )

summary_table <- data.frame(
  Type = sapply(dataset, class),
  Missing = sapply(dataset, function(x) sum(is.na(x))),
  Unique = sapply(dataset, function(x) length(unique(x)))
)

knitr::kable(summary_table, format = "latex", caption = "Summary of the Cleaned Dataset")

```
\large
- Several amenities were included in the dataset. These were pivoted wider, and dummy variables were attached to extract information on specific amenities.

- The dataset was split into training and testing data to remain consistent across all portions of the project.

\newpage

```{r,include=FALSE}
# Replace "null" with NA across all character columns in the dataset for easier handling
dataset[] <- lapply(dataset, function(x) {
  if (is.character(x)) {
    x[x == "null"] <- NA
  }
  return(x)
})

# Get a summary of NA counts for each column in the dataset
colSums(is.na(dataset))

# Remove rows with any NA values
dataset <- na.omit(dataset)

# Check the structure and summary of the dataset to verify changes
str(dataset)
summary(dataset)

```



```{r,include=FALSE}

# Change `has_photo` entries containing "No Thumbnail" into "No" category so only categories are "Yes" or "No"
dataset$has_photo[dataset$has_photo == "No Thumbnail"] <- "No"

# Correct the special character in `source`
dataset$source[dataset$source == "RENTCaf\xe9"] <- "RENTCafe"

```
\newpage


```{r,include=FALSE}
dataset_spread <- dataset %>%
  # Split and widen amenities
  mutate(amenities_split = str_split(amenities, ",")) %>%
  unnest(amenities_split) %>%
  mutate(has_amenity = 1) %>%
  pivot_wider(names_from = amenities_split, values_from = has_amenity, values_fill = 0) %>%
  select(-amenities, -id)

dataset_spread_copy <- dataset_spread

```

```{r,include=FALSE}
library(tidymodels)
set.seed(445)
dataset_spread$bathrooms <- as.numeric(dataset_spread$bathrooms)
dataset_spread$bedrooms <- as.numeric(dataset_spread$bedrooms)
mean(dataset_spread$price)
dataset_spread$bi_price[dataset_spread$price >=1356.544] <- "1"
dataset_spread$bi_price[dataset_spread$price < 1356.544] <- "0"
dataset_spread$bi_price <- as.factor(dataset_spread$bi_price)
dataset_split <- initial_split(dataset_spread, prop = 0.6)
dataset_train <- training(dataset_split)
dataset_test <- testing(dataset_split)
dataset_train
```
\newpage

# Distribution Map of the Apartments
```{r}
##Create dataset for the states with longitude and latitude. map_data function is in ggplot2
USmap <- map_data("state")

##Need to change abbreviations to full state name to match with USMap datatset 
dataset_spread$state_full <- state.name[match(dataset_spread$state, state.abb)]
##Lowercase letters for exact match
dataset_spread$state_full <- tolower(dataset_spread$state_full)
##Save the new name for state
dataset_spread$state <- as.factor(dataset_spread$state_full)

##Determine frequency for each state by the amount of records. *DC not mapped on map*
state_counts <- dataset_spread %>%
  group_by(state) %>%
  summarise(frequency = n())

##Join frequency for each state in the dataset
USmap <- USmap %>%
  left_join(state_counts, by = c("region" = "state"))

##Map out the United States using ggplot, and color by frequency used. 
ggplot(USmap, aes(x = long, y = lat, group = group, fill = frequency)) +
  geom_polygon(color = "black") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", na.value = "gray") +
  theme_minimal() +
  labs(title = "US Map with Highlighted States", fill = "Frequency")


```
\newpage

```{r,include=FALSE}
## Average income for each state 
table <- dataset_spread %>%
  group_by(state) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table

## Average price for each listing source
table2 <- dataset_spread %>%
  group_by(source) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table2
## Average price based on bathrooms
table3 <- dataset_spread %>%
  group_by(bathrooms) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table3
##Average price based on bedrooms 
table4 <- dataset_spread %>%
  group_by(bedrooms) %>%
  summarize(
    count =n(),
    avg_price = mean(price, na.rm = TRUE))
table4


```
\newpage
\LARGE
# Linear Model
\large
- A linear regression model with all remaining predictors was used to predict the price of apartments.
- Price is a continuous variable, making it an easy candidate for a linear model.
- Linear models are easily interpretable, as any predictor would showcase a dollar amount associated with the addition of features such as square footage, amenities, bathrooms, etc.
- The focus was on evaluating the accuracy of the model and determining whether other models learned in class would outperform simpler models.

\newpage

```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
#glimpse(dataset_train)

pricelm <- recipe(price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed +
                    state + source , data = dataset_train)

workflow <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm)

fitted_model <- workflow %>% 
  fit(data = dataset_train)

predictions <- fitted_model %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics <- predictions %>% 
  metrics(truth = price, estimate = .pred)

metrics %>%
  knitr::kable(
    format = "latex",
    caption = "Model Performance Metrics",
    align = "c"
  )
```
\large
- The model has limitations in fully explaining price variability (as seen from the R² value).
- The errors (RMSE and MAE) indicate that further optimization might improve performance.

\newpage

# Logistic Regression
Average price of house, above mean 1, below 0. See if we can get better predictions from binary logistic regression. In this model we get 19.45% error rate using a logistic model with the same predictors. 

```{r,include=FALSE}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog <- recipe(bi_price ~ square_feet + bathrooms + bedrooms + has_photo + pets_allowed + state + source , data = dataset_train)

workflow_log <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog)

fitted_model_log <- workflow_log %>% 
  fit(data = dataset_train)

predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)

class_predictions_log <- fitted_model_log %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)

metrics_log <- class_predictions_log %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log <-class_predictions_log %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

#print(matrix_log)

#(239+138)/(1104+239+138+457)

metrics_log %>%
  knitr::kable(
    format = "latex",
    caption = "Model Performance Metrics",
    align = "c"
  )
```
## Choosing the correct predictors:
- The model includes a significant number of factor variables, with over 90 potential predictors when categorical data is broken into categories (e.g., state includes 47 unique states).
- Forward and backward selection were used to identify the best predictors for making price predictions.
- Location (e.g., state) can influence apartment prices, but this analysis focused on separating out amenities, bedrooms, ad type,      etc.
- Forward selection using BIC and Adjusted R² identified the following predictors:
  - Amenities: Playground, Hot Tub, View, Wood Floors, Washer/Dryer, Patio/Deck, Elevator, Garbage Disposal, and Refrigerator.
  - Sources: Only two sources (Rent Lingo and ListedBuy) influenced price predictability.
  - Structural Features: Square footage and Bathrooms.
- The adjusted R² model explains about 25.28% of the variability in apartment prices and includes all the above predictors plus Additional Amenities of Tennis Court and Pool.
- A full linear model that includes all variables (e.g., state) naturally has a higher R² value because it explains more variability.
- Backward selection confirmed the same model as forward selection when using BIC and Adjusted R².
- Insignificant predictors (e.g., certain amenities) were excluded to simplify the model.

```{r, include=FALSE}
library(leaps)
sub_for <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "forward", nvmax=20)
summary(sub_for)
summary(sub_for)$bic
summary(sub_for)$adjr2

sub_back <- regsubsets(price ~ . -bi_price -state, data = dataset_spread, method = "backward", nvmax = 20)
summary(sub_back)
summary(sub_back)$bic
summary(sub_back)$adjr2
```
Linear model with predictors shows slight improvement in our R squared value where our model is explaining 58.9%.

```{r}
set.seed(445)
linear_model <- linear_reg() %>% 
  set_engine("lm")
#glimpse(dataset_train)
#dataset_spread$`Garbage Disposal`

pricelm2 <- recipe(price ~ square_feet + bathrooms + Pool + View + Refrigerator+ `Patio/Deck` +Playground +`Hot Tub` + Elevator + `Wood Floors`+ `Garbage Disposal`+ state + source , data = dataset_train)

workflow2 <- workflow() %>% 
  add_model(linear_model) %>% 
  add_recipe(pricelm2)

fitted_model2 <- workflow2 %>% 
  fit(data = dataset_train)

predictions2 <- fitted_model2 %>% 
  predict(new_data = dataset_test) %>% 
  bind_cols(dataset_test)

metrics2 <- predictions2 %>% 
  metrics(truth = price, estimate = .pred)

print(metrics2)
```
\newpage

# Logisitc Regression with Predictors

Not much changed in our prediction error or accuracy, but the mean of all prices was selected as an arbitrary cut off point. Logistic regression tends to outperform the linear model in accuracy for predicting whether the prices will be lower or higher than average, but since we dont know if this cut off point of the average of the overall price is a good classification compared to the other models. It is not yet determined whether the logistic model utilizing classification of low and high is better predictor than examining the prices of the model as a continuous response variable in a linear model, or other options.

```{r}
set.seed(445)
logistic_model <- logistic_reg() %>%
  set_engine("glm")

pricelog2 <- recipe(bi_price ~ square_feet + bathrooms + Pool + View + Refrigerator+ `Patio/Deck` +Playground +`Hot Tub` + Elevator + `Wood Floors`+ `Garbage Disposal`+ state + source, data = dataset_train)

workflow_log2 <- workflow() %>% 
  add_model(logistic_model) %>% 
  add_recipe(pricelog2)

fitted_model_log2 <- workflow_log2 %>% 
  fit(data = dataset_train)

predictions_log2 <- fitted_model_log2 %>%
  predict(new_data = dataset_test, type = "prob") %>%
  bind_cols(dataset_test)
#predictions_log2

class_predictions_log2 <- fitted_model_log2 %>%
  predict(new_data = dataset_test) %>%
  bind_cols(dataset_test)
#class_predictions_log2

metrics_log2 <- class_predictions_log2 %>%
  metrics(truth = bi_price, estimate = .pred_class)

matrix_log2 <-class_predictions_log2 %>%
  conf_mat(truth = bi_price, estimate = .pred_class)

#print(matrix_log2)

(235+138)/(1104+235+138+461)

print(metrics_log2)

``` 

\newpage

# Ridge Regression Technique 
```{r,include=FALSE}

library(tidymodels)
library(glmnet)
library(ggplot2)

set.seed(445)
```

```{r,warning=FALSE}
rec <- recipe(price ~ ., data = dataset_train) %>%
  step_novel(all_nominal_predictors()) %>% # Handle unseen factor levels
  step_dummy(all_nominal_predictors()) %>% # Convert categorical variables to dummy variables
  step_zv(all_predictors()) %>%  # Remove zero-variance 
  step_normalize(all_numeric_predictors()) # Normalize numeric predictors

prepped_rec <- prep(rec, training = dataset_train)

x_train <- bake(prepped_rec, new_data = dataset_train)
x_test <- bake(prepped_rec, new_data = dataset_test)

x_train <- as.matrix(select(x_train, -price))
y_train <- dataset_train$price

x_test <- as.matrix(select(x_test, -price))
y_test <- dataset_test$price

# Fit ridge regression model with cross-validation
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_ridge$lambda.min

# Fit final ridge regression model with best lambda
ridge_model <- glmnet(x_train, y_train, alpha = 0, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(ridge_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)


# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(ridge_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_ridge$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_ridge$lambda

# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  # Add circle around best_lambda
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20, label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  # Add annotation for best lambda

# Print the plot
print(cv_rmse_plot)
```
\newpage    
```{r,warning=FALSE}
comparison_plot <- data.frame(Actual = y_test, Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Reference line y=x
  labs(title = "Actual vs Predicted Values",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)
```

```{r,warning=FALSE}

comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  # Predicted values

comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   # Line for actual values
  geom_line(aes(y = Predicted, color = "Predicted"), size = 1, linetype = "dashed",alpha = 0.8) +  # Line for predicted values
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  # Custom colors
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```
\newpage

# Lasso Regression Technique 

```{r,warning=FALSE}
set.seed(445)

# Fit Lasso regression model with cross-validation (alpha = 1 for Lasso)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

# Get best lambda from cross-validation
best_lambda <- cv_lasso$lambda.min

# Fit final Lasso regression model with best lambda
lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda)

# Make predictions on test set
predictions_test <- predict(lasso_model, s = best_lambda, newx = x_test)

# Calculate Test MSE and RMSE
test_mse <- mean((y_test - predictions_test)^2)
test_rmse <- sqrt(test_mse)

# Calculate Test R2
r2_test <- 1 - sum((y_test - predictions_test)^2) / sum((y_test - mean(y_test))^2)

# Make predictions on training set for training error calculation
predictions_train <- predict(lasso_model, s = best_lambda, newx = x_train)

# Calculate Training R2
r2_train <- 1 - sum((y_train - predictions_train)^2) / sum((y_train - mean(y_train))^2)

# Calculate Training MSE and RMSE
train_mse <- mean((y_train - predictions_train)^2)
train_rmse <- sqrt(train_mse)

# Extract cross-validation results (RMSE for each lambda)
cv_rmse <- sqrt(cv_lasso$cvm)  # cvm contains MSE for each lambda
lambda_values <- cv_lasso$lambda



# Plot RMSE vs Lambda (cross-validation results)
cv_rmse_plot <- data.frame(lambda_values, cv_rmse) %>%
  ggplot(aes(x = log(lambda_values), y = cv_rmse)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  geom_point(aes(x = log(best_lambda), y = min(cv_rmse)), 
             shape = 1, size = 4, color = "green") +  # Add circle around best_lambda
  labs(title = "CV RMSE vs. Log(Lambda)",
       x = "Log(Lambda)",
       y = "Cross-Validation RMSE") +
  theme_minimal() +
  annotate("text", x = log(best_lambda), y = min(cv_rmse) + 20, label = paste("Best Lambda: ", round(best_lambda, 2)), 
           color = "green", angle = 0, vjust = 0, hjust = 1)  # Add annotation for best lambda

# Print the plot
print(cv_rmse_plot)
```
\newpage
```{r,warning=FALSE}
# Plot: Actual vs Predicted for Test Data
comparison_plot <- data.frame(Actual = y_test, Predicted = as.vector(predictions_test)) %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +  # Reference line y=x
  labs(title = "Actual vs Predicted Values for Lasso",
       x = "Actual Values",
       y = "Predicted Values") +
  theme_minimal()
print(comparison_plot)
```

```{r,warning=FALSE}


comparison_df <- data.frame(Index = 1:length(y_test),  # Index for x-axis
                            Actual = y_test,           # Actual values
                            Predicted = as.vector(predictions_test))  # Predicted values

comparison_plot2 <- comparison_df %>%
  ggplot(aes(x = Index)) +
  geom_line(aes(y = Actual, color = "Actual"), size = 1) +   # Line for actual values
  geom_line(aes(y = Predicted, color = "Predicted",alpha = 0.8), size = 1, linetype = "dashed", alpha = 0.8) +  # Line for predicted values
  labs(title = "Actual vs Predicted Values Over Index",
       x = "Index",
       y = "Values") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +  # Custom colors
  theme_minimal()

# Print the plot
print(comparison_plot2)

error_summary <- tibble(
  Metric = c("Training MSE", "Training RMSE", "Training R2", "Test MSE", "Test RMSE", "Test R2"),
  Value = c(train_mse, train_rmse, r2_train, test_mse, test_rmse, r2_test)
)

# Print the error summary tibble
print(error_summary)
```

Cross-validation is used to select an optimal regularization parameter, helping reduce overfitting to some extent; However, the results for both Lasso and Ridge models show a slight improvement comparing to the previously developed linear model. This suggests that the issue isn't due to the type of regularization but rather the dataset's complexity. Given that cross-validation with ten folds is a rational choice, the next steps should focus on using more complex models like decision trees, random forests, or gradient boosting, which may better capture the underlying patterns of the data.

\newpage
# Tree Based Methods
## Remove rare levels
```{r,echo=TRUE}
# Threshold for minimum occurrences
threshold <- 10
# Identify rows to keep
rows_to_keep <- seq(TRUE, nrow(dataset_spread_copy))
# Process each column
for (col_name in colnames(dataset_spread_copy)) {
  if (is.character(dataset_spread_copy[[col_name]]) || 
      is.factor(dataset_spread_copy[[col_name]])) {
    col_table <- table(dataset_spread_copy[[col_name]])
    valid_categories <- names(col_table[col_table >= threshold])
    rows_to_keep <- rows_to_keep & dataset_spread_copy[[col_name]] %in% 
      valid_categories
  }
}
dataset_spread_trees <- dataset_spread_copy[rows_to_keep, ]
```
Due to the extreme inbalance of the dataset, levels with less than 10 values were removed as they could cause a mismatch between the training and testing sets as well as the different folds for cross validation.

```{r}
set.seed(445)

dataset_spread_trees <- dataset_spread_trees %>% select(-state, -source)


dataset_trees_split <- initial_split(dataset_spread_trees, prop = 0.6)
dataset_trees_train <- training(dataset_trees_split)
dataset_trees_test <- testing(dataset_trees_split)
```
\newpage

# Decision Tree
## Tune depth
```{r, cache=TRUE}
library(rpart.plot)
library(vip)
library(doParallel)

set.seed(445)
train_cv_5fold <- vfold_cv(dataset_trees_train, v = 5)

tree_spec <- decision_tree(tree_depth = tune("depth")) |>
  set_engine("rpart") |>
  set_mode("regression")

tree_rec <- recipe(price ~ ., data = dataset_trees_train)

tree_wf <- workflow() |>
  add_model(tree_spec) |>
  add_recipe(tree_rec)

tune_df <- data.frame(depth = seq(1:10))

tree_tune_fit <- tree_wf |>
  tune_grid(resamples = train_cv_5fold, grid = tune_df)

tree_tune_fit |>
  autoplot()
```
\newpage
```{r}
library(rpart.plot)

best_result<- show_best(tree_tune_fit, metric = "rmse", n = 1)
knitr::kable(
  best_result,
  caption = "Best Tuning Parameters Based on RMSE",
  format = "latex"
)

tree_final <- finalize_workflow(tree_wf, select_best(tree_tune_fit, metric = "rmse"))

tree_final_fit <- tree_final |>
  fit(data = dataset_trees_train)

tree_final_fit |>
  extract_fit_engine() |>
  rpart.plot(roundint=FALSE)
```

```{r, results='asis'}
library(vip)

vip(tree_final_fit)

dataset_pred_tree <- predict(tree_final_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_tree))*(sum(dataset_trees_test$price - dataset_pred_tree$.pred))^2)
```

Tuning depth with 5 fold cross validation gives an RMSE of 35.64 and depth equal to just one layer.

\newpage

## Tune alpha
```{r, cache=TRUE}
set.seed(445)
train_cv_5fold <- vfold_cv(dataset_trees_train, v = 5)

tree_spec <- decision_tree(tree_depth = 1, cost_complexity = tune("alpha")) |>
  set_engine("rpart") |>
  set_mode("regression")

tree_rec <- recipe(price ~ ., data = dataset_trees_train)

tree_wf <- workflow() |>
  add_model(tree_spec) |>
  add_recipe(tree_rec)

tune_df <- data.frame(alpha = 10^seq(-3, -1, length.out = 10))

tree_tune_fit <- tree_wf |>
  tune_grid(resamples = train_cv_5fold, grid = tune_df)

tree_tune_fit |>
  autoplot()
```
\newpage
```{r}
best_result <- show_best(tree_tune_fit, metric = "rmse", n = 1)
knitr::kable(
  best_result,
  caption = "Best Tuning Parameters Based on RMSE",
  format = "latex"
)
```

```{r}
library(rpart.plot)
tree_final <- finalize_workflow(tree_wf, select_best(tree_tune_fit, metric = "rmse"))

tree_final_fit <- tree_final |>
  fit(data = dataset_trees_train)

tree_final_fit |>
  extract_fit_engine() |>
  rpart.plot(roundint=FALSE)
```

```{r}

vip(tree_final_fit)

dataset_pred_tree <- predict(tree_final_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_tree))*(sum(dataset_trees_test$price - dataset_pred_tree$.pred))^2)

```

Tuning alpha gives a value of 0.001 and an RMSE score of 35.64, which is the same as with just tuning the depth.

\newpage

# Random Forest

```{r, cache=TRUE}
set.seed(445)

rf_spec <- rand_forest(mtry = sqrt(.cols()), trees = 100) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("regression")

rf_rec <- recipe(price ~ ., data = dataset_trees_train)

rf_wf <- workflow() |>
  add_model(rf_spec) |>
  add_recipe(rf_rec)

rf_fit <- rf_wf |>
  fit(dataset_trees_train)

vip(rf_fit)

dataset_pred_rf <- predict(rf_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_rf))*(sum(dataset_trees_test$price - dataset_pred_rf$.pred))^2)

```

A random forest with 100 trees achieved an RMSE of 135.63, worse than the decision tree model but better than other linear models. Cross-validation was skipped due to long computation times, and 100 trees were manually chosen for optimal performance.

\newpage

## Tune lambda and depth

```{r, cache=TRUE}
#Setup parallel processing
#Comment out if causing issues.
all_cores <- parallel::detectCores(logical = FALSE)
if(all_cores >= 16){
  all_cores <- 32
}
cl <- makePSOCKcluster(all_cores)
registerDoParallel(cl)

set.seed(445)

boost_spec <- boost_tree(trees = 5000, tree_depth = tune('depth'), learn_rate = tune("lambda")) |>
  set_engine("xgboost") |>
  set_mode("regression")

boost_rec <- recipe(price ~ ., data = dataset_trees_train) |>
  step_dummy(all_nominal_predictors())

boost_wf <- workflow() |>
  add_model(boost_spec) |>
  add_recipe(boost_rec)

tune_df <- data.frame(depth = seq(1, 5, length.out = 5), lambda = 10^seq(-3, 1, length.out = 5))

start.time <- Sys.time()

boost_tune_fit <- boost_wf |>
  tune_grid(resamples = train_cv_5fold, grid = tune_df)

end.time <- Sys.time()
time.taken <- end.time - start.time
#print(paste("Time elapsed tuning:", time.taken))

boost_tune_fit |>
  autoplot()
```
\newpage
```{r}
# best_results<- show_best(boost_tune_fit, metric = "rmse", n = 1)
# knitr::kable(
#   best_result,
#   caption = "Best Tuning Parameters Based on RMSE",
#   format = "latex"
# )
boost_final <- finalize_workflow(boost_wf, select_best(boost_tune_fit, metric = "rmse"))

boost_final_fit <- boost_final |>
  fit(data = dataset_trees_train)

vip(boost_final_fit)

dataset_pred_boost <- predict(boost_final_fit, dataset_trees_test)
RMSE <- sqrt((1/nrow(dataset_pred_boost))*(sum(dataset_trees_test$price - dataset_pred_boost$.pred))^2)

```

Tuning depth and lambda simultaneously gives a depth of 2 and lambda equal to 0.01. The RMSE is 173.16, making it the worst of the tree based methods.

\newpage

# Conclusion
| Model         | RMSE   |
|---------------|--------|
| Linear        | 518    |
| Ridge         | 473    |
| Lasso         | 474    |
| Decision Trees| 35     |
| Random Forest | 173    |

- The regression model with the best RMSE is the decision tree model, which is surprising given its simpleness compared to random forests and boosted models, especially considering that XG boost handles unbalanced datasets better.
- Logistic regression had higher accuracy when predicting on higher or lower than average apartment prices based on the predictors provided in the dataset but little improvements can be made to the model past this.


\newpage
# References
- “Apartment for Rent Classified.” UCI Machine Learning Repository. https://doi.org/10.24432/C5X623.

